---
output: hugodown::hugo_document

slug: duckplyr-1-0-0
title: duckplyr fully joins the tidyverse!
date: 2025-02-13
author: Kirill Müller and Maëlle Salmon
description: >
    duckplyr 1.0.0 is on CRAN and part of the tidyverse!
    A drop-in replacement for dplyr, powered by DuckDB for speed.
    It is the most dplyr-like of dplyr backends.

photo:
  url: https://www.pexels.com/photo/a-mallard-duck-on-water-6918877/
  author: Kiril Gruev

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package]
tags:
  - duckplyr
  - dplyr
  - tidyverse
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're very chuffed to announce the release of [duckplyr](https://duckplyr.tidyverse.org) 1.0.0.
This is a new dplyr backend powered by [DuckDB](https://duckdb.org/), a fast in-memory analytical database system[^duckdb].
It joins the rank of dplyr backends together with [dtplyr](https://dtplyr.tidyverse.org) and [dbplyr](https://dbplyr.tidyverse.org).
You can use it instead of dplyr for data small or large.

<!-- FIXME:

We have many more dplyr backends, the two above are just from the tidyverse.
GitHub search: https://github.com/search?q=org%3Acran+%2FS3method%5B%28%5D%28mutate%7Csummarise%29+*%2C%2F&type=code
Do we need an "awesome dplyr" like https://github.com/krlmlr/awesome-vctrs/?

-->

```{r include = FALSE}
options(
  pillar.min_title_chars = 20,
  pillar.max_footer_lines = 7,
  pillar.bold = TRUE
)

Sys.setenv(DUCKPLYR_FALLBACK_INFO = TRUE)
```


You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("duckplyr")
```

This article shows how duckplyr can be used instead of dplyr with data of different size, explain how you can help improve the package, and share a selection of further resources.

## A drop-in replacement for dplyr

Imagine you have to wrangle a huge dataset.
Here we generate one using the [data generator from the TPC-H benchmark](https://duckdb.org/2024/04/02/duckplyr.html#benchmark-tpc-h-q1).

```{r}
lineitem_tbl <- duckdb:::sql("INSTALL tpch; LOAD tpch; CALL dbgen(sf=1); FROM lineitem;")
lineitem_tbl <- tibble::as_tibble(lineitem_tbl)
dplyr::glimpse(lineitem_tbl)
```

We could transform the data using dplyr but we could also transform it using a tool that'll scale well to ever larger data: duckplyr.
The duckplyr package is a _drop-in replacement for dplyr_ that uses _DuckDB for speed_.
You can simply _drop_ duckplyr into your pipeline by loading it, then computations will be efficiently carried out by DuckDB.

[^duckdb]: If you haven't heard about it, you can watch [Hannes Mühleisen's keynote at posit::conf(2024)](https://www.youtube.com/watch?v=GELhdezYmP0&feature=youtu.be).

Below, we express the standard "TPC-H benchmark query 1" in dplyr syntax, but execute it with duckplyr.

```{r include=FALSE}
options(conflicts.policy = list(warn = FALSE))
```


```{r}
library(conflicted)
library(duckplyr)
conflict_prefer("filter", "dplyr", quiet = TRUE)

out <- lineitem_tbl |>
  select(l_shipdate, l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
  filter(l_shipdate <= !!as.Date("1998-09-02")) |>
  select(l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
  summarise(
    sum_qty = sum(l_quantity),
    sum_base_price = sum(l_extendedprice),
    sum_disc_price = sum(l_extendedprice * (1 - l_discount)),
    sum_charge = sum(l_extendedprice * (1 - l_discount) * (1 + l_tax)),
    avg_qty = mean(l_quantity),
    avg_price = mean(l_extendedprice),
    avg_disc = mean(l_discount),
    count_order = n(),
    .by = c(l_returnflag, l_linestatus)
  ) |>
  arrange(l_returnflag, l_linestatus)
out
```

Like with other dplyr backends like dtplyr and dbplyr, duckplyr allows you to get faster results without learning a different syntax.
Unlike other dplyr backends, duckplyr does not require you to change existing code or learn specific idiosyncrasies.
Not only is the syntax the same, the semantics are too!

Start using duckplyr today by attaching it and running your existing dplyr code.
Many operations will be carried out with DuckDB, faster than with dplyr.
The duckplyr package is fully compatible with dplyr: if an operation cannot be carried out with DuckDB, it is automatically outsourced to dplyr.
Over time, we expect fewer and fewer fallbacks to dplyr to be needed.


## How to use duckplyr

To _replace_ dplyr with duckplyr, you can:

- Load duckplyr and then keep your pipeline as is. Calling `library(duckplyr)` overwrites dplyr methods, enabling duckplyr for the entire session no matter how data.frames are created.
    This is shown in the example above.

- Create individual "duck frames" using _conversion functions_ like `duckdb_tibble()` or `as_duckdb_tibble()`, or _ingestion functions_ like `read_csv_duckdb()`.


Then, the data manipulation pipeline uses the exact same syntax as a dplyr pipeline.
The duckplyr package performs the computation using DuckDB.

```{r}
out <- lineitem_tbl |>
  duckplyr::as_duckdb_tibble() |> # this is the only change :-)
  select(l_shipdate, l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
  filter(l_shipdate <= !!as.Date("1998-09-02")) |>
  select(l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
  summarise(
    sum_qty = sum(l_quantity),
    sum_base_price = sum(l_extendedprice),
    sum_disc_price = sum(l_extendedprice * (1 - l_discount)),
    sum_charge = sum(l_extendedprice * (1 - l_discount) * (1 + l_tax)),
    avg_qty = mean(l_quantity),
    avg_price = mean(l_extendedprice),
    avg_disc = mean(l_discount),
    count_order = n(),
    .by = c(l_returnflag, l_linestatus)
  ) |>
  arrange(l_returnflag, l_linestatus)
```


For programming, the resulting object is indistinguishable from a regular tibble, except for the additional class.


```{r}
typeof(out)
class(out)
out$count_order
```

The result could also be computed to a file.

```{r}
# to a file
csv_file <- withr::local_tempfile()
compute_csv(out, csv_file)
fs::file_size(csv_file)
```

Operations not yet supported by duckplyr are automatically outsourced to dplyr.
For instance, filtering on grouped data is not supported yet, still it works thanks to the fallback mechanism.

```{r}
lineitem_tbl |>
  duckplyr::as_duckdb_tibble() |>
  select(l_shipdate, l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
  filter(l_quantity == max(l_quantity), .by = c(l_returnflag, l_linestatus))
```

Using duckplyr is faster than using dplyr.
Below we compare the same pipeline, "TPC-H benchmark query 1", with dplyr and duckplyr.

_Because we compare dplyr with duckplyr here, we no longer can use duckplyr for all dplyr operations._

```{r}
# Undo the effect of library(duckplyr)
methods_restore()
```

Here is the function that runs the pipeline with dplyr.

```{r}
tpch_dplyr <- function(lineitem) {
  lineitem |>
    select(l_shipdate, l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
    filter(l_shipdate <= !!as.Date("1998-09-02")) |>
    select(l_returnflag, l_linestatus, l_quantity, l_extendedprice, l_discount, l_tax) |>
    summarise(
      sum_qty = sum(l_quantity),
      sum_base_price = sum(l_extendedprice),
      sum_disc_price = sum(l_extendedprice * (1 - l_discount)),
      sum_charge = sum(l_extendedprice * (1 - l_discount) * (1 + l_tax)),
      avg_qty = mean(l_quantity),
      avg_price = mean(l_extendedprice),
      avg_disc = mean(l_discount),
      count_order = n(),
      .by = c(l_returnflag, l_linestatus)
    ) |>
    arrange(l_returnflag, l_linestatus) 
}
```

Here is the function that runs it with duckplyr.
We can use exactly the same code (by forwarding to the dplyr function) and only need to change the input data to a duck frame.
The `collect()` at the end ensures materialization, otherwise the comparison isn't fair.

```{r}
tpch_duckplyr <- function(lineitem) {
  lineitem |>
    duckplyr::as_duckdb_tibble() |>
    tpch_dplyr() |>
    collect()
}
```

And now we compare the two:

```{r}
bench::mark(
  tpch_dplyr(lineitem_tbl),
  tpch_duckplyr(lineitem_tbl),
  check = ~ all.equal(.x, .y, tolerance = 1e-10)
)
```

In this example, the pipeline run with duckplyr is clearly faster than the pipeline run with dplyr.

## Data larger than memory

For data larger than memory, duckplyr is a legitimate alternative to dtplyr and dbplyr.

With datasets that approach or surpass the size of your machine's RAM, you want:

- input data in an efficient format, like Parquet files, which duckplyr allows thanks to its ingestion functions like `read_parquet_duckdb()`;
- efficient computation, which duckplyr provides via DuckDB's holistic optimization, without your having to adapt your code;
- large results to not clutter your memory by dumping them to files using  [`compute_parquet()`](https://duckplyr.tidyverse.org/reference/compute_parquet.html) or [`compute_csv()`](https://duckplyr.tidyverse.org/reference/compute_csv.html);
- small results processed seamlessly with dplyr, using all verbs and functions.

This workflow is fully supported by duckplyr.
See [`vignette("large")`](https://duckplyr.tidyverse.org/articles/large.html) for a walkthrough and more details.

## Help us improve duckplyr!

Our goals for future development of duckplyr include:

- Enabling users to provide [custom translations](https://github.com/tidyverse/duckplyr/issues/158) of dplyr functionality;
- Making it easier to contribute code to duckplyr;
- Supporting more dplyr and tidyr functionality natively in DuckDB.

You can help!

- Please report any issues, especially regarding unknown incompabilities. See [`vignette("limits")`](https://duckplyr.tidyverse.org/articles/limits.html).
- Contribute to the codebase after reading duckplyr's [contributing guide](https://duckplyr.tidyverse.org/CONTRIBUTING.html).
- Turn on telemetry to help us hear about the most frequent fallbacks so we can prioritize working on the corresponding missing dplyr translation. See [`vignette("telemetry")`](https://duckplyr.tidyverse.org/articles/telemetry.html) and the [`duckplyr::fallback_sitrep()`](https://duckplyr.tidyverse.org/reference/fallback.html) function.

## Additional resources

Eager to learn more about duckplyr -- beside by trying it out yourself?
The pkgdown website of duckplyr features several [articles](https://duckplyr.tidyverse.org/articles/).
Furthermore, the blog post ["duckplyr: dplyr Powered by DuckDB"](https://duckdb.org/2024/04/02/duckplyr.html) by Hannes Mühleisen provides some context on duckplyr including its inner workings, as also seen in a [section](https://blog.r-hub.io/2025/02/13/lazy-meanings/#duckplyr-lazy-evaluation-and-prudence) of the R-hub blog post ["Lazy introduction to laziness in R"](https://blog.r-hub.io/2025/02/13/lazy-meanings/) by Maëlle Salmon, Athanasia Mo Mowinckel and Hannah Frick. 

## Acknowledgements

A big thanks to all folks who filed issues, created PRs and generally helped to improve duckplyr and its workhorse [duckdb](https://r.duckdb.org/)!

[&#x0040;adamschwing](https://github.com/adamschwing), [&#x0040;alejandrohagan](https://github.com/alejandrohagan), [&#x0040;andreranza](https://github.com/andreranza), [&#x0040;apalacio9502](https://github.com/apalacio9502), [&#x0040;apsteinmetz](https://github.com/apsteinmetz), [&#x0040;barracuda156](https://github.com/barracuda156), [&#x0040;beniaminogreen](https://github.com/beniaminogreen), [&#x0040;bob-rietveld](https://github.com/bob-rietveld), [&#x0040;brichards920](https://github.com/brichards920), [&#x0040;cboettig](https://github.com/cboettig), [&#x0040;davidjayjackson](https://github.com/davidjayjackson), [&#x0040;DavisVaughan](https://github.com/DavisVaughan), [&#x0040;Ed2uiz](https://github.com/Ed2uiz), [&#x0040;eitsupi](https://github.com/eitsupi), [&#x0040;era127](https://github.com/era127), [&#x0040;etiennebacher](https://github.com/etiennebacher), [&#x0040;eutwt](https://github.com/eutwt), [&#x0040;fmichonneau](https://github.com/fmichonneau), [&#x0040;hadley](https://github.com/hadley), [&#x0040;hannes](https://github.com/hannes), [&#x0040;hawkfish](https://github.com/hawkfish), [&#x0040;IndrajeetPatil](https://github.com/IndrajeetPatil), [&#x0040;JanSulavik](https://github.com/JanSulavik), [&#x0040;JavOrraca](https://github.com/JavOrraca), [&#x0040;jeroen](https://github.com/jeroen), [&#x0040;jhk0530](https://github.com/jhk0530), [&#x0040;joakimlinde](https://github.com/joakimlinde), [&#x0040;JosiahParry](https://github.com/JosiahParry), [&#x0040;kevbaer](https://github.com/kevbaer), [&#x0040;larry77](https://github.com/larry77), [&#x0040;lnkuiper](https://github.com/lnkuiper), [&#x0040;lorenzwalthert](https://github.com/lorenzwalthert), [&#x0040;lschneiderbauer](https://github.com/lschneiderbauer), [&#x0040;luisDVA](https://github.com/luisDVA), [&#x0040;math-mcshane](https://github.com/math-mcshane), [&#x0040;meersel](https://github.com/meersel), [&#x0040;multimeric](https://github.com/multimeric), [&#x0040;mytarmail](https://github.com/mytarmail), [&#x0040;nicki-dese](https://github.com/nicki-dese), [&#x0040;PMassicotte](https://github.com/PMassicotte), [&#x0040;prasundutta87](https://github.com/prasundutta87), [&#x0040;rafapereirabr](https://github.com/rafapereirabr), [&#x0040;Robinlovelace](https://github.com/Robinlovelace), [&#x0040;romainfrancois](https://github.com/romainfrancois), [&#x0040;sparrow925](https://github.com/sparrow925), [&#x0040;stefanlinner](https://github.com/stefanlinner), [&#x0040;szarnyasg](https://github.com/szarnyasg), [&#x0040;thomasp85](https://github.com/thomasp85), [&#x0040;TimTaylor](https://github.com/TimTaylor), [&#x0040;Tmonster](https://github.com/Tmonster), [&#x0040;toppyy](https://github.com/toppyy), [&#x0040;wibeasley](https://github.com/wibeasley), [&#x0040;yjunechoe](https://github.com/yjunechoe), [&#x0040;ywhcuhk](https://github.com/ywhcuhk), [&#x0040;zhjx19](https://github.com/zhjx19), [&#x0040;ablack3](https://github.com/ablack3), [&#x0040;actuarial-lonewolf](https://github.com/actuarial-lonewolf), [&#x0040;ajdamico](https://github.com/ajdamico), [&#x0040;amirmazmi](https://github.com/amirmazmi), [&#x0040;anderson461123](https://github.com/anderson461123), [&#x0040;andrewGhazi](https://github.com/andrewGhazi), [&#x0040;Antonov548](https://github.com/Antonov548), [&#x0040;appiehappie999](https://github.com/appiehappie999), [&#x0040;ArthurAndrews](https://github.com/ArthurAndrews), [&#x0040;arthurgailes](https://github.com/arthurgailes), [&#x0040;babaknaimi](https://github.com/babaknaimi), [&#x0040;bcaradima](https://github.com/bcaradima), [&#x0040;bdforbes](https://github.com/bdforbes), [&#x0040;bergest](https://github.com/bergest), [&#x0040;bill-ash](https://github.com/bill-ash), [&#x0040;BorgeJorge](https://github.com/BorgeJorge), [&#x0040;brianmsm](https://github.com/brianmsm), [&#x0040;chainsawriot](https://github.com/chainsawriot), [&#x0040;ckarnes](https://github.com/ckarnes), [&#x0040;clementlefevre](https://github.com/clementlefevre), [&#x0040;cregouby](https://github.com/cregouby), [&#x0040;cy-james-lee](https://github.com/cy-james-lee), [&#x0040;daranzolin](https://github.com/daranzolin), [&#x0040;david-cortes](https://github.com/david-cortes), [&#x0040;DavZim](https://github.com/DavZim), [&#x0040;denis-or](https://github.com/denis-or), [&#x0040;developertest1234](https://github.com/developertest1234), [&#x0040;dicorynia](https://github.com/dicorynia), [&#x0040;dsolito](https://github.com/dsolito), [&#x0040;e-kotov](https://github.com/e-kotov), [&#x0040;EAVWing](https://github.com/EAVWing), [&#x0040;eddelbuettel](https://github.com/eddelbuettel), [&#x0040;edward-burn](https://github.com/edward-burn), [&#x0040;elefeint](https://github.com/elefeint), [&#x0040;eli-daniels](https://github.com/eli-daniels), [&#x0040;elysabethpc](https://github.com/elysabethpc), [&#x0040;erikvona](https://github.com/erikvona), [&#x0040;florisvdh](https://github.com/florisvdh), [&#x0040;gaborcsardi](https://github.com/gaborcsardi), [&#x0040;ggrothendieck](https://github.com/ggrothendieck), [&#x0040;hdmm3](https://github.com/hdmm3), [&#x0040;hope-data-science](https://github.com/hope-data-science), [&#x0040;IoannaNika](https://github.com/IoannaNika), [&#x0040;jabrown-aepenergy](https://github.com/jabrown-aepenergy), [&#x0040;JamesLMacAulay](https://github.com/JamesLMacAulay), [&#x0040;jangorecki](https://github.com/jangorecki), [&#x0040;javierlenzi](https://github.com/javierlenzi), [&#x0040;Joe-Heffer-Shef](https://github.com/Joe-Heffer-Shef), [&#x0040;kalibera](https://github.com/kalibera), [&#x0040;lboller-pwbm](https://github.com/lboller-pwbm), [&#x0040;lgaborini](https://github.com/lgaborini), [&#x0040;m-muecke](https://github.com/m-muecke), [&#x0040;meztez](https://github.com/meztez), [&#x0040;mgirlich](https://github.com/mgirlich), [&#x0040;mtmorgan](https://github.com/mtmorgan), [&#x0040;nassuphis](https://github.com/nassuphis), [&#x0040;nbc](https://github.com/nbc), [&#x0040;olivroy](https://github.com/olivroy), [&#x0040;pdet](https://github.com/pdet), [&#x0040;phdjsep](https://github.com/phdjsep), [&#x0040;pierre-lamarche](https://github.com/pierre-lamarche), [&#x0040;r2evans](https://github.com/r2evans), [&#x0040;ran-codes](https://github.com/ran-codes), [&#x0040;rplsmn](https://github.com/rplsmn), [&#x0040;Saarialho](https://github.com/Saarialho), [&#x0040;SimonCoulombe](https://github.com/SimonCoulombe), [&#x0040;tau31](https://github.com/tau31), [&#x0040;thohan88](https://github.com/thohan88), [&#x0040;ThomasSoeiro](https://github.com/ThomasSoeiro), [&#x0040;timothygmitchell](https://github.com/timothygmitchell), [&#x0040;vincentarelbundock](https://github.com/vincentarelbundock), [&#x0040;VincentGuyader](https://github.com/VincentGuyader), [&#x0040;wlangera](https://github.com/wlangera), [&#x0040;xbasics](https://github.com/xbasics), [&#x0040;xiaodaigh](https://github.com/xiaodaigh), [&#x0040;xtimbeau](https://github.com/xtimbeau), [&#x0040;yng-me](https://github.com/yng-me), [&#x0040;Yousuf28](https://github.com/Yousuf28), [&#x0040;yutannihilation](https://github.com/yutannihilation), and [&#x0040;zcatav](https://github.com/zcatav)

Special thanks to Joe Thorley ([&#x0040;joethorley](https://github.com/joethorley)) for help with choosing the right words.
