---
output: hugodown::hugo_document

slug: duckplyr-1-0-0
title: duckplyr fully joins the tidyverse!
date: 2025-02-11
author: Kirill Müller and Maëlle Salmon
description: >
    duckplyr 1.0.0 is on CRAN and part of the tidyverse!
    A drop-in replacement for dplyr, powered by DuckDB for speed.
    It is the most dplyr-like of dplyr backends.

photo:
  url: https://www.pexels.com/photo/a-mallard-duck-on-water-6918877/
  author: Kiril Gruev

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package]
tags:
  - duckplyr
  - dplyr
  - tidyverse
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're very chuffed to announce the release of [duckplyr](https://duckplyr.tidyverse.org) 1.0.0.
duckplyr is a drop-in, fully compatible replacement for dplyr, powered by [DuckDB](https://duckdb.org/) for speed.
It joins the rank of dplyr backends together with [dtplyr](https://dtplyr.tidyverse.org) and [dbplyr](https://dbplyr.tidyverse.org).
You can use it instead of dplyr for data small or large.

<!-- FIXME:

We have many more dplyr backends, the two above are just from the tidyverse.
GitHub search: https://github.com/search?q=org%3Acran+%2FS3method%5B%28%5D%28mutate%7Csummarise%29+*%2C%2F&type=code
Do we need an "awesome dplyr" like https://github.com/krlmlr/awesome-vctrs/?

-->

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("duckplyr")
```

In this article, we'll introduce you to the basic concepts behind duckplyr, show how it can help you data of different sizes, and explain how you can help improve the package.

## A drop-in replacement for dplyr

The duckplyr package is a _drop-in replacement for dplyr_ that uses _DuckDB for speed_.
You can simply _drop_ duckplyr into your pipeline by loading it, then computations will be efficiently carried out by DuckDB.
DuckDB is a fast in-memory analytical database system.
If you haven't heard about it, watch [Hannes Mühleisen's keynote at posit::conf(2024)](https://www.youtube.com/watch?v=GELhdezYmP0&feature=youtu.be).

```{r}
library(conflicted)
library(duckplyr)
conflict_prefer("filter", "dplyr", quiet = TRUE)
library(babynames)

out <- babynames |>
  mutate(is_frequent = (prop >= 0.01)) |>
  summarize(
    .by = c(sex, year, is_frequent),
    babies_n = sum(n)
  ) |>
  filter(sex == "F")
class(out)
```

Like with other dplyr backends like dtplyr and dbplyr, duckplyr allows you to get faster results.
Unlike other dplyr backends, duckplyr does not require you to learn a different syntax.
The duckplyr package is fully compatible with dplyr: if an operation cannot be carried out with DuckDB, it is automatically outsourced to dplyr.
Over time, we expect fewer and fewer fallbacks to dplyr to be needed.

## How to use duckplyr

To _replace_ dplyr with duckplyr, you can either

- load duckplyr and then keep your pipeline as is. Calling `library(duckplyr)` overwrites dplyr methods, enabling duckplyr for the entire session no matter how data.frames are created.

```{r}
library(conflicted)
library(duckplyr)
conflict_prefer("filter", "dplyr", quiet = TRUE)
```

- Create individual "duck frames" which allows you to control their automatic materialization parameters to [protect memory](https://duckplyr.tidyverse.org/articles/prudence.html). To do so, you can use _conversion functions_ like `duckdb_tibble()` or `as_duckdb_tibble()`, or _ingestion functions_ like `read_csv_duckdb()`.

Then, the data manipulation pipeline uses the exact same syntax as a dplyr pipeline.
The duckplyr package performs the computation using DuckDB.

```{r}
library(babynames)
out <- babynames |>
  filter(n > 1000) |>
  summarize(
    .by = c(sex, year),
    babies_n = sum(n)
  ) |>
  filter(sex == "F")
```

The result can finally be materialized to memory, or computed temporarily, or computed to a file.

```{r}
# to memory
out

# to a file
csv_file <- withr::local_tempfile()
file.size(csv_file)
compute_csv(out, csv_file)
file.size(csv_file)
```

When duckplyr itself does not support specific functionality, it falls back to dplyr.
For instance, row names are not supported yet:

```{r}
mtcars |>
  summarize(
    .by = cyl,
    disp = mean(disp, na.rm = TRUE),
    sd = sd(disp, na.rm = TRUE)
  )
```

Current limitations are documented in a [vignette](https://duckplyr.tidyverse.org/articles/limits.html).
You can change the verbosity of fallbacks, refer to [`duckplyr::fallback_sitrep()`](https://duckplyr.tidyverse.org/reference/fallback.html).

### For large data

For large data, duckplyr is a legitimate alternative to dtplyr and dbplyr.

With large datasets, you want:

- input data in an efficient format, like Parquet files, which duckplyr allows thanks to its ingestion functions like `read_parquet_duckdb()`.
- efficient computation, which duckplyr provides via DuckDB's holistic optimization, without your having to use another syntax than dplyr.
- the output to not clutter all the memory, which duckplyr supports through two features:
    - the [control of automatic materialization](https://duckplyr.tidyverse.org/articles/prudence.html) (collection of results into memory) thanks to the `prudence` parameter. You can disable automatic materialization completely or, as a compromise, disable it up to a certain output size.
    - [computation to files](https://duckplyr.tidyverse.org/reference/compute_file.html) using  `compute_parquet()` or `compute_csv()`.

A drawback of analyzing large data with duckplyr is that the limitations of duckplyr won't be compensated by fallbacks, since fallbacks to dplyr necessitate putting data into memory.
Therefore, if your pipeline encounters fallbacks, you might want to work around them by converting the duck frame into a table through `compute()` then running SQL code through the experimental `read_sql_duckdb()` function. Again, over time, we expect more native support for dplyr functionality.

```{r}
data <-
  duckdb_tibble(a = 2) |>
  mutate(b = 3)

computed_data <-
  data |>
  compute(name = "computed_data")

sql_data <-
  read_sql_duckdb("SELECT *, a * b AS c FROM computed_data")

sql_data
```

## Help us improve duckplyr!

Our goals for future development of duckplyr include:

- Increasing the native support for dplyr functionality;
- Enabling users to provide [custom translations](https://github.com/tidyverse/duckplyr/issues/158) of dplyr functionality;
- Making it easier to contribute code to duckplyr.

You can help!

- Please report any issue especially regarding unknown incompabilities. See [`vignette("limits")`](https://duckplyr.tidyverse.org/articles/limits.html).
- Contribute to the codebase after reading duckplyr's [contributing guide](https://duckplyr.tidyverse.org/CONTRIBUTING.html).
- Turn on telemetry to help us hear about the most frequent fallbacks so we can prioritize working on the corresponding missing dplyr translation. See [`vignette("telemetry")`](https://duckplyr.tidyverse.org/articles/telemetry.html) and the [`duckplyr::fallback_sitrep()`](https://duckplyr.tidyverse.org/reference/fallback.html) function.

## Acknowledgements

A big thanks to all 54 folks who filed issues, created PRs and generally helped to improve duckplyr!

[&#x0040;adamschwing](https://github.com/adamschwing), [&#x0040;andreranza](https://github.com/andreranza), [&#x0040;apalacio9502](https://github.com/apalacio9502), [&#x0040;apsteinmetz](https://github.com/apsteinmetz), [&#x0040;barracuda156](https://github.com/barracuda156), [&#x0040;beniaminogreen](https://github.com/beniaminogreen), [&#x0040;bob-rietveld](https://github.com/bob-rietveld), [&#x0040;brichards920](https://github.com/brichards920), [&#x0040;cboettig](https://github.com/cboettig), [&#x0040;davidjayjackson](https://github.com/davidjayjackson), [&#x0040;DavisVaughan](https://github.com/DavisVaughan), [&#x0040;Ed2uiz](https://github.com/Ed2uiz), [&#x0040;eitsupi](https://github.com/eitsupi), [&#x0040;era127](https://github.com/era127), [&#x0040;etiennebacher](https://github.com/etiennebacher), [&#x0040;eutwt](https://github.com/eutwt), [&#x0040;fmichonneau](https://github.com/fmichonneau), [&#x0040;github-actions[bot]](https://github.com/github-actions[bot]), [&#x0040;hadley](https://github.com/hadley), [&#x0040;hannes](https://github.com/hannes), [&#x0040;hawkfish](https://github.com/hawkfish), [&#x0040;IndrajeetPatil](https://github.com/IndrajeetPatil), [&#x0040;JanSulavik](https://github.com/JanSulavik), [&#x0040;JavOrraca](https://github.com/JavOrraca), [&#x0040;jeroen](https://github.com/jeroen), [&#x0040;jhk0530](https://github.com/jhk0530), [&#x0040;joakimlinde](https://github.com/joakimlinde), [&#x0040;JosiahParry](https://github.com/JosiahParry), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;larry77](https://github.com/larry77), [&#x0040;lnkuiper](https://github.com/lnkuiper), [&#x0040;lorenzwalthert](https://github.com/lorenzwalthert), [&#x0040;luisDVA](https://github.com/luisDVA), [&#x0040;maelle](https://github.com/maelle), [&#x0040;math-mcshane](https://github.com/math-mcshane), [&#x0040;meersel](https://github.com/meersel), [&#x0040;multimeric](https://github.com/multimeric), [&#x0040;mytarmail](https://github.com/mytarmail), [&#x0040;nicki-dese](https://github.com/nicki-dese), [&#x0040;PMassicotte](https://github.com/PMassicotte), [&#x0040;prasundutta87](https://github.com/prasundutta87), [&#x0040;rafapereirabr](https://github.com/rafapereirabr), [&#x0040;Robinlovelace](https://github.com/Robinlovelace), [&#x0040;romainfrancois](https://github.com/romainfrancois), [&#x0040;sparrow925](https://github.com/sparrow925), [&#x0040;stefanlinner](https://github.com/stefanlinner), [&#x0040;thomasp85](https://github.com/thomasp85), [&#x0040;TimTaylor](https://github.com/TimTaylor), [&#x0040;Tmonster](https://github.com/Tmonster), [&#x0040;toppyy](https://github.com/toppyy), [&#x0040;wibeasley](https://github.com/wibeasley), [&#x0040;yjunechoe](https://github.com/yjunechoe), [&#x0040;ywhcuhk](https://github.com/ywhcuhk), and [&#x0040;zhjx19](https://github.com/zhjx19).
