---
title: testthat 2.0.0
author: Hadley Wickham
date: '2017-11-15'
slug: testthat-2-0-0
categories:
  - package
tags: []
---

```{r setup, include = FALSE}
library(testthat)
```

We are very excited to announce that testthat 2.0.0 is now available on CRAN! Testthat makes it easy to turn your existing informal tests into formal automated tests that you can rerun quickly and easily. Learn more at <http://r-pkgs.had.co.nz/tests.html>. Install the latest version with:

```{r, eval = FALSE}
install.packages("testthat")
```

testthat 2.0.0 is a massive release containing a bunch of new features. 

In addition, as part of a general process to make tidyverse and r-lib packages available more widely, we are now testing testthat on R 3.1, 3.2, 3.3, 3.4, and 3.5. We hope this makes it easier to use our packages in slower moving environments.



* New default reporter
* New and improved expectations
* Quasiquotation support
* More setup and teardown tools

Unfortunately along with these new features we had to make a few breaking changes to the API. These affected around 1 in 40 packages using testthat, and are described at the end of this post.

## New default reporter

A new default reporter, `ReporterProgress`, revamps the output to use colour more effectively and reveals details of failures as they occur rather than only showing at the end:

INSERT GIF HERE

If you prefer the previous version, you can turn back time by setting `option(testthat.default_reporter = "summary")`.

## Setup and teardown

There are two new ways to run code before and after tests

* `setup()` and `teardown()` functions allow you to run at the start and
  end of each test file.

* `setup-xyz.R` files are run before all tests (and unlike the existing 
  `helpers-xyz.R` are not run by `devtools::load_all()`)); 
  `teardown-xyz.R` are run after all tests.
  
## New and improved expectations

We have identified a useful family of expectations that compares the results of an expression to a known good value stored in a file. They are designed to be use in conjunction with git so that you can see what precisely has changed, and revert it if needed.

* `expect_known_output()` replaces `expect_output_file()`, which has
  been soft-deprecated. It now defaults to `update = TRUE` and warn, rather
  than failing on the first run. It gains a `print` argument to automatically 
  print the input (#627). It also sets the width option to 80 to ensure 
  consistent output across environments (#514)

* `expect_known_value()` replaces `expect_equal_to_reference()`, which
  has been soft-deprecated. It gains an update argument defaulting to `TRUE`.
  This changes behaviour from the previous version, and soft-deprecated
  `expect_equal_to_reference()` gets `update = FALSE`. 

We've also improved tools for failures:

* `expect_known_failure()` stores and compares the failure message from
  an expectation. It's a useful regression test when developing informative
  failure messges for your own expectations.

* `expect_condition()` works like `expect_error()` but captures any
  condition, not just error conditions (#621).

*  `expect_error()` gains a `class` argument that allows you to make an 
  assertion about the class of the error object (#530).

* `expect_setequal()` compares two sets (stored in vectors), ignoring
  duplicates and differences in order (#528).

And a few new skip helperes

* `skip_if()` makes it easy to skip a test when a condition is true (#571).
  For example, use `skip_if(getRversion() <= 3.1)` to skip a test in older
  R versions.

* `skip_if_translated()` skips tests if you're running in an locale
  where translations are likely to occur (#565). Use this to avoid
  spurious failures when checking the text of error messages in non-English
  locales.
  
* `skip_if_not_installed()` gains new `minimum_version` argument. This allows
  you to only test if a minimum version requrement is met:
  `skip_if_not_installed("ggplot2", "2.0.0")`

## Quasiquotation support

All expectations can now use [unquoting](???). This makes it much easier to generate informative failure messages when running tests in a for loop. For example take this test:

```{r, error = TRUE}
f <- function(i) if (i > 3) i * 9 else i * 10

for (i in 1:5) {
  expect_equal(f(i), i * 10)
}
```

The error message is not great because you don't know which iteration caused the problem! You can resolve that problem by using the unquoting operator `!!` (pronounced bang-bang):

```{r, error = TRUE}
for (i in 1:5) {
  expect_equal(f(!!i), !!(i * 10))
}
```

(Note that this is not tidy evaluation per se, but is closely related. This works a little differently in testthat to tidyverse package because the only quoting involved is for error message. See `?quasi_label()` for more details. At this time you can not unquote quosures.)

## Breaking API changes

Unfotunately

* "Can't mock functions in base packages": You can no longer use `with_mock()` 
  to mock functions in base packages, because this no longer works in 
  R-devel due to changes with the byte code compiler. I recommend using
  [mockery](https://github.com/n-s-f/mockery) instead.

* The order of arguments to `expect_equivalent()` and `expect_error()` has
  changed slightly as both now pass `...` on another function. This reveals
  itself with a number of different errors, like:
  
    * 'what' must be a character vector
    * 'check.attributes' must be logical
    * 'tolerance' should be numeric
    * argument is not interpretable as logical
    * threw an error with unexpected class
    * argument "quo" is missing, with no default
    * argument is missing, with no default
    
    If you see one of these errors, check the number, order, and names of 
    arguments to the expectation.

* "Failure: (unknown)". The last release mistakenly failed to test 
  bare expectations not wrapped inside `test_that()`. If you see "(unknown)"
  in a failure message, this is a failing expectation that you previously
  weren't seeing. As well as fixing the failure, please also wrap inside
  a `test_that()` with an informative name.
  
* "Error: the argument has already been evaluated": the way in which 
  expectations now need create labels has changed, which caused a couple 
  of failures with unusual usage when combined with `Reduce`, `lapply()`, 
  and `Map()`. Avoid these functions in favour of for loops. I also recommend
  reading the section below on quasiquotation support in order to create more 
  informative failure messages.
  
* `is_null()` and `matches()` have been deprecated because they conflict
  with other functions in the tidyverse. As far as we could tell, no
  CRAN packages used these functions

## Acknowledgements

A big thanks goes to [Kirill Müller](https://github.com/krlmlr) for his help running R CMD check on all the packages that use testthat - in total he ran R CMD check over 10,000 times! 
