---
title: styler 1.0.0
author: Lorenz Walthert, Kirill Müller
date: '2017-12-14'
slug: styler-1.0.0
categories:
  - package
photo:
  url: https://unsplash.com/photos/mpdIPhYqZ4Y
  author: Heng Films
description: >
  styler, a package that can format your code according to the tidyverse style guide, is now on CRAN.

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, promt = TRUE)
knitr::knit_engines$set(list(
  styler = function(options) {
    options$comment <- ""
    knitr::engine_output(
      options,
      
      c("# Before", options$code),
      c("# After", styler::style_text(options$code))
    )
  }
))

```


We're pleased to announce the release of [styler 1.0.0](https://github.com/r-lib/styler/releases/tag/v1.0.0). styler is a source code formatter - a package to format R code according to a style guide. We've implemented the [tidyverse style guide](http://style.tidyverse.org), so you can format your code according to it. A coherent style is important for consistency and legibility. Just as it is important to putspcesbetweenwords. 
You can install styler from CRAN:

```{r, eval = FALSE}
install.packages("styler")
```

styler can style text, single files, packages and entire
R source trees with the following functions:

* `style_text()` styles a character vector.
* `style_file()` styles R and Rmd files.
* `style_dir()` styles all R and/or Rmd files in a directory.
* `style_pkg()` styles the source files of an R package.
* An RStudio Addin that styles the active R or Rmd file, the current 
  package or the highlighted code.

## styling options

A distinguishing feature of styler is its flexibility. The following options are available:

* `scope`. What is to be styled, e.g. just spacing or also indention, line breaks, tokens?
* `strict`. How strict should be rules be applied?
* `math_token_spacing`. Granular control for spacing around math tokens.
* `reindention`. Allows to set the indention of certain regex patterns manually.

We will briefly describe all of them below.

### scope

`scope` can take one of the following values:

* "none": Performs no transformation at all.
* "spaces": Manipulates spacing between token on the same line.
* "indention": In addition to "spaces", this option also manipulates the indention level.
* "line_breaks": In addition to "indention", this option also manipulates line breaks.
* "tokens": In addition to "line_breaks", this option also manipulates tokens.

Hence, `scope < "tokens"` will never change code in the sense that the expression remains the same before and after styling, as only spaces and line breaks between the tokens are modified. styler can internally validate the styling, so you can be sure that it has not broken your code. Although extensively tested, we recommend to inspect the styling result with `scope = "tokens"`, since in this case, an automatic validation is not possible.

Let's now see a few examples. We can limit ourselves to styling just spacing information:

```{r}
library("styler")
style_text("a=3; 2", scope = "spaces")
```

Or, on the other extreme of the scale, styling spaces, indention, line breaks 
and tokens:

```{r}
style_text("a=3; 2", scope = "tokens")
```

In the next example, we want to style spaces and indention, but not line breaks
and tokens:
```{r}
style_text(
  "a=  function( x) {
  'x'
  }", 
  scope = "indention"
)
```


### strict

Another option that is helpful to determine the level of 'invasiveness' is `strict`. If set to `TRUE`, spaces and line breaks before or after tokens are set to either zero or one. However, in some situations, this might be undesirable (so we set `strict = FALSE`), as the following example shows:

```{r}
style_text(
  "data_frame(
     small  = 2 ,
     medium = 4,#comment without space
     large  =6
   )", 
  strict = FALSE
)
```

We prefer to keep the equal sign after "small", "medium" and large aligned, so we set `strict = FALSE` to set spacing to *at least* one around `=`.

Also, with `scope >= "line_breaks"`, `strict = TRUE` will break the lines in the following call, whereas `strict = FALSE` will leave them as is:
```{r}
style_text(
  "do_a_long_and_complicated_fun_cal('which', has, way, to, 
                              'and longer then lorem ipsum in its full length'
                              )"
)
```


In addition, the `strict` option also adds braces (if `scope = "tokens"`) to conditional statements if they are multi-line:
```{r}
style_text(
  "if (set_q(x))
    3 else 
    5",
  strict = TRUE
)
```


### math token spacing

styler can identify and handle unary operators and other math tokens:

```{styler}
1++1-1-1/2
```

This is tidyverse style. However, styler offers very granular control for math token spacing. Assuming you like spacing around `+` and `-`, but not around `/` and `*` and `^`. This can be achieved as follows:
```{r}
style_text(
  "1++1/2*2^2", 
  math_token_spacing = specify_math_token_spacing(zero = c("'/'", "'*'", "'^'"))
)
```

### reindention

You can set the indention of lines that match a certain regular expression manually. For example, if you don't want comments starting with `###` to be indented, you can formulate an unindention rule:
```{r}
style_text(
    "a <- function() {
    ### not to be indented
    # indent normally
    33
    }",
  reindention = specify_reindention(regex_pattern = "^###", indention = 0)
)
```


These are options that allow flexible reformatting. You can even go further, as described in the ["Customizing styler"](http://styler.r-lib.org/articles/customizing_styler.html) vignette.

# A few more examples

Now that you know that the rules can be applied in a flexible fashion, let's focus on `scope = "tokens"` and `strict = TRUE` and illustrate how specific rules of the tidyverse style guide are implemented. 

Function declarations are indented if multi-line:
```{styler}
my_fun <- function(x, 
y, 
z) {
  just(z)
}
```

styler can also format complicated expressions that involve line breaking and indention based on both brace expressions and operators:

```{styler}
if (x >3) {stop("this is an error")} else {
c(there_are_fairly_long,
1 / 33 * 
2 * long_long_variable_names)%>% k(

) }
```

Lines are broken after `(` if a function call spans multiple lines:

```{styler}
do_a_long_and_complicated_fun_cal("which", has, way, to, 
                              "and longer then lorem ipsum in its full length"
                              )
```

styler replaces `=` with `<-` for assignment if necessary:
```{styler}
one = 3
data_frame(a = 3)
```

It also handles single quotes according to the tidyverse style guide:
```{styler}
'one string'
'one string in a "string"'
"one string in a 'string'"

```

Braces are added to function calls in pipes:

```{styler}
a %>%
  b %>%
  c
```

tidyeval syntax is handled properly:
```{styler}
mtcars %>%
  group_by(!!!my_vars)
```

That was a quick overview of styler. Please feel free to grab styler from CRAN, run it on your own package and tell us what you think.
