---
title: Erratum tidyr 0.7.0
slug: erratum-tidyr-0.7.0
description: >
  We have updated tidyselect to revert a behaviour introduced in tidyr 0.7.0.
author: Lionel Henry
date: 2017-09-04
photo:
  url: https://unsplash.com/photos/0vY082Un2pk
  author: Edu Grande
categories: [packages]
---

```{r setup, include = FALSE}
library("tidyr")
library("dplyr")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

In [tidyr 0.7.0](http://www.tidyverse.org/articles/2017/08/tidyr-0.7.0/),
we introduced a stronger distinction between data expressions and
context expressions for selecting verbs like `gather()`. However that
change caused a lot of trouble and confusion and we have updated
`tidyselect` (the backend for selection functions) to revert that
behaviour. In that article, we provide a few comments on these changes
as well as some notes on good practices for writing functions with
tidyverse tools. Finally we introduce two new selection features that
help write safe code: improved support for strings and character
vectors and a new selection helper `last_col()`.

You can install the new version of tidyselect from CRAN:

```{r, eval = FALSE}
install.packages("tidyselect")
```


### Updated selection rules

Since tidyr 0.7.0, selecting functions like `gather()` use the
tidyselect package as backend. Tidyselect was extracted from dplyr and
provides the mechanism for helpers like `starts_with()` or
`everything()`. However, tidyselect had one big change compared to
dplyr: data expressions could no longer refer to contextual
variables. A **data** expression is defined as either a bare symbol
like `var`, or a call like `var1:var2` or `c(var1, var2)`. Any other
expressions is a **context** expression. The semantic change meant
that it was no longer legal to refer to a contextual variable in a
data expression:

```{r, eval = FALSE}
var <- 5
mtcars %>% gather("key", "value", 1:var)
```

We thought this was a relatively uncommon occurrence in practice.
However that broke a lot of code that had this form:

```{r, eval = FALSE}
df %>% gather("key", "value", 1:ncol(df))
```

Although the change was well-intentioned, it proved to be too
disruptive and we have reverted it.

Note that we still maintain a distinction between data and context
expressions. The notion of context expression (anything that's not a
symbol or a call to `:` or `c()`) was introduced in dplyr 0.7.0. Since
that version, context expressions cannot refer to the data. This makes
it safer to refer to other objects.


### Safety versus convenience

Most datawise functions in R obey the same set rules regarding the
visibility of objects. Visibility is hierarchical: data frame objects
override those found in the context. This is convenient for
interactive use and scripts, but can also cause issues. In the
following example, should `gather()` select the first three columns
(using the `x` defined in the global environment), or should it select
the first two columns (using the column named `x`)?

```{r}
x <- 3
df <- tibble(w = 1, x = 2, y = 3)
df %>% gather("key", "value", 1:x)
```

The answer is that it selects the first two variables because `x` is
first found in the data.

In practice, the hierarchical ambiguity is not a big problem when you
use these tools interactively or in scripts. However it becomes
worrying when you're writing reusable functions because you don't know
in advance the variables in the input data frame. For those cases
where ambiguity matters, the tidy eval feature of quasiquotation
allows you to explicitly pick a variable from the context:

```{r}
df %>% gather("key", "value", 1 : (!! x))
```


### The special semantics of selection functions

Selection functions have always been a bit special in the tidyverse.
They don't behave exactly like regular quoting functions. In almost
all quoting functions in R, variables represent a data frame *column*.
That's why expressions like this are natural:

```{r, eval = FALSE}
lm(mass ~ scale(height), data = starwars)
```

Since variables represent actual columns, you can include them in
expressions as if they were actual objects. For instance
`scale(height)` is equivalent to `scale(starwars$height)`. The same is
true for most tidyverse tools, e.g. `dplyr::mutate()`:

```{r, eval = FALSE}
starwars %>% mutate(height = scale(height))
```

However in selection functions, variables do not represent columns but
*column positions*. That is a subtle but important distinction. When
you type `height`, tidyselect actually sees the integer `2`. This
makes sense for several reasons:

* Expressions such as `name:mass` evaluates to `1:3` in a natural way.

* You can select the same column multiple times. For example if you
  supply a selection for the dataset `starwars` with
  `starts_with("s")` and `ends_with("s")`, the variables `species` and
  `starships` would be matched twice. It is easy for tidyselect to
  take the intersection of the two sets of column positions. If the
  sets contained vectors instead, it could not determine whether it
  got two different but identical vectors rather than the same
  vector selected twice.

* Finally and most importantly, if the variables evaluated to the
  column vectors, we would have no information about their names or
  positions, which we need to reconstitute the data frame.


### Safety in selection functions

Given the special semantics of selection functions, we had more
freedom to solve the hierarchical ambiguity of quoting functions.
Indeed, apart from `:` or `c()`, there rarely is any need for
referring to column positions in helpers like `starts_with()` or
`contains()`. For this reason, dplyr 0.7.0 introduced the notion of
context expressions. Data frame columns are no longer in scope in
these calls in order to solve the hierarchical ambiguity. This has the
downside that context expressions in selection functions behave a bit
differently from the rest of the tidyverse, but we gain safety in
exchange.

Given these special semantics, it seemed to make sense to give data
expressions the opposite behaviour and only allow references to the
data. This would solve the ambiguity in the opposite direction.  As
explained above, this broke too much code. We had to change it back
and the issue of hierarchical ambiguity along with it.

Luckily tidyselect 0.2.0 also introduces a few features that help
writing safer code for data expressions. First, the support for
strings and character vectors has been improved. All data expressions
fully support strings. It is now valid to supply strings to `-` and `:`:

```{r}
starwars %>% gather("key", "value", "name" : "films")
starwars %>% gather("key", "value", -"height")
```

Note that this only applies to data expressions and it would not make
sense to write `seq("name", "mass")`. Also, it only makes sense to
support strings in this way because of the special nature of selection
functions. This wouldn't work with `mutate()` or `lm()` since they
wouldn't be able to differentiate between a column name or an actual
column (by recycling the string to column length if the data frame has
more than one row).

The purpose of supporting strings in selection function is to make it
easier to unquote column names. Excluding columns with quasiquotation
is now as simple as this:

```{r}
x <- "height"
starwars %>% gather("key", "value", -(!! x))
```

The second feature introduced in tidyselect 0.2.0 is the `last_col()`
helper. We noticed in bug reports that many people use variants of:

```{r}
x <- starwars
x %>% gather("key", "value", 3 : ncol(x))
```

That is potentially unsafe in functions since the data frame might
contain a column named `x`. You can now use `last_col()` instead:

```{r}
# Importing last_col() because it's not exported in dplyr yet
last_col <- tidyselect::last_col

x %>% gather("key", "value", 3 : last_col())
```
