---
title: purrr 0.2.3
slug: purrr-0.2.3
description: >
  A new version of purrr is on CRAN! It features a new family of
  generic mappers, a tool for tidy plucking of deep data structures,
  and many other features and fixes.
author: Lionel Henry
date: 2018-08-09
photo:
  url: https://unsplash.com/photos/TAN1KpDS7Rg
  author: Erika Lanpher
categories: [packages]
---

```{r setup, include = FALSE}
library("purrr")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

We are pleased to announce that purrr 0.2.3 is now on CRAN! Despite
the small increment in the version number (our mistake, it should have
been 0.3.0), this is a rather major release that includes many new
features and bug fixes.

Install the latest version of purrr with:

```{r, eval = FALSE}
install.packages("purrr")
```

You can read about the complete set of changes
at <https://github.com/tidyverse/purrr/releases/tag/v0.2.3>. In this
blog post, I will present the highlights of this version which are a
family of generic mappers, a new `pluck()` function for accessing deep
elements in a readable way, and a few nifty features and improvements.


### Generic mapping

The new `modify` family of functions introduces genericity to mapping.
This means it is now easier to map functions over vector-like S3
objects. The reason is twofold:

* `map()` and friends apply functions on the elements of the
  underlying data structure but that is not always appropriate. An
  object sometimes contains metadata. A user of such a class probably
  wants to map over the *elements* of the data of interest rather than
  over the metadata fields.

* `map()` is type-stable and always returns a list. If you're mapping
  over an object, chances are you want a similar object back.

`modify()` and its variants solve both these problems via S3
dispatch. For instance, let's try a conditional map over a data
frame. Since `map_if()` is type-stable, it returns a list and we lose
the data frame information:

```{r}
iris %>% map_if(is.factor, as.character) %>% str()
```

On the other hand, `modify_if()` is generic and returns a data frame:

```{r}
iris %>% modify_if(is.factor, as.character) %>% str()
```

The flip side of genericity is that the vectors returned by the mapped
function should obey the constraints of the container type. For
instance, data frames require vectors of equal size and it wouldn't be
appropriate to use `modify()` on a function that returns vectors of
variable sizes. In this case you should use `map()`:

```{r}
map(mtcars, unique) %>% str()
```

As a developer there are two ways to make your class compatible with
`modify()` . The easiest is to implement a method for the
subset-assign operator `[<-` which should be sufficient in most
cases. Alternatively you can implement methods for the `modify`
functions themselves as they are all generics.


### pluck() for deep subsetting

The plucking mechanism used for indexing into data structures with
`map()` has been extracted into the function `pluck()`. Plucking is
often more readable when extracting an element buried in a deep data
structure. Compare this base R code which reads non-linearly:

```{r, eval = FALSE}
accessor(x[[1]])$foo
```

To the equivalent pluck:

```{r, eval = FALSE}
x %>% pluck(1, accessor, "foo")
```

The new function `attr_getter()` generates attribute accessors that
can be used in `pluck()`:

```{r}
dfs <- list(iris, mtcars)
dfs %>% pluck(2, attr_getter("row.names"))
```

Or in mapped indexing:

```{r}
x <- list(
  list(vec = letters, df = mtcars),
  list(vec = LETTERS, df = iris)
)
x %>% map(list("df", attr_getter("row.names")))
```


### set_names() is much more flexible

`set_names()` (now reexported from the rlang package) has become much
more flexible. It now takes `...` arguments and concatenates them into
one vector of names. This saves a bit of typing as you don't have to
concatenate explicitly with `c()`. The code also looks a bit leaner:

```{r}
x <- letters[1:3] %>% set_names("foo", "bar", "baz")
x
```

In addition `set_names()` now accepts functions as its first
argument. The function is applied to existing names. Let's transform
the names of our new vector to uppercase with `base::toupper()`:

```{r}
x %>% set_names(toupper)
```

When you supply a function, the `...` arguments are forwarded to the
function which is often handy. Here we might want to pass further
arguments to `base::paste()`:

```{r}
x %>% set_names(paste, "suffix", sep = "_")
```


### Reducing with a three-argument function

`reduce2()` makes it possible to reduce with a three-argument reducing
function. `reduce2()` takes two vectors `.x` and `.y`, the first of
which is reduced in the usual manner. The accumulated value is passed
to the reducing function as first argument while the next value is
passed as second argument. `.y` on the other hand is mapped, not
reduced. At each reducing round, the next value of `.y` is passed to
the reducing function as third argument.

In the following example we have a binary paster that takes `sep` as
third argument. With the ordinary `reduce()` we are stuck with a
single separator during the whole reduction:

```{r}
paste2 <- function(x, y, sep) paste(x, y, sep = sep)
x <- letters[1:4]
reduce(x, paste2, sep = ".")
```

If we want to vary the separator for each value of the input vector,
`reduce2()` allows us to pass a second vector containing specific
separators. This auxiliary vector should have one fewer elements than
the reduced vector:

```{r}
seps <- c("-", ".", "_")
reduce2(x, seps, paste2)
```


### Variadic lambda-formulas

You can now refer to arguments by position in lambda-formulas. The
`..1` symbol refers to the first argument, `..2` to the second and so
on. This makes it easier to use functions like `pmap()` with the
formula shortcut:

```{r}
pmap_chr(mtcars, ~paste(..2, ..4, sep = " - "))
```



### API changes

purrr no longer depends on lazyeval or Rcpp (or dplyr, as of the
previous version). This makes the dependency graph of the tidyverse
simpler, and makes purrr more suitable as a dependency of lower-level
packages.

A number of functions have been renamed or deprecated:

* `is_numeric()` and `is_scalar_numeric()` are deprecated because they
  don't test for what you might expect at first sight.

* `as_function()` is now `as_mapper()` because it is a tranformation
  that makes sense primarily for mapping functions, not in general.
  The rlang package has an `as_function()` coercer that is smaller in
  scope.

* The data frame suffix `_df` has been (soft) deprecated in favour of
  `_dfr` to more clearly indicate that it's a row-bind. All variants
  now also have a `_dfc` for column binding. They currently don't
  handle functions returning vectors because of a bug, but they will
  in the next minor version of purrr.

* `cross_n()` has been renamed to `cross()`. The `_n` suffix was
  removed for consistency with `pmap()` (originally called `map_n()`
  at the start of the project) and `transpose()` (originally called
  `zip_n()`). Similarly, `cross_d()` has been renamed to `cross_df()`
  for consistency with `map_df()`.

* `contains()` has been renamed to `has_element()` to avoid conflicts
  with dplyr.

* `at_depth()` has been renamed to `modify_depth()`.


Finally, these functions have been removed from the package:

* The previously deprecated functions `flatmap()`, `map3()`,
  `map_n()`, `walk3()`, `walk_n()`, `zip2()`, `zip3()`, `zip_n()` have
  been removed.

* `order_by()`, `sort_by()` and `split_by()` have been removed.
  `order_by()` conflicted with `dplyr::order_by()` and the complete
  family doesn't feel that useful. Use tibbles instead.
