---
title: tidyr 0.7.0
slug: tidyr-0.7.0
description: >
  The next installment of tidyr is finally on CRAN! This version
  brings tidy eval to a crucial component of the tidyverse workflow.
author: Lionel Henry
date: 2018-08-15
photo:
  url: https://unsplash.com/photos/8tem2WpFPhM
  author: Radek Grzybowski
categories: [packages]
---

```{r setup, include = FALSE}
library("tidyr")
library("dplyr")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

We are happy to announce that tidyr 0.7.0 is now available on CRAN.
This version is mostly about ensuring consistent semantics across the
tidyverse. Part of this is the switch to the new tidyselect package as
selection backend. This update also brings tidy evaluation to make it
easier to program with tidyr. See the (programming with
dplyr)[http://dplyr.tidyverse.org/articles/programming.html] vignette
for an introduction to tidy eval.

Install the latest version of tidyr with:

```{r, eval = FALSE}
install.packages("tidyr")
```

You can read about the complete set of changes
at <https://github.com/tidyverse/tidyr/releases/tag/v0.7.0>.


### New selection rules

Following the switch to tidyselect, selecting functions are now
stricter in their arguments to avoid ambiguous cases. For example,
take `gather()` and its `...` argument. Consider the following code:

```{r}
x <- 3
df <- tibble(w = 1, x = 2, y = 3)
gather(df, "variable", "value", 1:x)
```

Should it select the first three columns (using the `x` defined in the
global environment), or should it select the first two columns (using
the column named `x`)?

To solve this ambiguity, we now make a strict distinction between data
and context expressions. A **data expression** is either a bare name
or an expression of the form `x:y` or `c(x, y)`. In a data expression,
you can only refer to columns from the data frame. Everything else is
a **context expression** in which you can only refer to objects that
you have defined by assigning with `<-`.

In practice this means that you can no longer refer to contextual
objects like this:

```{r, eval = FALSE}
mtcars %>% gather(var, value, 1:ncol(mtcars))

x <- 3
mtcars %>% gather(var, value, 1:x)
mtcars %>% gather(var, value, -(1:x))
```

You now have to be explicit about where to find objects. One way of
being explicit is to use the quasiquotation operator `!!` which will
evaluate its argument early and inline the result:

```{r, eval = FALSE}
mtcars %>% gather(var, value, !! 1:ncol(mtcars))
mtcars %>% gather(var, value, !! 1:x)
mtcars %>% gather(var, value, !! -(1:x))
```

Read more about quasiquotation in the tidy eval section.


### Tidy evaluation

Tidy evaluation is a principled set of tools that allow programming
with quoting functions (also called NSE functions) in a principled
way. It was first introduced in dplyr 0.7.0 and you can learn more
about it in the (programming with
dplyr)[http://dplyr.tidyverse.org/articles/programming.html] vignette.
At its core, tidy evaluation is the combination of two features:
quasiquotation and quosures.

The tidy eval tools live in rlang and many of them are reexported in
dplyr. This includes `quo()`, `enquo()` and `quos()`. In addition,
`rlang::expr()`, `rlang::sym()` and `rlang::syms()` may be useful as
well and will be exported in the next version of dplyr.

```{r}
# Let's import some tidy eval tools that we'll use in examples below
library("dplyr")
sym <- rlang::sym
```

**Quasiquotation** is essential to program with quoting functions. It
refers to the ability of unquoting part of a quoted expression, and
makes it possible to program with the quoting grammars of dplyr and
tidyr. With quasiquotation, you can change what a function "sees".
You'll typically want to unquote a symbol representing a data frame
column with the `!!` operator.

Here, `expand()` sees `vs` and `cyl`:

```{r}
expand(mtcars, vs, cyl)
```

Thanks to quasiquotation, we can change what `expand()` sees by
unquoting the `am` symbol:

```{r}
x <- sym("am")
expand(mtcars, vs, !! x)
```

The second tidy eval feature are **quosures**. A quosure is a special
kind of expression that evaluates in both the data context (so you can
refer to data frame columns) and the original context of the
expression (e.g. a function context, so you can refer to local
variables created there).

```{r}
# For nicer printing
iris <- tibble::as_tibble(iris)

# Let's create a quosure within a local context
quo <- local({
  prefix <- "Sepal"
  quo(starts_with(prefix))
})
```

`prefix` only exist in the local context but the quosure can safely
refer to it

```{r}
quo
```

In tidyr 0.7.0, all functions now support quosures:

```{r}
gather(iris, key, value, !! quo)
```

Typically you'll use quosures to create wrappers around tidyr
functions. To this end you'll need `enquo()` which does two things: it
transforms your function to a dplyr-like quoting function; and it
returns the quoted expression as a quosure. Creating a wrapper
function is often a simple matter of enquosing and unquoting:

```{r}
my_gather <- function(df, expr) {
  quo <- enquo(expr)
  tidyr::gather(df, key, value, !! quo)
}
```

Thanks to the enquosing, you can safely call your wrapper in local
contexts (e.g. within a function) and refer to variables defined
there:

```{r}
local({
  prefix <- "Sepal"
  my_gather(iris, starts_with(prefix))
})
```


### Special tidy evaluation rules

The tidy evaluation implementation of tidyr is a bit special. While
the philosophy of tidy evaluation is that R code should refer to real
objects (from the data frame or from the context), we had to make some
exceptions to this rule for tidyr. The reason is that several
functions accept bare symbols to specify the names of _new_ columns to
create (`gather()` being a prime example). This is not tidy because
the symbol do not represent any actual object.

Our workaround is to capture these arguments using `rlang::quo_name()`
(so they still support quasiquotation and you can unquote symbols or
strings). This workaround is only provided for backward compatibility
as this UI is not consistent with other tidyr functions that require
strings (e.g. the `into` argument of `separate()`). More generally
this type of NSE it is now discouraged in the tidyverse. We now
prefer strings to refer to variables that don't yet exist.


### Breaking changes

* The underscored SE variants are softly deprecated.

* Selecting functions now make a distinction between data expressions
  and context expressions. The latter can refer only to contextual
  objects while the former can refer only to data variables. See above
  for more information.
