---
title: "conflicted: a new approach to resolving ambiguity"
author: Hadley Wickham
date: '2018-06-02'
slug: 2018-06-conflicted
categories:
  - package
tags: []
---

We're stoked to announce the initial release of the [conflicted](https://github.com/r-lib/conflicted#conflicted) package. The goal of conflicted is to provide alternative way of resolving conflicts caused by ambiguous function names. R handles ambiguity by reporting conflicts when you load a package, then having the most recently loaded package win. This can make it hard to detect conflicts, because it's easy to miss the messages since you often load packages at the top of the script, and you don't see a problem until much later. conflicted takes a different approach to resolving ambiguity, instead making every conflict an error and forcing you to always choose which function to use.

Installed conflicted by running:

```{r, eval = FALSE}
install.packages("conflicted")
```

## How do I use it?

conflicted does not export any functions. To use it, you just need to load it:

```{r, error = TRUE}
library(conflicted)
library(dplyr)

filter(mtcars, am & cyl == 8)
```

To resolve conflicts, remove the ambiguity by using `::` (i.e. `dplyr::filter()` or `base::filter()`). To resolve conflicts for your entire session, use `<-`:

```{r}
filter <- dplyr::filter
filter(mtcars, am & cyl == 8)
```

If you want to make this behaviour the default, you can load conflicted in your `~/.Rprofile` (the easiest way to find and edit this file is with `usethis::edit_r_profile()`):

```{r}
if (interactive()) {
  suppressMessages(suppressWarnings(require(conflicted)))
}
```

## How does it work?

Loading conflicted creates a new "conflicted" environment that is attached just after the global environment. This environment contains an active bindings for any object that is exported by multiple packages; the active binding will throw an error message describing how to disambiguate the name.

The conflicted environment also contains bindings for `library()` and `require()`, these  suppress the default conflict reporting (i.e. `warn.conflicts = FALSE`) and ensure that the conflicted environment is updated with any new conflicts. Because they occur in the search path before the base package, they will be called instead of `base::library()` and `base::require()`.   

## What does the future hold?

In my use of conflicted, I have noticed two main pain points:

* When creating the dplyr package I made what I now believe to be a mistake
  and I gave two functions the same name as existing base R functions: 
  `filter()` and `lag()` (I did this because I thought those functions were
  rarely used, but they're actually important to some communities). However,
  it's frustrating to have do `filter <- dplyr::filter()` everytime you
  load dplyr, so I need to figure out someway to allow you to globally 
  prefer either `dplyr::filter()` or `base::filter()`. It's possible 
  `library(tidyverse)` could do this for you, but I'm not sure if that's too
  aggressive or not.

* When we move a function to a different package, we usually leave a function 
  behind that tells you the new location. Ideally, this sort of function would
  never trigger a conflict so there's no additional hassle if you have both
  new and old packages loaded. (Currently the biggest hassle is usethis and
  devtools since so many usethis functions originally lived in devtools). I 
  think the right way to handle this is to set an attribute on deprecated 
  functions so that conflicted knows to ignore them.
  
My goal is to figure out how to eliminate these pain points before the next release.
  
## Acknowledgements 

Thanks to [\@krlmlr](https://github.com/krlmlr) for the basic idea. This code was previously part of the experimental [strict](https://github.com/hadley/strict) package, but I decided improved conflict resolution is useful by itself and worth its own package.

