---
title: dplyr 0.8.0 release candidate
author: Romain François
date: '2018-10-23'
slug: dplyr-0-8-0-release-candidate
description: > 
  dplyr 0.8.0 release candidate
categories:
  - package
tags:
  - dplyr
  - tidyverse
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
```

# Tidy grouping structure

Previous versions of `dplyr` used a messy set of attributes in grouped
tibbles to keep track of the groups and their indices. This has been 
re-organized into a tibble that can be accessed with the new 
`group_data()` function. 

```{r}
iris %>% 
  group_by(Species) %>% 
  group_data()
```

The first columns of that tibble describe the groups in terms of the 
grouping variables, and the last column (always called `.rows`)
is a list of integer vectors identifying the (one-based) indices of 
each group. 

The related function `group_rows()` gives just that last column. 

```{r}
iris %>% 
  group_by(Species) %>% 
  group_rows()
```

# New grouping algorithm

The algorithm behind `group_by()` has been redesigned to better respect factors. 
In previous versions of `dplyr` groups were created to only match the observed data,
which was potentially losing groups when factors are involved. 

The new algorithm respects the levels of the factors that are involved in the 
grouping, by creating groups even if there is no data in that group. 

```{r}
df <- tibble(
  x = c(1,1,2,2), 
  f = factor(c("a", "b", "a", "b"), levels = c("a", "b", "c"))
)
df %>% 
  group_by(f) %>% 
  group_data()
```

Where previous versions of `dplyr` would have created only two groups (for levels `a` and `b`), 
it now creates one group per level, and the group related to the level `c` just happens to be 
empty. 

Groups are still made to match the data on other variables: 

```{r}
df %>% 
  group_by(x) %>% 
  group_data()
```

Consider this tibble using a mix of factors and non factors

```{r}
df <- tibble(
  f1 = factor(c("a", "a", "a", "b", "b"), levels = c("a", "b", "c")), 
  f2 = factor(c("d", "e", "d", "e", "f"), levels = c("d", "e", "f")), 
  x  = c(1, 1, 1, 2, 2), 
  y  = 1:5
)
```

If we group by the factors `f1` and `f2`, we get 9 groups

```
# 9 groups : 3 x 3
df %>% 
  group_by(f1, f2) %>% 
  group_data()
```

When factors and non factors are involved in the grouping, the number of 
groups depends on the order. At each level of grouping, factors are always expanded
to one group per level, but non factors only create groups based on observed data. 

```
# 3 groups
df %>% 
  group_by(f1, x) %>% 
  group_data()
```

In this example, we group by `f1` then `x`. At the first layer, grouping on `f1` creates
two groups. Each of these grouops is then subdivided based on the values of the second 
variable `x`. Since `x` is always 1 when `f1` is `a` the group is not 
further divided. 

The last group, associated with the level `c` of the factor `f1` is empty, and 
consequently has no values for the vector `x`. In that case, `group_by()` uses 
`NA`. 

```
# 6 groups
df %>% 
  group_by(x, f1) %>% 
  group_data()
```

When we group by `x` then `f1` we initially split the data according to `x` which 
gives 2 groups. Each of these two groups is then further divided in 3 groups, 
i.e. one for each level of `f1`. 

# Performance

TODO: 
 - unwind protect

# Less surprising hybrid evaluation

Hybrid evaluation has been entirely redesigned to be more predictable 
and less surprising. 

When we evaluate an expression in `mutate()` or `summarise()`, the hybrid system
first assert that it can entirely evaluate the expression independently with native code, 
if it can it does. 

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(Petal.Length = mean(Petal.Length))
```

The expression `mean(Petal.Length)` is handled by native code because there is an hybrid 
internal version of `mean`, and its first argument is a column of the data, and the type of 
that column can be treated by the internal function. 

If the expression can not be evaluated completely, it gets evaluated using R standard 
evaluation rules in a special environment that makes the columns available (the data mask). 

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(Petal.Length = sum(Petal.Length) / n())
```

The expression `sum(Petal.Length) / n()` is not recognized, so evaluation goes through the 
standard path. For each group, the subset of `Petal.Length` is extracted, then the R function `sum` 
is called on it and this is divided by the result of calling the `n()` function from `dplyr`. 

Previous versions of `dplyr` attempted to partially evaluate `sum(Petal.Length) / n()` with the 
hybrid evaluator, because `sum(Petal.Length)` and `n()` can both independently be handled. This was 
making it hard to obey R semantics because R is too flexible. 

# nest_by, nest_join

TODO

# colwise verbs

TODO 
