---
title: "stringr 1.3.0"
slug: stringr-1-3-0
description: >
    stringr 1.3.0 is now on CRAN.
date: 2018-02-21
author: Mara Averick
photo:
  url: https://unsplash.com/photos/dLTpk6N31Fc
  author: Igor Ovsyannykov
categories: [package]
---

```{r setup, include = FALSE}
library(stringr)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

We are happy to announce that [stringr](http://stringr.tidyverse.org/) 1.3.0
is now on CRAN. stringr provides a cohesive set of functions designed to make
working with strings as easy as possible. For a complete list of changes, please
see the [release notes](https://github.com/tidyverse/stringr/releases/tag/v1.3.0).

## What's new?

*  Two new wrappers from the [glue](http://glue.tidyverse.org/)
package: `str_glue()` and `str_glue_data()`. If you haven't heard of glue, you're in for a treat! Glue lets you easily interpolate data into strings.

    ```{r gluex}
    name <- "Chet Manley"
    str_glue("My name is {name}.")
    ```
    Since stringr is loaded with tidyverse, this means that you can now access glue's functionality without loading     another package.

*  `str_flatten()`, a wrapper for the `stri_flatten()` function, which 
flattens a character vector into a single string. This is equivalent to `paste(x, collapse = "-")` but it is a bit more explicit in your code -- `str_flatten()` always returns a single string.

*  `str_remove()` and `str_remove_all()`, which wrap `str_replace()` and `str_replace_all()` for removing patterns from strings.

*  `str_squish()`, which removes whitespace from the left and right sides of strings, and converts multiple spaces or space-like characters from the middle of strings into a single space.

    ```{r spacejam}
    str_squish("\n\nString \t\nwith all \tthis     \rspace   in it.\n\n")
    ```

## API changes

The long deprecated `str_join()`, `ignore.case()`, and `perl()` functions
have been removed. You may see `Error : object ‘ignore.case’ is not exported by 'namespace:stringr'` during package build.

## Contributors

Thanks to all of the community members who helped with this release by contributing code and filing issues: [\@alanault](https://github.com/alanault), [\@AlunHewinson](https://github.com/AlunHewinson), [\@amhrasmussen](https://github.com/amhrasmussen), [\@andreypanchenko](https://github.com/andreypanchenko), [\@austin3dickey](https://github.com/austin3dickey), [\@bachlaw](https://github.com/bachlaw),  [\@bpasanek](https://github.com/bpasanek), [\@cderv](https://github.com/cderv), [\@ClaytonJY](https://github.com/ClaytonJY), [\@cwickham](https://github.com/cwickham), [\@danielsjf](https://github.com/danielsjf), [\@dchiu911](https://github.com/dchiu911), [\@dhicks](https://github.com/dhicks), [\@dhodge](https://github.com/dhodge), [\@djrobust](https://github.com/djrobust), [\@dracodoc](https://github.com/dracodoc), [\@duarteguilherme](https://github.com/duarteguilherme), [\@etienne-s](https://github.com/etienne-s), [\@ewallace](https://github.com/ewallace), [\@florisvdh](https://github.com/florisvdh), [\@frootor](https://github.com/frootor), [\@geotheory](https://github.com/geotheory), [\@gonulate](https://github.com/gonulate), [\@guiastrennec](https://github.com/guiastrennec),  [\@hongshaoyang](https://github.com/hongshaoyang), [\@IndrajeetPatil](https://github.com/IndrajeetPatil), [\@IngaPa](https://github.com/IngaPa), [\@jameshowison](https://github.com/jameshowison), [\@jennybc](https://github.com/jennybc), [\@jerryfuyu0104](https://github.com/jerryfuyu0104), [\@jimvine](https://github.com/jimvine), [\@jonmcalder](https://github.com/jonmcalder), [\@jonocarroll](https://github.com/jonocarroll), [\@landroni](https://github.com/landroni), [\@lozalojo](https://github.com/lozalojo), [\@m0nhawk](https://github.com/m0nhawk), [\@mgirlich](https://github.com/mgirlich), [\@mplourde](https://github.com/mplourde), [\@msgoussi](https://github.com/msgoussi), [\@nokome](https://github.com/nokome), [\@novakowd](https://github.com/novakowd), [\@nthun](https://github.com/nthun), [\@pgensler](https://github.com/pgensler), [\@richierocks](https://github.com/richierocks), [\@shapiromatron](https://github.com/shapiromatron), [\@Shians](https://github.com/Shians), [\@sibojan](https://github.com/sibojan), [\@sieste](https://github.com/sieste), [\@simonthelwall](https://github.com/simonthelwall), [\@skanskan](https://github.com/skanskan), [\@stephlocke](https://github.com/stephlocke), [\@swvanderlaan](https://github.com/swvanderlaan), [\@thorvall](https://github.com/thorvall), [\@tjmahr](https://github.com/tjmahr), [\@tomwwagstaff](https://github.com/tomwwagstaff), [\@yeedle](https://github.com/yeedle), [\@yutannihilation](https://github.com/yutannihilation), [\@zapashcanon](https://github.com/zapashcanon), and [\@zhaoy](https://github.com/zhaoy).
