---
title: rlang 0.2.0
slug: rlang-0.2.0
description: >
  rlang 0.2.0 is on CRAN! This version brings many improvements to
  tidy evaluation.
author: Lionel Henry
date: 2018-03-02
categories: [package]
photo:
  url: https://unsplash.com/photos/K5gBXHSvtkc
  author: Denise Johnson
---

```{r setup, include = FALSE}
library("rlang")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r crayon, include = FALSE}
colourise_chunk <- function(type) {
  function(x, options) {
    x <- pkgdown:::escape_html(x)
    x <- ansistrings::ansi_to_html(x, fullpage = FALSE)
    class <- tolower(options$engine)
    sprintf('<pre class="knitr %s">%s</pre>\n', class, x)
  }
}

knitr::knit_hooks$set(
  output = colourise_chunk("output"),
  message = colourise_chunk("message"),
  warning = colourise_chunk("warning"),
  error = colourise_chunk("error")
)

options(crayon.enabled = FALSE)
```

We are happy to announce a new version of [rlang](http://rlang.tidyverse.org/)
that features many improvements to tidy evaluation:

* The quirks in the quasiquotation syntax have been much reduced.

* Quosures gained a much improved printing method with colour support.

* The performance of quoting, splicing and quosure evaluation was
  vastly improved.

* Many bugs have been fixed.


Install the latest version of rlang with:

```{r, eval = FALSE}
install.packages("rlang")
```


### Syntax adjustments for the !! operator

The headline improvement is a much improved syntax for the `!!`
operator (pronounced bang-bang). In a way bang-bang is a lie, or
a [polite fiction](https://adv-r.hadley.nz/quasiquotation.html#the-polite-fiction-of).
It is not a real R operator and only works in tidy evaluation
functions. To implement `!!`, we simply look at the code and
reinterpret two consecutive `!` negations as the unquoting
operator. The major downside of this approach was that our `!!`
operator inherited from the operation precedence of `!`, which is very
low. As a result `!!a + b` was interpreted by R as `!!(a + b)` rather
than `(!!a) + b`. To work around this, tidy eval users had to add
explicit parentheses, which was often confusing.

Happily this is no longer necessary. We now reprocess the whole syntax
tree to give `!!` a more natural operation precedence, i.e. that of
unary `+` and `-`. This means that expressions like `!!a > b` now do
the expected thing (only `a` is unquoted):

```{r}
a <- sym("foo")
expr(!!a > b)
```

In addition, we automatically strip one layer when `!!` is wrapped in
parentheses. This is useful for unquoting function names cleanly:

```{r}
expr((!!a)(bar, baz))

op <- sym(">")
expr((!!op)(bar, baz))
```


### A new print method for quosures

#### Problems with base R deparsing

When R code is printed at the console (be it a function, an
expression, a formula or a quosure), R runs the function `deparse()`
to transform the code to a printable string. The deparsing mechanism
in base R is not optimal for tidy eval:

  - It adds redundant parentheses between the `!` of `!!`:

    ```{r}
    quote(!!x)
    ```

    If you ever encounter these weird `!(!x)` statements (for instance
    by printing the code of a tidy eval function in the console), it's
    because the R printer adds the pair of parentheses at
    printing-time. The good news is that it is merely a printing
    problem.

  - The way it prints inlined vectors is ambiguous. R expressions can
    contain actual data structures not just symbolic code. It is
    especially easy to inline vectors and other objects with
    quasiquotation:

    ```{r}
    expr(mean(!!c(1, 2, 3)))
    ```

    Note how the output above is indistinguishable from the following
    output:

    ```{r}
    expr(mean(c(1, 2, 3)))
    ```

    In both cases, the R printer outputs the code that creates the
    vector, even if the vector is already created as in the first
    case!

  - Quosures are currently implemented as formulas but that's an
    implementation detail. Quosures and formulas behave differently
    and it is misleading to have them print as formulas:

    ```{r}
    expr(mean(!!quo(c(1, 2, 3))))
    ```

    Furthermore, it would be nice to have a way to distinguish
    quosures coming from different contexts.


#### Deparsing with rlang

rlang now features `expr_print()` powered by `expr_deparse()`. These
functions are used in the print method of quosures and nicely solve
the printing issues:

  - Inlined vectors are represented with angular brackets and are
    abbreviated to avoid taking too much space:

    ```{r}
    quo(list(code = 1:10, data = !!(1:10)))
    ```

    S3 objects are printed based on the extensible
    `pillar::type_sum()` generic:

    ```{r}
    quo(lm(!!(disp ~ cyl), data = !!mtcars))
    ```

  - As you can see in the output above quosures are now displayed with
    a leading `^` instead of a tilde in order to distinguish them from
    formulas. In addition, they are now colourised according to their
    environments. Quosures from the global environment are always
    printed with the default colour. Quosures from local environments
    are printed with a unique colour. To illustrate this, let's create
    two quosures referring to `arg`, where `arg` represents a different
    object:

    ```{r, include = FALSE}
    options(crayon.enabled = TRUE)
    ```

    ```{r}
    make_quo <- function(arg) {
      quo(arg)
    }

    q1 <- make_quo("foo")
    q2 <- make_quo("bar")
    q3 <- quo(list(!!q1, !!q2))

    q3
    ```

    ```{r, include = FALSE}
    # Work around ansistrings issue
    # See https://github.com/r-lib/ansistrings/issues/20
    options(crayon.enabled = FALSE)
    ```

    `q3` contains two nested quosures that both point to objects
    called `arg`. While they have the same names these objects come
    from different contexts, the contexts that were created each time
    we called the function `make_quo()`. Let's evaluate `q3` to verify
    this:

    ```{r}
    eval_tidy(q3)
    ```

    The colours should help you figure out what is happening when
    quosures from disparate contexts are unquoted in a larger
    expression.

Note that if you are ever confused by how a quosure prints out
(perhaps because of those intervening `^` symbols), you can use the
new function `quo_squash()` to get rid of all the nested quosures and
print a naked version:

```{r}
quo_squash(q3)
```

Finally, raw expressions created by `expr()` are still printed with
the base R deparser. If you would like to debug tidy eval code and
inspect the result of quasiquotation, use the new `qq_show()` function
to display a raw expression with the new printer:

```{r}
my_wrapper <- function(data, expr) {
  expr <- enquo(expr)
  qq_show(data %>% mutate(mean(!!expr)))
}

my_wrapper(mtcars, cyl)
```
