---
title: Resource Cleanup in C and the R API
date: '2019-05-22'
slug: resource-cleanup-in-c-and-the-r-api
author: Gábor Csárdi, Lionel Henry
categories: [package, programming]
description: >
    How to avoid resource leaks in C code in R packages?
    Introducing the cleancall package.
photo:
  url: https://pixabay.com/photos/plumbing-pipe-wrench-plumber-840835
  author: Steve Buissinne
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7,
  fig.align = 'center',
  fig.asp = 0.618, # 1 / phi
  out.width = "700px"
)
```

<style>
h1 code, h2 code {
    font-size: 0.8em;
}
</style>

# Introduction

We have just released the [cleancall] package to CRAN. [cleancall] is a
new package that helps avoiding resource leaks in the C code of your
R package. In this blog post, we show some example resource leaks from
base R and R packages, and also show three different ways to fix them,
one of which uses the [cleancall] package.

# The problem

When writing C code in R packages, one non-trivial issue is to make
sure that resources acquired by a C function are correctly released
or wrapped before giving back control to R. The difficulties partially
arise from the many ways a C function can terminate and transfer control
back to R:

* regular termination,
* error,
* warning or other condition caught with `tryCatch()`,
* debugger exit,
* keyboard interruption.

Resources that need to be released include:

* memory areas,
* file handles,
* connections (sockets),
* database connections,
* Windows handles for threads, events, etc.,
* others.

## Regular termination

Most functions terminate successfully, most of the time. Releasing
resources is easy in this case, we can simply do it at the end of the
function, before returning.

## Error

Errors may happen two ways. They can be thrown directly, or by the R API.
A function that reads a file will throw an error it if
fails to open the file for reading. These errors are easy to handle, we
can release all resources before throwing the error.

Errors can also be thrown by the R API. The R C API functions do not
return on error, instead they perform an error (a `longjmp` technically),
that can be caught by R or (recently) C code. Releasing resources is
trickier in this case, as our C function never gets back the control after
the failed R API call.

## Warning or other condition caught with `tryCatch()`

Usually we assume that throwing a warning (or other non-error condition)
does not cause an early exit, but this is not always the case. Warnings can
be caught by an exiting `tryCatch()`.

## Debugger exit

If the C function calls back to R to evaluate R expressions, then
these might invoke the debugger, see `?debug` or `?trace`. The `Q` debugger
command invokes the `abort` restart, to get back to the top-level prompt.
In this case most functions on the (R and C) stack are removed at once.

## Keyboard interruption

C code can call `R_CheckUserInterrupt()`, and it is indeed suggested to do
so during long computation or I/O. If the user initiates an interrupt via
the keyboard or a signal, then `R_CheckUserInterrupt()` never returns, and
the control goes back to top level.

To illustrate the resource cleanup problem more, we'll show some examples
from base R and CRAN packages that contain potential resource leaks.

# Example 1: `download.file(method = "internal")`

The C implementation of `download.file()` opens a file for saving the
downloaded data to, and it does not clean or even close this file on a
keyboard interrupt. The file is opened here:
https://github.com/wch/r-source/blob/f3a67c36c5cf4c715dac374e33168cfc348106aa/src/modules/internet/internet.c#L553

```c
	FILE *out;
    [...]
	out = R_fopen(R_ExpandFileName(file), mode);
    [...]
```

We can easily show the problem in practice as well.
First, we create a handy function that interrupts the current process
one second after calling it. This allows us to simulate an interrupt from
the keyboard. It starts a background process that will send an interrupt
(SIGINT on Unix, CTRL+C on Windows) to our R process.

```{r}
interrupt <- function(expr, after = 1) {
  tryCatch({
    p <- callr::r_bg(function(pid, after) {
      Sys.sleep(after)
      ps::ps_interrupt(ps::ps_handle(pid))
    }, list(pid = Sys.getpid(), after = after))
    expr
    p$kill()
  }, interrupt = function(e) e)
}
```

`ps::ps_open_files()` lists all open files of a process,`ps::ps_handle()`
returns a handle for the current R process. You should see the three
file descriptors corresponding to standard input, output and error here,
and another file is used by the blogdown engine rendering this page. If you
run this code in your R session you'll not see that file, but of course you
may see others.

```{r}
ps::ps_open_files(ps::ps_handle())
```

Now we will start downloading a file, and while it is downloading, we'll
send an interrupt to our R process, causing an early exit.

```{r}
interrupt(
  download.file("http://httpbin.org/delay/3", tempfile(), method = "internal")
)
```

The list of open files will now include the one opened by `download.file()`:

```{r}
ps::ps_open_files(ps::ps_handle())
```

Because `download.file()` fails to clean up this file handle, it is not
released until the R session quits. On Windows this file is probably locked,
and cannot be removed until R exits.

# Example 2: `download.file(method = "libcurl")`

There is a very similar bug in the implementation of the `download.file()`
libcurl method, which opens an output file here:
https://github.com/wch/r-source/blob/def075db88ae87104c38437a57e9327b078bb804/src/modules/internet/libcurl.c#L566
but it does not close it on error or interrupt. This method can download
many files in parallel, in which case multiple file handles may be lost:

```{r}
ps::ps_open_files(ps::ps_handle())
interrupt(
  download.file(
    rep("https://httpbin.org/delay/3", 3), paste0(tempfile(), 1:3),
    method = "libcurl")
)
ps::ps_open_files(ps::ps_handle())
```

# Example 3: `edit()`

`edit()` invokes the text editor specified by the `editor` option, with the
supplied object to be edited. `edit()` can also write the edited version
to a file. It opens a file here to create its first version, before starting
the editor:
https://github.com/wch/r-source/blob/019f16d3dec4f97c1b4e4f0ec905148e36979e37/src/main/edit.c#L121-L128

```c
121    if((fp=R_fopen(R_ExpandFileName(filename), "w")) == NULL)
122        errorcall(call, _("unable to open file"));
123    if (LENGTH(STRING_ELT(fn, 0)) == 0) EdFileUsed++;
124    PROTECT(src = deparse1(x, 0, FORSOURCING)); /* deparse for sourcing, not for display */
125    for (i = 0; i < LENGTH(src); i++)
126        fprintf(fp, "%s\n", translateChar(STRING_ELT(src, i)));
127    UNPROTECT(1); /* src */
128    fclose(fp);
```

If all goes well, then it will close it just 7 code lines later.
However, it also calls the R API a number of times before closing the file,
so if any of these API calls fail, the file is never closed. In particular,
it calls `deparse1()` which is roughly equivalent to the R `deparse()`
function, and `deparse`()` fails on long vectors:

```{r, error = TRUE}
ps::ps_open_files(ps::ps_handle())
edit(1:10e10, file = tempfile())
ps::ps_open_files(ps::ps_handle())
```

Notice that we have one more extra open file here.

# Example 4: the `wait()` method in [processx]:

`processx::process` is an R6 class for a subprocess. Its `$wait()` method
waits for the subprocess to finish, with a timeout. To implement `$wait()`
on Unix, [processx] opens a pair of pipe file descriptors. These are
temporary and should be closed once the function exits. `$wait()` is
interruptible, it calls `R_CheckUserInterrupt()` periodically. However,
in the current, 3.3.0 version of [processx] it does not close the pipe
file descriptors on an interrupt. Here is an illustration:

```{r}
ps::ps_num_fds(ps::ps_handle())

p <- processx::process$new("sleep", "10")
interrupt(
  p$wait()
)
p$kill()
gc()

ps::ps_num_fds(ps::ps_handle())
```

`ps_num_fds()` prints the number of open file descriptors of a process.
The two extra file descriptors that are open after garbage collection are
the two ends of the (supposedly) temporary pipe, just opened by `$wait()`.

# Fix 1: External pointer and finalizer

One generic solution to resource cleanup is to wrap all C resources into
one or more R external pointer objects, and add finalizers to them. We
show how this can fix the [processx] problem in the previous example.

This is how the `processx_wait()` C function looks like before the fix:
https://github.com/r-lib/processx/blob/a8f09d147fead78347a87fcf4e0fbd1c07de1c21/src/unix/processx.c#L507-L589

First, we need to create a finalizer function, that will be called by
the R garbage collector, after `$wait()` has finished, at the next
garbage collection:

```c
static void processx__wait_finalizer(SEXP xptr) {
  SEXP tag = R_ExternalPtrTag(xptr);
  if (INTEGER(tag)[0] >= 0) close(INTEGER(tag)[0]);
  if (INTEGER(tag)[1] >= 0) close(INTEGER(tag)[1]);
}
```

An external pointer can have a _tag_, which is an R object that is kept
alive as long the pointer object itself is alive. In this case we can put
the file descriptors in the tag, in an integer vector of length two.
In more complicated cases the resources cannot easily be represented as
R objects, so you would use the actual C pointer, with a custom C
`struct` to store them.

Now we need to create the external pointer, before the pipes are opened:

```c
  SEXP tag = PROTECT(allocVector(INTSXP, 2));
  INTEGER(tag)[0] = INTEGER(tag)[1] = -1;
  SEXP xptr = PROTECT(R_MakeExternalPtr(NULL, tag, R_NilValue));
  R_RegisterCFinalizerEx(xptr, processx__wait_finalizer, 0);
```

We initialize the file descriptors to -1, which is guard value, meaning
that no cleanup is needed.

We can now open the pipes and save their file descriptors in the tag of
the external pointer.

```c
  if (pipe(handle->waitpipe)) {
    processx__unblock_sigchld();
    error("processx error: %s", strerror(errno));
  }
  INTEGER(tag)[0] = handle->waitpipe[0];
  INTEGER(tag)[1] = handle->waitpipe[1];
```

At the end of the function, we need to unprotect the external pointer,
and the tag:

```c
  UNPROTECT(2);
  return ScalarLogical(ret != 0);
}
```

One potential problem with the external pointer fix is that the resources
will only be cleaned up at the next garbage collection, and sometimes this
is too late. For example, if an open file is locked by the operating system,
then we won't be able to remove that file, or its directory, until the
garbage collector runs and closes it. It is easier to program if
resource cleanup is immediate, and luckily the other two fixes below are.

# Fix 2: the `R_ExecWithCleanup()` function

`R_ExecWithCleanup()` is a function in the R API, that can add a cleanup
function to a regular C function call. The cleanup function is always
executed, even on early exit:

```c
SEXP R_ExecWithCleanup(SEXP (*fun)(void *), void *data,
               void (*cleanfun)(void *), void *cleandata);
```

`R_ExecWithCleanup()` calls `fun` with `data`, and then calls
`cleanfun` with `cleandata`. If `fun` calls the R API and exits early,
then it will still call `cleanfun` with `cleandata`, before throwing
the error up the stack.

To fix `$wait()` with `R_ExecWithCleanup()`, we'll need to define a
cleanup function first. The cleanup function receives the cleanup data
as a `void` pointer, we'll define a `struct` for this. For simplicity,
this struct will also include the arguments to `processx_wait()`, so
we can use the same struct for both `fun` and `cleanupfun`.

```c
struct processx_wait_data {
  SEXP status;
  SEXP timeout;
  int fds[2];
};

void processx_wait_cleanup(void *data) {
  struct processx_wait_data *pdata = data;
  if (pdata->fds[0] >= 0) close(pdata->fds[0]);
  if (pdata->fds[1] >= 0) close(pdata->fds[1]);
}
```

The new `processx_wait()` function will call `R_ExecWithCleanup()`,
to call the original `processx_wait()`, under a new name:

```c
SEXP processx_wait_internal(void *data);
SEXP processx_wait(SEXP status, SEXP timeout) {
  struct processx_wait_data pdata = { status, timeout, { -1, -1 } };
  SEXP result = R_ExecWithCleanup(processx_wait_internal, &pdata,
                                  processx_wait_cleanup, &pdata);
  return result;
}
```

`processx_wait_internal()` is very much like `processx_wait()` used to
be, but we need to extract the arguments from the struct at the beginning:

```c
SEXP processx_wait_internal(void *data) {
  struct processx_wait_data *pdata = data;
  SEXP status = pdata->status;
  SEXP timeout = pdata->timeout;
  int *fds = pdata->fds;
  ...
```

Now the only other thing we need to do is saving the file descriptors
in the data struct:

```
  if (pipe(handle->waitpipe)) {
    processx__unblock_sigchld();
    error("processx error: %s", strerror(errno));
  }
  fds[0] = handle->waitpipe[0];
  fds[1] = handle->waitpipe[1];
```

# Fix 3: the [cleancall] package

`R_ExecWithCleanup()` is a good fix to the resource cleanup problem,
but it can be verbose and error prone. It also requires that you
replace your original function with a wrapper that packs the original
function arguments into a `struct` and an internal function that
only has a `void*` argument.

We created the [cleancall] package to make resource cleanup easier.
This package automates wrapping your functions with `R_ExecWithCleanup()`.
[cleancall] has been just published on CRAN. Here we show how to use it to
fix the [processx] `$wait()` method.

To use it in your package, you need to specify [cleancall] as a dependency,
both as `LinkingTo` and `Imports`:
```
...
Imports: cleancall
LinkingTo: cleancall
...
```

[cleancall] defines the `call_with_cleanup()` R function and the
`r_call_on_exit()` and `r_call_on_early_exit()` C functions.

You need to replace `.Call()` with `call_with_cleanup()` in your R code:

```
cleancall::call_with_cleanup(c_processx_wait, private$status,
                               as.integer(timeout))
```

In your C code, include the `cleancall.h` header:
```c
#include <cleancall.h>
```

Next, create a cleanup function for the resource type:
```c
void processx__close_fd(void *ptr) {
  int *fd = ptr;
  if (*fd >= 0) close(*fd);
}
```

This cleanup function closes a file descriptor. Once your resource type
has a cleanup function, you can call `r_call_on_exit()` or
`r_call_on_early_exit()` every time you acquire a resource of that type.
Use `r_call_on_exit()` if the resource must be released on normal
termination as well, and `r_call_on_early_exit()` if it must be released
on early termination only:

```c
  if (pipe(handle->waitpipe)) {
    processx__unblock_sigchld();
    error("processx error: %s", strerror(errno));
  }
  r_call_on_exit(processx__close_fd, handle->waitpipe);
  r_call_on_exit(processx__close_fd, handle->waitpipe + 1);
```

Typically `r_call_on_exit()` is more convenient for temporary resources.
`r_call_on_early_exit()` is more convenient if the C function returns a
handle (e.g. external pointer), for which it needs to allocate resources
gradually. If all resource allocations are successful, and the function
returns normally, then no cleanup is needed. However if an intermediate
step fails, you need to release the resources acquired before the failure.

In does not matter much in this simple example, but it in important in
general that exit handlers are always executed in reverse order (last one in
is the first one out, LIFO). This makes it easy to build a resource
gradually. Exit handlers installed via `r_call_on_exit()` and
`r_call_on_early_exit()` share the same stack.

We suggest that exit handlers are kept as simple and fast as possible.
In particular, errors (and other early exits) triggered from exit handlers
are not caught currently. If an exit handler exits early the others do not
run. If this is an issue, you can wrap the exit handler in `R_tryCatch()`
(available for R 3.4.0 and later). But in general the best error handlers
do not call the R API at all.

# What about C++?

The resource cleanup problem is also present if you interface C++ code.
While [cleancall] can be used with C++, it works best with C code,
since it follows C idioms.

Using external pointers and finalizers works for C++ as well, as does
wrapping R API calls in `R_ExecWithCleanup()`.

Alternatively, you can wrap your R API calls in `R_tryCatch()` (R 3.4.0
and later), or use `R_UnwindProtect()` (R 3.5.0 and later).

If your C++ code needs to support older R versions, that is more
challenging, and one possibility is to call back to R and call `tryCatch()`
there.

# Summary

Resource cleanup in C code can be challenging, especially given that there
is not very much documentation on this topic.

Hopefully this post and the [cleancall] package will make resource
cleanup much simpler, and fewer R packages will suffer from resource
leaks in the future.

[cleancall]: https://github.com/r-lib/cleancall#readme
[processx]: https://github.com/r-lib/processx#readme
