---
title: pkgdown 1.0.0
author: Hadley Wickham
date: '2018-05-03'
slug: pkgdown-1-0-0
categories: [package]
photo:
  url: https://unsplash.com/photos/rx1iJ59jRyU
  author: Annie Spratt
---

I'm very pleased to announce the initial release of [pkgdown](http://pkgdown.r-lib.org) to CRAN! pkgdown is designed to make it quick and easy to build a website for your package. You can see pkgdown in action at <http://pkgdown.r-lib.org>: this is the output of pkgdown applied to the latest version of pkgdown. This initial release is version 1.0.0 in recognition that pkgdown has been in development for 6 years (!) and has already been used to make websites for [over 1600 packages](https://github.com/search?q=pkgdown.css+in%3Apath+path%3Adocs&type=Code) (!!). 

Install pkgdown with:

```{r, eval = FALSE}
install.packages("pkgdown")
```

## Usage

Basic usage is simple: run `build_site()` within your package directory:

```{r, eval = FALSE}
pkgdown::build_site()
```

By default, this will generate a `docs/` directory. The home page will be generated from your `README.md`, a function reference will be generated from the documentation in the `man/` directory, and any `vignettes/` will be converted into articles

If you are using GitHub, the easiest way to make this your package website is to check it in, push it, then go to repo settings and ensure that the __GitHub pages__ source is set to "master branch /docs folder".

## Customisation

To customise your site, first read [`vignette("pkgdown")`](http://pkgdown.r-lib.org/articles/pkgdown.html), then read the documentation for the functions responsible for building each part of the site:

* [`init_site()`](http://pkgdown.r-lib.org/reference/init_site.html)
* [`build_home()`](http://pkgdown.r-lib.org/reference/build_home.html)
* [`build_reference()`](http://pkgdown.r-lib.org/reference/build_reference.html)
* [`build_articles()`](http://pkgdown.r-lib.org/reference/build_articles.html)
* [`build_tutorials()`](http://pkgdown.r-lib.org/reference/build_tutorials.html)
* [`build_news()`](http://pkgdown.r-lib.org/reference/build_news.html)

## In the wild

A great way to see what you can do with pkgdown is to look at existing websites. Here are a few examples created by some of the contributors to pkgdown:

* [bayesplot](http://mc-stan.org/bayesplot/index.html)
  [[source](https://github.com/stan-dev/bayesplot/tree/gh-pages)]:
  plotting functions for posterior analysis, model checking, and MCMC 
  diagnostics.

* [valr](https://rnabioco.github.io/valr/) 
  [[source](https://github.com/rnabioco/valr)]: 
  read and manipulate genome intervals and signals.

* [mkin](http://jranke.github.io/mkin/)
  [[source](https://github.com/jranke/mkin)]:
  calculation routines based on the FOCUS Kinetics Report

* [NMF](http://renozao.github.io/NMF/master/index.html)
  [[source](https://github.com/renozao/NMF)]:
  a framework to perform non-negative matrix factorization (NMF).

## Acknowledgements

pkgdown has been a long time coming, and it wouldn't have been possible without a devoted community of users, many of whom have gone on to contribute fixes. I'd like to particularly thank [Jay Hesselberth](https://hesselberthlab.org) who put a lot of work into implementing site search in partnership with [Algolia's docsearch](https://community.algolia.com/docsearch/). 

A big thanks goes out to all 184 other people who contributed issues, pull requests, and comments: [\@aaronrudkin](https://github.com/aaronrudkin), [\@adamkski](https://github.com/adamkski), [\@adibender](https://github.com/adibender), [\@AdrianHordyk](https://github.com/AdrianHordyk), [\@adrtod](https://github.com/adrtod), [\@agrueneberg](https://github.com/agrueneberg), [\@alekrutkowski](https://github.com/alekrutkowski), [\@amatsuo](https://github.com/amatsuo), [\@amrrs](https://github.com/amrrs), [\@andrie](https://github.com/andrie), [\@antuki](https://github.com/antuki), [\@aravind-j](https://github.com/aravind-j), [\@batpigandme](https://github.com/batpigandme), [\@bborgesr](https://github.com/bborgesr), [\@bhaskarvk](https://github.com/bhaskarvk), [\@bowerth](https://github.com/bowerth), [\@BrianDiggs](https://github.com/BrianDiggs), [\@carlosparadis](https://github.com/carlosparadis), [\@cboettig](https://github.com/cboettig), [\@cderv](https://github.com/cderv), [\@chfleming](https://github.com/chfleming), [\@choisy](https://github.com/choisy), [\@christophergandrud](https://github.com/christophergandrud), [\@ck37](https://github.com/ck37), [\@corybrunson](https://github.com/corybrunson), [\@crew102](https://github.com/crew102), [\@czeildi](https://github.com/czeildi), [\@Dasonk](https://github.com/Dasonk), [\@davebraze](https://github.com/davebraze), [\@davidcarslaw](https://github.com/davidcarslaw), [\@DavisVaughan](https://github.com/DavisVaughan), [\@dfalbel](https://github.com/dfalbel), [\@dincerti](https://github.com/dincerti), [\@dirkschumacher](https://github.com/dirkschumacher), [\@dmenne](https://github.com/dmenne), [\@dracodoc](https://github.com/dracodoc), [\@edgararuiz](https://github.com/edgararuiz), [\@edzer](https://github.com/edzer), [\@eibanez](https://github.com/eibanez), [\@eliotmcintire](https://github.com/eliotmcintire), [\@Emaasit](https://github.com/Emaasit), [\@emhart](https://github.com/emhart), [\@Enchufa2](https://github.com/Enchufa2), [\@espinielli](https://github.com/espinielli), [\@famuvie](https://github.com/famuvie), [\@flaviobarros](https://github.com/flaviobarros), [\@flying-sheep](https://github.com/flying-sheep), [\@friendly](https://github.com/friendly), [\@gaborcsardi](https://github.com/gaborcsardi), [\@GegznaV](https://github.com/GegznaV), [\@GeoBosh](https://github.com/GeoBosh), [\@ghost](https://github.com/ghost), [\@goldingn](https://github.com/goldingn), [\@gr8694](https://github.com/gr8694), [\@GShotwell](https://github.com/GShotwell), [\@Guilz](https://github.com/Guilz), [\@gvegayon](https://github.com/gvegayon), [\@hadley](https://github.com/hadley), [\@iagomosqueira](https://github.com/iagomosqueira), [\@ijlyttle](https://github.com/ijlyttle), [\@IndrajeetPatil](https://github.com/IndrajeetPatil), [\@isteves](https://github.com/isteves), [\@jacob-long](https://github.com/jacob-long), [\@jakob-r](https://github.com/jakob-r), [\@jakobbossek](https://github.com/jakobbossek), [\@jamesdunham](https://github.com/jamesdunham), [\@janeshdev](https://github.com/janeshdev), [\@jaredhuling](https://github.com/jaredhuling), [\@javierluraschi](https://github.com/javierluraschi), [\@jbkunst](https://github.com/jbkunst), [\@jefferis](https://github.com/jefferis), [\@jeffwong](https://github.com/jeffwong), [\@jemus42](https://github.com/jemus42), [\@jennybc](https://github.com/jennybc), [\@jentjr](https://github.com/jentjr), [\@jeroenjanssens](https://github.com/jeroenjanssens), [\@jflournoy](https://github.com/jflournoy), [\@jhoobergs](https://github.com/jhoobergs), [\@jimhester](https://github.com/jimhester), [\@jjallaire](https://github.com/jjallaire), [\@jkennel](https://github.com/jkennel), [\@JohnCoene](https://github.com/JohnCoene), [\@JohnMount](https://github.com/JohnMount), [\@jolars](https://github.com/jolars), [\@jonasfoe](https://github.com/jonasfoe), [\@jranke](https://github.com/jranke), [\@jrosen48](https://github.com/jrosen48), [\@jthurner](https://github.com/jthurner), [\@KasperSkytte](https://github.com/KasperSkytte), [\@katrinleinweber](https://github.com/katrinleinweber), [\@kbenoit](https://github.com/kbenoit), [\@KellyLoucks](https://github.com/KellyLoucks), [\@kent37](https://github.com/kent37), [\@keurcien](https://github.com/keurcien), [\@kevinushey](https://github.com/kevinushey), [\@kimyen](https://github.com/kimyen), [\@kmiddleton](https://github.com/kmiddleton), [\@kohske](https://github.com/kohske), [\@krlmlr](https://github.com/krlmlr), [\@kuhnrl30](https://github.com/kuhnrl30), [\@kwstat](https://github.com/kwstat), [\@kylebmetrum](https://github.com/kylebmetrum), [\@lbusett](https://github.com/lbusett), [\@lgatto](https://github.com/lgatto), [\@LiNk-NY](https://github.com/LiNk-NY), [\@lionel-](https://github.com/lionel-), [\@luca-scr](https://github.com/luca-scr), [\@maelle](https://github.com/maelle), [\@MarcinKosinski](https://github.com/MarcinKosinski), [\@marcosci](https://github.com/marcosci), [\@MarkEdmondson1234](https://github.com/MarkEdmondson1234), [\@markusdumke](https://github.com/markusdumke), [\@maurolepore](https://github.com/maurolepore), [\@maxheld83](https://github.com/maxheld83), [\@mdsumner](https://github.com/mdsumner), [\@merliseclyde](https://github.com/merliseclyde), [\@michaellevy](https://github.com/michaellevy), [\@michelk](https://github.com/michelk), [\@mjsteinbaugh](https://github.com/mjsteinbaugh), [\@mooresm](https://github.com/mooresm), [\@Mullefa](https://github.com/Mullefa), [\@mwillumz](https://github.com/mwillumz), [\@nacnudus](https://github.com/nacnudus), [\@nbenn](https://github.com/nbenn), [\@nealrichardson](https://github.com/nealrichardson), [\@newtux](https://github.com/newtux), [\@nhejazi](https://github.com/nhejazi), [\@nite](https://github.com/nite), [\@noamross](https://github.com/noamross), [\@nsheff](https://github.com/nsheff), [\@nuno-agostinho](https://github.com/nuno-agostinho), [\@pachevalier](https://github.com/pachevalier), [\@PascalKieslich](https://github.com/PascalKieslich), [\@pat-s](https://github.com/pat-s), [\@patperry](https://github.com/patperry), [\@privefl](https://github.com/privefl), [\@rajanand](https://github.com/rajanand), [\@randy3k](https://github.com/randy3k), [\@ras44](https://github.com/ras44), [\@renozao](https://github.com/renozao), [\@restonslacker](https://github.com/restonslacker), [\@rickhelmus](https://github.com/rickhelmus), [\@RLesur](https://github.com/RLesur), [\@RLumSK](https://github.com/RLumSK), [\@RMHogervorst](https://github.com/RMHogervorst), [\@robertzk](https://github.com/robertzk), [\@Robinlovelace](https://github.com/Robinlovelace), [\@romainfrancois](https://github.com/romainfrancois), [\@rstub](https://github.com/rstub), [\@sahilseth](https://github.com/sahilseth), [\@samuel-rosa](https://github.com/samuel-rosa), [\@saurfang](https://github.com/saurfang), [\@sckott](https://github.com/sckott), [\@sebastian-c](https://github.com/sebastian-c), [\@sfirke](https://github.com/sfirke), [\@slwu89](https://github.com/slwu89), [\@statsandthings](https://github.com/statsandthings), [\@stephlocke](https://github.com/stephlocke), [\@StevenMMortimer](https://github.com/StevenMMortimer), [\@surmann](https://github.com/surmann), [\@t-kalinowski](https://github.com/t-kalinowski), [\@terrytangyuan](https://github.com/terrytangyuan), [\@ThierryO](https://github.com/ThierryO), [\@thk686](https://github.com/thk686), [\@tjmahr](https://github.com/tjmahr), [\@tomaskrehlik](https://github.com/tomaskrehlik), [\@topepo](https://github.com/topepo), [\@trinker](https://github.com/trinker), [\@Tutuchan](https://github.com/Tutuchan), [\@UweBlock](https://github.com/UweBlock), [\@WastlM](https://github.com/WastlM), [\@wch](https://github.com/wch), [\@wjakethompson](https://github.com/wjakethompson), [\@wlandau](https://github.com/wlandau), [\@wlandau-lilly](https://github.com/wlandau-lilly), [\@wrathematics](https://github.com/wrathematics), [\@wviechtb](https://github.com/wviechtb), [\@xguardi](https://github.com/xguardi), [\@yiluheihei](https://github.com/yiluheihei), [\@yutannihilation](https://github.com/yutannihilation), [\@zappingseb](https://github.com/zappingseb), [\@zbjornson](https://github.com/zbjornson), [\@zkamvar](https://github.com/zkamvar), and [\@znmeb](https://github.com/znmeb)
