---
output: hugodown::hugo_document

slug: usethis-2-0-0
title: usethis 2.0.0
date: 2020-12-10
author: Jennifer Bryan
description: >
    This is a big release aimed at improving usability, especially around Git
    and GitHub functionality.

photo:
  url: https://unsplash.com/photos/knzXwBCtEeM
  author: Kiana Bosman
categories: [package] 
tags: [devtools,r-lib,usethis]
---

We're ecstatic to announce the release of usethis v2.0.0 ([usethis.r-lib.org](https://usethis.r-lib.org/)).
usethis is a package that facilitates interactive workflows for R project creation and development.
It's mostly focussed on easing day-to-day package development, but many of its functions are also useful for non-package projects.

You can install usethis from CRAN with:

```{r, eval = FALSE}
install.packages("usethis")
```

This is a major release, involving lots of change under the hood, and yet there are very few truly new features.
Instead, this release is mostly about making more things "just work" the way it seems like they should.

This blog post will only hit a few highlights, mostly relating to Git and GitHub functionality:

* We've switched to a new package, gert, for Git operations. This resolves some
  long-standing difficulties around credential-finding and provides more
  reliable support for both HTTPS and SSH remotes.
* GitHub personal access tokens are now handled in the same way as command line
  Git. This means we can all manage our tokens more securely and, for example,
  can have different tokens for different hosts.
* Anything that works for github.com should now work on any GitHub Enterprise
  deployment.
* A default branch named `main` should be fully supported.
* We've added more functions and features for working with GitHub pull requests. 

You can see a full list of changes in the [release notes](https://usethis.r-lib.org/news/index.html#usethis-2-0-0-2020-12-10).

```{r setup}
library(usethis)
```

## Topic 1

## Topic 2

## Acknowledgements

A big thanks to everyone who helped with this release by reporting bugs, discussing issues, and contributing code:
[&#x0040;albersonmiranda](https://github.com/albersonmiranda), [&#x0040;andrader](https://github.com/andrader), [&#x0040;arashHaratian](https://github.com/arashHaratian), [&#x0040;Athanasiamo](https://github.com/Athanasiamo), [&#x0040;batpigandme](https://github.com/batpigandme), [&#x0040;cderv](https://github.com/cderv), [&#x0040;cstepper](https://github.com/cstepper), [&#x0040;dmenne](https://github.com/dmenne), [&#x0040;friep](https://github.com/friep), [&#x0040;hadley](https://github.com/hadley), [&#x0040;ijlyttle](https://github.com/ijlyttle), [&#x0040;jamesmyatt](https://github.com/jamesmyatt), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;jeroen](https://github.com/jeroen), [&#x0040;jimhester](https://github.com/jimhester), [&#x0040;jtr13](https://github.com/jtr13), [&#x0040;KoderKow](https://github.com/KoderKow), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;lionel-](https://github.com/lionel-), [&#x0040;maelle](https://github.com/maelle), [&#x0040;malcolmbarrett](https://github.com/malcolmbarrett), [&#x0040;maurolepore](https://github.com/maurolepore), [&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [&#x0040;mrcaseb](https://github.com/mrcaseb), [&#x0040;vinhtantran](https://github.com/vinhtantran), and [&#x0040;whtns](https://github.com/whtns).

