---
output: hugodown::hugo_document

slug: usethis-2-0-0
title: usethis 2.0.0
date: 2020-12-10
author: Jennifer Bryan
description: >
    This is a big release aimed at improving usability, especially around Git
    and GitHub functionality.

photo:
  url: https://unsplash.com/photos/knzXwBCtEeM
  author: Kiana Bosman
categories: [package] 
tags: [devtools,r-lib,usethis]
---

We're ecstatic to announce the release of usethis v2.0.0 ([usethis.r-lib.org](https://usethis.r-lib.org/)). usethis is a package that facilitates interactive workflows for R project creation and development. It's mostly focussed on easing day-to-day package development, but many of its functions are also useful for non-package projects. Install usethis with:

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("usethis")
```

This blog post will ...

You can see a full list of changes in the [release notes](https://usethis.r-lib.org/news/index.html#usethis-2-0-0-2020-12-10).

```{r setup}
library(usethis)
```

## Topic 1

## Topic 2

## Acknowledgements

A big thanks to everyone who helped with this release by reporting bugs, discussing issues, and contributing code:
[&#x0040;albersonmiranda](https://github.com/albersonmiranda), [&#x0040;andrader](https://github.com/andrader), [&#x0040;arashHaratian](https://github.com/arashHaratian), [&#x0040;Athanasiamo](https://github.com/Athanasiamo), [&#x0040;batpigandme](https://github.com/batpigandme), [&#x0040;cderv](https://github.com/cderv), [&#x0040;cstepper](https://github.com/cstepper), [&#x0040;dmenne](https://github.com/dmenne), [&#x0040;friep](https://github.com/friep), [&#x0040;hadley](https://github.com/hadley), [&#x0040;ijlyttle](https://github.com/ijlyttle), [&#x0040;jamesmyatt](https://github.com/jamesmyatt), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;jeroen](https://github.com/jeroen), [&#x0040;jimhester](https://github.com/jimhester), [&#x0040;jtr13](https://github.com/jtr13), [&#x0040;KoderKow](https://github.com/KoderKow), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;lionel-](https://github.com/lionel-), [&#x0040;maelle](https://github.com/maelle), [&#x0040;malcolmbarrett](https://github.com/malcolmbarrett), [&#x0040;maurolepore](https://github.com/maurolepore), [&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [&#x0040;mrcaseb](https://github.com/mrcaseb), [&#x0040;vinhtantran](https://github.com/vinhtantran), and [&#x0040;whtns](https://github.com/whtns).

