---
output: hugodown::hugo_document

slug: nanoparquet-0-5-1
title: nanoparquet 0.5.1
date: 2026-04-22
author: Gábor Csárdi
description: >
    nanoparquet 0.5.1 (and 0.5.0) bring lots of bug fixes and some new
    features: reading and writing of list columns, bit64::integer64 and
    blob::blob support and the ability to write a parquet file to the
    standard output.

photo:
  url: https://pixabay.com/photos/chocolate-box-square-brown-white-7892542/
  author: ykaiavu

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package]
tags: [package, parquet]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're very chuffed to announce the release of
[nanoparquet](https://nanoparquet.r-lib.org/) 0.5.1 (and 0.5.0).
nanoparquet is a small, self-sufficient R package for reading and
writing Parquet files.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("nanoparquet")
```

This blog post will go over some of the improvements in nanoparquet 0.5.0
and 0.5.1.

You can see a full list of changes in the release notes
[here](https://github.com/r-lib/nanoparquet/releases/tag/v0.5.0) and
[here](https://github.com/r-lib/nanoparquet/releases/tag/v0.5.1).

```{r setup}
library(nanoparquet)
```

```{r include = FALSE}
library(pillar)
```

## List columns

Parquet has a `LIST` type for columns whose values are variable-length
sequences of scalars. nanoparquet 0.5.0 adds support for reading and
writing such columns.

> **Note:** for now nanoparquet supports one level of nesting: each
> element of a list column must be an atomic vector of a single type,
> not a list of lists. All elements in a column must have the same scalar
> type.

To write a list column, put a regular R list into your data frame.
Each element must be an atomic vector (integer, double, or character),
`NULL` for a missing list, or an empty vector for an empty list.
`NA` values inside an element vector encode missing elements.

```{r list-write}
df <- data.frame(id = 1:4)
df$scores <- list(c(80L, 95L, 70L), c(100L), NULL, integer(0))
write_parquet(df, tmp <- tempfile(fileext = ".parquet"))
```

```{r list-write-schema}
read_parquet_schema(tmp)
```


`read_parquet()` reads `LIST` columns back as R list columns:

```{r list-read}
as.data.frame(read_parquet(tmp))
```

`infer_parquet_schema()` shows how nanoparquet maps each column to a Parquet
type. For list columns, the `r_type` shows e.g. `list(integer)`:

```{r list-schema}
infer_parquet_schema(df)[2:7]
```

A `LIST` column occupies three rows in the schema: the outer list node,
a repeated group node, and the leaf element node.

When you need to specify the element type explicitly, you can use
`parquet_schema()`:

```{r list-explicit-schema}
schema <- parquet_schema(
  id     = "INT32",
  scores = list("LIST", element = "INT32")
)
write_parquet(df, tmp2 <- tempfile(fileext = ".parquet"), schema = schema)
```

```{r list-explicit-schema-schema}
read_parquet_schema(tmp2)
```

## New types

### `bit64::integer64`

Parquet's `INT64` type holds 64-bit integers. R's native `integer` is
only 32 bits, so nanoparquet has mapped `INT64` to `double` by default.
nanoparquet 0.5.1 adds support for `bit64::integer64`, which gives you true
64-bit integer arithmetic in R.

`write_parquet()` now writes `bit64::integer64` columns as `INT64`:

```{r include = FALSE}
suppressPackageStartupMessages(library(bit64))
```

```{r bit64-write}
library(bit64)
df2 <- data.frame(id = as.integer64(c(1e18, 2e18, 3e18)))
write_parquet(df2, tmp3 <- tempfile(fileext = ".parquet"))
read_parquet_schema(tmp3)
```

To read `INT64` columns back as `bit64::integer64` instead of the default
`double`, use the `read_int64_type` option. The bit64 package must be
installed; if it isn't, nanoparquet throws a clear error.

```{r bit64-read}
read_parquet(tmp3, options = parquet_options(read_int64_type = "integer64"))
```

### `blob::blob`

`read_parquet()` previously returned raw `BYTE_ARRAY` and
`FIXED_LEN_BYTE_ARRAY` columns (i.e. those without a string, UUID, or
decimal annotation) as plain lists of raw vectors. They are now returned as
`blob::blob` objects, which print more neatly and come with the full set of
blob helpers. `write_parquet()` now also accepts `blob::blob` columns, so
round-tripping binary data is straightforward:

```{r blob}
library(blob)
df3 <- data.frame(id = 1:3)
df3$payload <- blob::blob(
  charToRaw("hello"),
  charToRaw("world"),
  charToRaw("!")
)
write_parquet(df3, tmp4 <- tempfile(fileext = ".parquet"))
as.data.frame(read_parquet(tmp4))
```

## nanoparquet as a filter

In Unix, a _filter_ is a program that reads from standard input and writes
to standard output, making it a composable building block in shell pipelines.
`write_parquet()` now supports writing to standard output via
`file = ":stdout:"`:

```{r stdout, eval = FALSE}
write_parquet(mtcars, ":stdout:")
```

The most common use case is from the command line:

```sh
Rscript --quiet -e 'nanoparquet::write_parquet(mtcars, ":stdout:")' > mtcars.parquet
```

You can build this into a data pipeline. For example, to convert a CSV
to Parquet, and then process Parquet with another tool in one shot,
without an intermediate `.parquet` file on the disk, you can do:

```sh
cat data.csv |
  Rscript --quiet -e '
    df <- read.csv(file("stdin"))
    nanoparquet::write_parquet(df, ":stdout:")
  ' | another-parquet-tool
```

Since nanoparquet 0.4.0, `read_parquet()` can also read from an R
connection, so you can pipe Parquet data _in_ as well:

```{r stdin, eval = FALSE}
url <- "https://raw.githubusercontent.com/r-lib/nanoparquet/main/inst/extdata/userdata1.parquet"
con <- pipe(paste("curl --silent", url))
df <- read_parquet(con)
```

## Acknowledgements

We thank all contributors to nanoparquet so far, for opening issues,
submitting pull requests, and providing feedback:
[&#x0040;Aariq](https://github.com/Aariq),
[&#x0040;alvarocombo](https://github.com/alvarocombo),
[&#x0040;apalacio9502](https://github.com/apalacio9502),
[&#x0040;atsyplenkov](https://github.com/atsyplenkov),
[&#x0040;cboettig](https://github.com/cboettig),
[&#x0040;ChandlerLutz](https://github.com/ChandlerLutz),
[&#x0040;cmrnp](https://github.com/cmrnp),
[&#x0040;D3SL](https://github.com/D3SL),
[&#x0040;damonbayer](https://github.com/damonbayer),
[&#x0040;DavideMessinaARS](https://github.com/DavideMessinaARS),
[&#x0040;eitsupi](https://github.com/eitsupi),
[&#x0040;gksmyth](https://github.com/gksmyth),
[&#x0040;hadley](https://github.com/hadley),
[&#x0040;jack-davison](https://github.com/jack-davison),
[&#x0040;jeroenjanssens](https://github.com/jeroenjanssens),
[&#x0040;lbm364dl](https://github.com/lbm364dl),
[&#x0040;lschneiderbauer](https://github.com/lschneiderbauer),
[&#x0040;mrcaseb](https://github.com/mrcaseb),
[&#x0040;pmarks](https://github.com/pmarks),
[&#x0040;PMassicotte](https://github.com/PMassicotte),
[&#x0040;r2evans](https://github.com/r2evans),
[&#x0040;RealTYPICAL](https://github.com/RealTYPICAL),
[&#x0040;tanho63](https://github.com/tanho63),
[&#x0040;thisisnic](https://github.com/thisisnic),
[&#x0040;torfason](https://github.com/torfason),
[&#x0040;TurnaevEvgeny](https://github.com/TurnaevEvgeny),
[&#x0040;Upipa](https://github.com/Upipa),
[&#x0040;vankesteren](https://github.com/vankesteren),
[&#x0040;vincentarelbundock](https://github.com/vincentarelbundock),
[&#x0040;wlandau](https://github.com/wlandau),
[&#x0040;YipengUva](https://github.com/YipengUva), and
[&#x0040;yutannihilation](https://github.com/yutannihilation).
