---
output: hugodown::hugo_document

slug: stringr-1-5-0
title: stringr 1.5.0
date: 2022-10-31
author: Hadley Wickham
description: >
    It's been three years but a new version of stringr is now on CRAN!
    This release includes a bunch of small but useful new functions and 
    some increased consistency with the rest of the tidyverse.

photo:
  url: https://unsplash.com/photos/XGqS569rdgk
  author: Amie Bell

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [stringr, tidyverse]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [ ] `hugodown::use_tidy_thumbnails()`
* [ ] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->
```
We're chuffed to announce the release of [stringr](https://stringr.tidyverse.org) 1.5.0.
stringr provides a cohesive set of functions designed to make working with strings as easy as possible.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("{package}")
```

It's been over three and half years since the last stringr release, and in that time we've accumulated a bunch of small but useful new functions in the developer version.
Thorough documentation review and a new vignette on transitioning from base R by Sara Stoudt in the 2019 Tidyverse developer day.
Finally, stringr is now officially [re-licensed as MIT](https://www.tidyverse.org/blog/2021/12/relicensing-packages/).

You can see a full list of changes in the [release notes](%7B%20github_release%20%7D).

```{r setup}
library(stringr)
```

## Breaking changes

Lets start with the important stuff: the breaking changes.
We've tried to keep these small and we don't believe they'll affect much code in the wild as they only affected \~20 of the \~1600 packages that use stringr.
But we're believe they're important as a consistent set of rules makes the tidyverse as a whole more predictable and easier to learn.

### Recycling rules

stringr functions now consistently implement the tidyverse recycling rules[^1], which are stricter than the previous rules in two ways.
Firstly, we no longer recycle shorter vectors that are an integer multiple of longer vectors:

[^1]: You might wonder why we developed our own set of recycling rules for the tidyverse instead of using the base R rules.
    That's because, unfortunately, there isn't a consistent set of rules used by base R, but a [suite of variations](https://vctrs.r-lib.org/articles/type-size.html#appendix-recycling-in-base-r).

```{r}
#| error: true

str_detect(letters, c("x", "y"))
```

Secondly, a 0-length vector doesn't automatically imply a 0-length output, it's instead recycled using the usual rules:

```{r}
#| error: true

str_detect(letters, character())
```

Neither of these situations occurs very commonly with stringr, so this change primarily brings consistency with the rest of the tidyverse without affecting much existing code.

There's one other small change that mostly affects `str_c()`: `NULL`s are ignored, rather than being treated like 0-length vectors.
This is mostly useful in programming.

```{r}
emphasise <- FALSE

str_c(
  "Hello",
  if (emphasise) "!"
)
```

Finally, stringr functions are generally a little stricter because we require the inputs to be vectors of some type.
Again, this is unlikely to affect your data analysis code and will result in a clearer error if you accidentally pass in something weird:

```{r}
#| error: TRUE

str_detect(mean, "x")
```

### Empty patterns

In many stringr functions, `""` will match or split on every character motivated.
This is motivated by base R's `strsplit()`:

```{r}
strsplit("abc", "")[[1]]
str_split("abc", "")[[1]]
```

When creating stringr (over 13 years ago!) I took this idea and ran with it, implementing similar support in every function where it might possibly work.
But I missed an important problem with `str_detect()`.

What should `str_detect(X, "")` return?
You can argue two ways:

-   To be consistent with `str_split()`, it should return `TRUE` whenever there are characters to match, i.e. `x != ""`.
-   It's not common to build up a set of possible matches by doing `str_flatten(matches, "|")`. What should this match if `matches` is empty? Ideally it would match nothing implying that `str_detect(x, "")` should be equivalent to `x == ""`.

This inconsistency potentially leads to some subtle bugs, so use of `""` in `str_detect()` (and a few other related functions) is now an error:

```{r}
#| error: true

str_detect(letters, "")
```

## New features

Now that we've got the breaking changes out of the way we can focus on the new features 😃.
The biggest improvement is to `str_view()` which has gained a bunch of new features, including using the cli package to work in more places.
We also have a grab bag of new functions.

### `str_view()`

`str_view()` uses ANSI colouring rather than an HTML widget which means it works in more places and requires fewer dependencies.
It:

-   Displays strings with special characters.

    ```{r}
    x <- c("\\", "\"\nabcdef\n\"")
    x

    str_view(x)
    ```

-   Highlights unusual whitespace characters.

    ```{r}
    str_view("\t")
    ```

-   Shows all matches.

    ```{r}
    str_view(fruit, "(.)\\1")
    ```

    (This makes `str_view_all()` redundant and hence deprecated.)

### Locale sensitive functions

There are three new functions related to comparing strings:

-   `str_equal()` compares two character vectors using unicode rules, optionally ignoring case

    ```{r}
    str_equal("a", "A")
    str_equal("a", "A", ignore_case = TRUE)
    ```

-   `str_rank()` completes the set of order/rank/sort functions:

    ```{r}
    str_rank(c("a", "c", "b", "b"))
    # compare to:
    str_order(c("a", "c", "b", "b"))
    ```

-   `str_unique()` returns unique values, optionally ignoring case:

    ```{r}
    str_unique(c("a", "a", "A"))
    str_unique(c("a", "a", "A"), ignore_case = TRUE)
    ```

### Splitting

`str_split()` gains two useful variants:

-   `str_split_1()` is tailored for the special case of splitting up a single string.
    It returns a character vector, not a list, and errors if you try and give it multiple values:

    ```{r}
    #| error: true

    str_split_1("x-y-z", "-")
    str_split_1(c("x-y", "a-b-c"), "-")
    ```

-   `str_split_i()` extracts a single piece from the splits string:

    ```{r}
    x <- c("a-b-c", "d-e", "f-g-h-i")
    str_split_i(x, "-", 2)
    str_split_i(x, "-", 4)
    ```

### Miscellaneous

-   `str_escape()` escapes regular expression metacharacters, providing an alternative to `fixed()` if you want to compose a pattern from user supplied strings:

    ```{r}
    str_view(str_escape("[]|"))
    ```

-   `str_extract()` can optionally extract a capturing group instead of the complete match:

    ```{r}
    x <- c("Chapter 1", "Section 2.3", "Chapter 3", "Section 4.1.1")
    str_extract(x, "([A-Za-z]+) ([0-9.]+)", group = 1)
    str_extract(x, "([A-Za-z]+) ([0-9.]+)", group = 2)
    ```

-   `str_flatten()` gains a `last` argument which is used to power the new `str_flatten_comma()`:

    ```{r}
    str_flatten_comma(c("cats", "dogs", "mice"))
    str_flatten_comma(c("cats", "dogs", "mice"), last = " and ")
    str_flatten_comma(c("cats", "dogs", "mice"), last = ", and ")

    # correctly handles the two element case with the Oxford comma
    str_flatten_comma(c("cats", "dogs"), last = ", and ")
    ```

-   `str_like()` works like `str_detect()` but uses SQL's LIKE syntax:

    ```{r}
    fruit <- c("apple", "banana", "pear", "pineapple")
    fruit[str_like(fruit, "%apple")]
    fruit[str_like(fruit, "p__r")]
    ```

## Acknowledgements
