---
title: bench 1.0.1
date: '2018-06-28'
slug: bench-1.0.1
author: Jim Hester
categories: [package]
description: >
  bench 1.0.1 is now available on CRAN. bench allows you to benchmark
  code, by tracking execution time, memory allocations and garbage
  collections.
photo:
  url: https://unsplash.com/photos/nIkuMWT4Imc
  author: Rula Sibai
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
options(width = 120)
```

[bench] is now available on CRAN!

The goal of [bench] is to benchmark code, by tracking execution time,
memory allocations and garbage collections.

Install the latest version with:
``` r
install.packages("bench")
```

### Usage

Benchmarks can be run with `bench::mark()`, which takes one or more expressions
to benchmark against each other.

```{r example}
library(bench)
set.seed(42)
dat <- data.frame(x = runif(10000, 1, 1000), y=runif(10000, 1, 1000))
```

`bench::mark()` will throw an error if the results are not equivalent, so you
don't accidentally benchmark non-equivalent code.
```{r example1-1, error = TRUE}
bench::mark(
  dat[dat$x > 500, ],
  dat[which(dat$x > 499), ],
  subset(dat, x > 500))
```

Results are easy to interpret, with human readable units in a rectangular data frame.
```{r example1-2}
bnch <- bench::mark(
  dat[dat$x > 500, ],
  dat[which(dat$x > 500), ],
  subset(dat, x > 500))
bnch
```

By default, the summary uses absolute measures, however relative results can be
obtained by using `relative = TRUE` in your call to `bench::mark()` or by calling
`summary(relative = TRUE)` on the results.
```{r example1-3}
summary(bnch, relative = TRUE)
```

`bench::press()` is used to run benchmarks against a grid of parameters.
Provide setup and benchmarking code as a single unnamed argument then define
sets of values as named arguments. The full combination of values will be
expanded and the benchmarks are then _pressed_ together in the result. This
allows you to benchmark a set of expressions across a wide variety of input
sizes, perform replications and other useful tasks.

```{r example2, cache = TRUE, message = FALSE}
set.seed(42)

create_df <- function(rows, cols) {
  as.data.frame(setNames(
    replicate(cols, runif(rows, 1, 1000), simplify = FALSE),
    rep_len(c("x", letters), cols)))
}

results <- bench::press(
  rows = c(10000, 100000),
  cols = c(10, 100),
  {
    dat <- create_df(rows, cols)
    bench::mark(
      min_iterations = 100,
      bracket = dat[dat$x > 500, ],
      which = dat[which(dat$x > 500), ],
      subset = subset(dat, x > 500)
    )
  }
)
results
```

### Plotting

`ggplot2::autoplot()` can be used to generate an informative default plot. This
plot is colored by GC level (0, 1, or 2) and faceted by parameters (if any). By
default it generates a
[beeswarm](https://github.com/eclarke/ggbeeswarm#geom_quasirandom) plot,
however you can also specify other plot types (`jitter`, `ridge`,
`boxplot`, `violin`). See `?autoplot.bench_mark` for full details. This gives
you a nice overview of the runs and allows you to gauge the effects of garbage
collection on the results.

```{r autoplot, message = FALSE, warning = FALSE, cache = TRUE, dependson = "example2"}
ggplot2::autoplot(results)
```

You can also produce fully custom plots by un-nesting the results and
working with the data directly. In this case we are exploring how the amount of
memory allocated by each expression interacts with the time taken to run.

```{r custom-plot, message = FALSE, cache = TRUE, dependson = "example2"}
library(tidyverse)
results %>%
  unnest() %>%
  filter(gc == "none") %>%
  ggplot(aes(x = mem_alloc, y = time, color = expression)) +
    geom_point() +
    scale_color_brewer(type = "qual", palette = 3) +
    geom_smooth(method = "lm", se = F, colour = "grey50")
```

### Compared to existing methods

Compared to other methods such as [system.time], [rbenchmark], [tictoc] or
[microbenchmark] we feel it has a number of benefits.

- Uses the highest precision APIs available for each operating system (often nanosecond-level).
- Tracks memory allocations for each expression.
- Tracks the number and type of R garbage collections per run.
- Verifies equality of expression results by default, to avoid accidentally
  benchmarking non-equivalent code.
- Uses adaptive stopping by default, running each expression for a set amount
  of time rather than for a specific number of iterations.
- Runs expressions in batches and calculates summary statistics after
  filtering out iterations with garbage collections. This allows you to isolate
  the performance and effects of garbage collection on running time (for more details see [Neal
  2014](https://radfordneal.wordpress.com/2014/02/02/inaccurate-results-from-microbenchmark/)).
- Allows benchmarking across a grid of input values with `bench::press()`.

### Dependency load

When the development version of **bench** was
[introduced](https://twitter.com/jimhester_/status/996063591433416704) a few
people expressed concern over the number of dependencies in the package. I will
attempt to explain why these dependencies exist and why the true load may actually
be less than you might think.

While bench currently has 19 dependencies, only 8 of these are hard
dependencies; that is they are needed to install the package. Of these 8 hard
dependencies 3 of them (methods, stats, utils) are base packages installed with
R. Of these 5 remaining packages 3 have no additional dependencies (glue,
profmem, rlang). The two remaining packages (tibble and pillar) are used to
provide nice printing of the times and memory sizes and support for list
columns to store the timings, garbage collections, and allocations. These are
major features of the bench package and it would not work without these
dependencies.

The remaining 11 packages are soft dependencies, used either for testing or for
optional functionality, most notably plotting. They will not be installed
unless explicitly requested.

The [microbenchmark] package is a good alternative for those looking for a
package with only base dependencies.

### Feedback wanted!

We hope **bench** is a useful tool for benchmarking short expressions of code.
Please open [GitHub issues](https://github.com/r-lib/bench) for any feature
requests or bugs.

Learn more about **bench** at http://bench.r-lib.org

A big thanks goes to all the community members who contributed code and opened
issues since for this release!
[\@espinielli](https://github.com/espinielli),
[\@hadley](https://github.com/hadley),
[\@HughParsonage](https://github.com/HughParsonage),
[\@jasonserviss](https://github.com/jasonserviss),
[\@jimhester](https://github.com/jimhester),
[\@jonocarroll](https://github.com/jonocarroll),
[\@lionel-](https://github.com/lionel-),
[\@MilesMcBain](https://github.com/MilesMcBain),
[\@njtierney](https://github.com/njtierney), and
[\@zkamvar](https://github.com/zkamvar)

[bench]: https://bench.r-lib.org
[rbenchmark]: https://cran.r-project.org/package=rbenchmark
[tictoc]: https://cran.r-project.org/package=tictoc
[microbenchmark]: https://cran.r-project.org/package=microbenchmark
[system.time]: https://www.rdocumentation.org/packages/base/versions/3.5.0/topics/system.time
