---
output: hugodown::hugo_document

slug: archive-1-1-0
title: archive 1 1 0
date: 2021-11-04
author: Jim Hester
description: >
    archive 1.1.0 is now on CRAN! archive lets you work with file archives, such as ZIP, tar, 7-Zip and RAR and compression formats like gzip, bzip2, XZ and Zstandard.

photo:
  url: https://unsplash.com/photos/IEiAmhXehwE
  author: Nana Smirnova
  

# one of: "deep-dive", "learn", "package", "programming", or "other"
categories: [package] 
tags: [r-lib]
---

[archive](https://archive.r-lib.org/) 1.1.0 is now on CRAN. archive lets you work with file archives, such as [ZIP](https://en.wikipedia.org/wiki/ZIP_(file_format)), [tar](https://en.wikipedia.org/wiki/Tar_(computing)), [7-Zip](https://en.wikipedia.org/wiki/7-Zip) and [RAR](https://en.wikipedia.org/wiki/RAR_(file_format)) and compression formats like [gzip](https://en.wikipedia.org/wiki/Gzip), [bzip2](https://en.wikipedia.org/wiki/Bzip2), [XZ](https://en.wikipedia.org/wiki/XZ_Utils) and [Zstandard](https://en.wikipedia.org/wiki/Zstandard).
It does this by building on top of the [libarchive](https://www.libarchive.org/) C library.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("archive")
```

This blog post will explain the main functions of archive, and show how you can use them to read from and write to archives.

You can see a full list of changes in the [release notes](https://archive.r-lib.org/news/index.html)

```{r setup}
library(archive)
my_dir <- fs::file_temp() |> fs::dir_create()
knitr::opts_knit$set(root.dir = my_dir)
```

## Displaying archive contents

Use `archive()` to return a tibble of the files contained in a given archive.

```{r, include = FALSE}
library(nycflights13)
fs::dir_create("nycflights13")
for (data in data(package="nycflights13")$results[,"Item"]) {
  readr::write_csv(head(get(data), n = 1000), fs::path("nycflights13", data, ext = "csv"))
}
archive_write_files("nycflights13.zip", fs::dir_ls("nycflights13"))
```

```{r}
archive("nycflights13.zip")
```

## Reading single files from an archive

`archive_read()` is used to read a single file from an archive. This function returns an R connection, which can be passed to many R functions that takes a connection object as input. All base R file system functions use connections, as well as some packages like readr.

The `file=` argument accepts numeric positions in the archive, or filenames as input.

```{r}
readLines(
  archive_read("nycflights13.zip", file = 2),
  n = 5
)

readLines(
  archive_read("nycflights13.zip", file = "nycflights13/planes.csv"),
  n = 5
)
```

## Writing single files to an archive

Similarly `archive_write()` is used to write a single file to an archive. Again this creates a writable R connection. Like reading, many base R functions work with writable connections, as well as some packages like readr.

The archive and compression formats are automatically guessed based on the output filename file extensions. However you can also specify them explicity with the `format` and `filter` options.

Here we create a new zip archive containing the file `mtcars.csv`.
```{r}
readr::write_csv(mtcars, archive_write("my-cars.zip", "mtcars.csv"))

archive("my-cars.zip")
```

## Writing multiple files to an archive

`archive_write_files()` writes multiple files to a new archive. In this case the files to be added to the archive should already be written on disk.

`archive_write_dir()` is a helper to archive all the files in a given directory.

```{r}
library(readr)

# Write a few files to the temp directory
write_csv(iris, "iris.csv")
write_csv(mtcars, "mtcars.csv")
write_csv(airquality, "airquality.csv")

# Add them to a new XZ compressed tar archive
archive_write_files("data.tar.xz",
                    c("iris.csv", "mtcars.csv", "airquality.csv"))

# View archive contents
archive("data.tar.xz")
```

## Extracting multiple files from an archive

`archive_extract()` allows you to extract one or more files to disk from an archive.

Note the archive and compression formats will be automatically detected.

```{r}
# Create a new directory
my_dir <- fs::file_temp() |> fs::dir_create()

# Extract two of the files in the archive to that directory
archive_extract("data.tar.xz", dir = my_dir, files = c("iris.csv", "mtcars.csv"))

# Show the extracted files
fs::dir_ls(my_dir) |> fs::path_file()
```
```{r cleanup, include = FALSE}
unlink(my_dir, recursive = TRUE)
```
## Acknowledgements

Thanks to the following users who have tried out the development versions of archive and opened issues and feature suggestions to improve it! [&#x0040;cboettig](https://github.com/cboettig), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;jeroen](https://github.com/jeroen), and [&#x0040;JMcrocs](https://github.com/JMcrocs).
