---
output: hugodown::hugo_document
slug: ggplot2-3-4-0-size-to-linewidth
title: Make your ggplot2 extension package understand the new linewidth aesthetic
date: 2022-08-24
author: Thomas Lin Pedersen
description: >
    The next release of ggplot2 will contain a number of internal improvements 
    and fixes long-time inconsistencies. One of these are the conflation of 
    point size and linewidth into the same aesthetic. This post will go into 
    detail with how you can make your extension package work well with the new
    linewidth aesthetic.
photo:
  url: https://unsplash.com/photos/GsZLXA4JPcM
  author: Ricardo Gomez Angel
# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [deep-dive] 
tags: [ggplot2]
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
set.seed(340)
options(rlib_message_verbosity = "verbose")
options(lifecycle_verbosity = "warning")
library(rlang)
```

We are hard at work finishing the next release of ggplot2. While this
release is mostly about internal changes, there are a few quite user
visible changes as well. One of these upends the idea that the `size`
aesthetic is responsible for *both* the sizing of point/text and the
width of lines. With the next release we will have a `linewidth`
aesthetic to take care of the latter, while `size` will continue
handling the former.

There are many excellent reasons for this change, all of which will have
to wait until the release post to be discussed. This blog post is for
those that maintain an extension package for ggplot2 and are left
wondering how they should respond to this --- if that is you, please
read on!

## The way it works

Before going into technicalities we'll describe how it is intended to
work. We are well aware that we can't just make a change that would
instantly break everyone's code. So, we have gone to great length to
make old code work as before while gently coercing users into adopting
the new paradigm. For example, take a look at this old code:

```{r}
library(ggplot2)

ggplot(airquality) + 
  geom_line(
    aes(x = Day, y = Temp, size = Wind, group = Month), 
    lineend = "round"
  )
```

As you can see, ggplot2 detects the use of the `size` aesthetic and
informs the user about the new `linewidth` aesthetic but otherwise
proceeds as before, producing the expected plot. As expected,
`scale_size()` also works in this situation:

```{r}
ggplot(airquality) + 
  geom_line(
    aes(x = Day, y = Temp, size = Wind, group = Month), 
    lineend = "round"
  ) + 
  scale_size("Windspeed (mph)", range = c(0.5, 3))
```

but ultimately we want users to migrate to the following code:

```{r}
ggplot(airquality) + 
  geom_line(
    aes(x = Day, y = Temp, linewidth = Wind, group = Month), 
    lineend = "round"
  ) + 
  scale_linewidth("Windspeed (mph)", range = c(0.5, 3))
```

> Note that there's an important difference between these two plots (and
> one of the reasons we're making the change): The last two plots differ
> because the default `linewidth` scale correctly use a linear transform
> instead of a square root transform (which is only sensible for scaling
> of areas).

## How to adopt this

We have been able to add this automatic translation in a quite
non-intrusive way which means that you as a package developer don't need
to do that much to adapt to the new naming. To show this I'll create a
geom drawing circles then update it to use linewidth instead:

```{r}
GeomCircle <- ggproto("GeomCircle", Geom,
  draw_panel = function(data, panel_params, coord) {
    # Expand x, y, radius data to points along circle
    circle_data <- Map(function(x, y, r) {
      radians <- seq(0, 2*pi, length.out = 101)[-1]
      data.frame(
        x = cos(radians) * r + x,
        y = sin(radians) * r + y
      )
    }, x = data$x, y = data$y, r = data$radius)
    
    circle_data <- do.call(rbind, circle_data)
    
    # Transform to viewport coords
    circle_data <- coord$transform(circle_data, panel_params)
    
    # Draw as grob
    grid::polygonGrob(
      x = circle_data$x,
      y = circle_data$y,
      id.lengths = rep(100, nrow(data)),
      default.units = "native",
      gp = grid::gpar(
        col = data$colour,
        fill = data$fill,
        lwd = data$size * .pt,
        lty = data$linetype
      )
    )
  },
  required_aes = c("x", "y", "radius"),
  default_aes = aes(
    colour = "black",
    fill = "grey",
    size = 0.5,
    linetype = 1,
    alpha = NA
  ),
  draw_key = draw_key_polygon
)

geom_circle <- function(mapping = NULL, data = NULL, stat = "identity", 
                        position = "identity", ..., na.rm = FALSE, 
                        show.legend = NA, inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomCircle,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      na.rm = na.rm,
      ...
    )
  )
}
```

As a sanity check, let us check that this actually works:

```{r}
random_points <- data.frame(
  x = runif(20),
  y = runif(20),
  radius = runif(20, max = 0.1),
  value = runif(20)
)

ggplot(random_points) + 
  geom_circle(aes(x = x, y = y, radius = radius, size = value))
```

It seems to work as intended. As can be seen from the code above, the
`size` aesthetics is not used much and is passed directly into
`polygonGrob()`. It follows that updating the code to using linewidth is
not a huge operation.

> There is nothing preventing you from keeping the code as is --- it
> will continue to work as always. However, your users may begin to feel
> a disconnect with the style as they adapt to the new `linewidth`
> aesthetic so it is highly recommended to make the proposed changes

### The fix

There are a few things you need to do to update the old code but they
are all pretty benign. The changes are commented in the code below and
will also be discussed afterwards.

```{r}
GeomCircle <- ggproto("GeomCircle", Geom,
  draw_panel = function(data, panel_params, coord) {
    # Expand x, y, radius data to points along circle
    circle_data <- Map(function(x, y, r) {
      radians <- seq(0, 2*pi, length.out = 101)[-1]
      data.frame(
        x = cos(radians) * r + x,
        y = sin(radians) * r + y
      )
    }, x = data$x, y = data$y, r = data$radius)
    
    circle_data <- do.call(rbind, circle_data)
    
    # Transform to viewport coords
    circle_data <- coord$transform(circle_data, panel_params)
    
    # Draw as grob
    grid::polygonGrob(
      x = circle_data$x,
      y = circle_data$y,
      id.lengths = rep(100, nrow(data)),
      default.units = "native",
      gp = grid::gpar(
        col = data$colour,
        fill = data$fill,
        # Use linewidth or fall back to size in old ggplot2 versions
        lwd = (data$linewidth %||% data$size) * .pt,
        lty = data$linetype
      )
    )
  },
  required_aes = c("x", "y", "radius"),
  default_aes = aes(
    colour = "black",
    fill = "grey",
    # Switch size to linewidth
    linewidth = 0.5,
    linetype = 1,
    alpha = NA
  ),
  draw_key = draw_key_polygon,
  # To allow using size in ggplot2 < 3.4.0
  non_missing_aes = "size",
  
  # Tell ggplot2 to perform automatic renaming
  rename_size = TRUE
)
```

As we can see above, we need two changes and two additions to our
implementation. First (but last in the code), we add
`rename_size = TRUE` to our geom implementation. This instructs ggplot2
that this layer has a `size` aesthetic that should be converted
automatically with a deprecation warning. Setting this to `TRUE` allows
you to rest assured that as far as your code goes you can expect to have
a `linewidth` aesthetic. Second, we update the `default_aes` to use
`linewidth` instead of `size`. Third, wherever we use `size` in our geom
logic we instead use `linewidth %||% size`. The reason for the fallback
is that if your package is used together with an older version of
ggplot2 the `rename_size = TRUE` line has no effect and you need to fall
back to `size` if that is what the user has specified. Fourth, we add
`size` to the `non_missing_aes` field. As with the last point, this is
only relevant for use with older versions of ggplot2 as it instructs the
geom to not warn when `size` is used.

Let's try out the new implementation:

```{r}
ggplot(random_points) + 
  geom_circle(aes(x = x, y = y, radius = radius, size = value))
```

We see that we get the deprecation warning we know and that everything
also renders as expected. Using the new naming also works, picks up the
linear `linewidth` scale, and doesn't have a warning.

```{r}
ggplot(random_points) + 
  geom_circle(aes(x = x, y = y, radius = radius, linewidth = value))
```

The legend looks a bit wonky, but that is because the polygon key
function caps the linewidth at a certain size relative to the size of
the key. We can see that it works fine using a lower range:

```{r}
last_plot() + scale_linewidth(range = c(0.1, 2))
```

## FAQ

*I'm creating a geom as a subclass of one of the ggplot2 geoms that now
uses `linewidth` --- what should I do?*

If your geom inherits from e.g. `geom_polygon()` which in the next
version will begin using `linewidth` all you have to do is to update
your code to refer to `linetype` instead of `size` if it uses that
anywhere. Your geom will already inherit the correct `rename_size`
value.

*I'm creating a stat --- should I do anything?*

Probably not. The only exception is if you set `size` in `default_aes`
to a calculated value and the expectance is that the geom used with the
stat will change to using `linewidth`. In such situations you should
change the `default_aes` setting to use `linewidth` instead. We haven't
had any such situations in the ggplot2 code base so the chance of this
being relevant is pretty low.

*I'm creating a geom that uses `size` for both point sizing and line
width --- how should I proceed?*

If you have a geom where `size` doubles for both point sizes and
linewidth (an example from ggplot2 is `geom_pointrange()`) you shouldn't
set `rename_size = TRUE` since `size` remains a valid aesthetic.
However, you should add `linewidth` to `default_aes` and use this
wherever in your code `size` was used for linewidth scaling before. Do
note that this is a breaking change for your users since the same piece
of code may no longer produce the same output.
