---
title: withr 2.1.0
date: '2017-11-16'
slug: withr-2.1.0
author: Jim Hester
categories: [package]
description: >
  withr 2.1.0 is now available on CRAN.
photo:
  url: https://unsplash.com/photos/V1YEEItoOTE
  author: Suganth
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE
)
library(withr)
```

[withr 2.1.0](http://withr.r-lib.org/news/index.html) is now available on CRAN!
[withr](http://withr.r-lib.org) makes working with global state in R safer and
less error prone. It has only base package dependencies so is easily included
in packages.

Install the latest version with:
```{r, eval = FALSE}
install.packages("withr")
```

# Whither withr?

Many functions in R modify global state in some fashion. Some common examples
are `par()` for graphics parameters, `dir()` to change the current directory
and `options()` to set a global option. Using these functions is handy
when using R interactively, because you can set them early in your
experimentation and they will remain set for the duration of the session.
However this makes programming with these settings difficult, because they make
your function impure by modifying a global state. Therefore you should always
strive to reset the previous state when the function exits.

One common idiom for dealing with this problem is to save the current state,
make your change, then restore the previous state.

```{r}
par("col" = "black")
my_plot <- function(new) {
  old <- par(col = "red", pch = 19)
  plot(mtcars$hp, mtcars$wt)
  par(old)
}
my_plot()
par("col")
```

However this approach can fail if there's an error before you are able to reset
the options.

```{r, error = TRUE}
par("col" = "black")
my_plot <- function(new) {
  old <- par(col = "red", pch = 19)
  plot(mtcars$hpp, mtcars$wt)
  par(old)
}
my_plot()
par("col")
```

Using the base function `on.exit()` is a robust solution to this problem.
`on.exit()` will run the code when the function is exited, regardless
of whether it exits normally or with an error.

```{r, error = TRUE}
par("col" = "black")
my_plot <- function(new) {
  old <- par(col = "red", pch = 19)
  on.exit(par(old))
  plot(mtcars$hpp, mtcars$wt)
}
my_plot()
par("col")

options(test = 1)
{
  print(getOption("test"))
  on.exit(options(test = 2))
}
getOption("test")
```

However this solution is somewhat cumbersome to work with. You
need to remember to use an `on.exit()` call after each stateful call. In
addition by default each `on.exit()` action will overwrite any previous
`on.exit()` action in the same function unless you use the `add = TRUE` option.
`add = TRUE` also adds additional code to the _end_ of existing code, which
means the code is not run in the [Last-In,
First-Out](https://en.wikipedia.org/wiki/FIFO_and_LIFO_accounting) order you
would generally prefer. It is also not possible to have this cleanup code
performed before the function has finished.

[withr](http://withr.r-lib.org) is a solution to these issues. It defines a
[large set of
functions](http://withr.r-lib.org/#withr---run-code-with-modified-state) for
dealing with global settings in R, such as `with_par()`. These functions set one of
the global settings for the duration of a block of code, then automatically
reset it after the block is completed.

```{r}
par("col" = "black")
my_plot <- function(new) {
  with_par(list(col = "red", pch = 19),
    plot(mtcars$hp, mtcars$wt)
  )
  par("col")
}
my_plot()
par("col")
```

In addition to the `with_*` functions there are `local_*` variants whose effects
last until the end of the function they are included in. These work similar to
`on.exit()`, but you can set the options in one call rather than two.

```{r}
par("col" = "black")
my_plot <- function(new) {
  local_par(list(col = "red", pch = 19))
  plot(mtcars$hp, mtcars$wt)
}
my_plot()
par("col")
```

# New features

Here are some highlights of new functions for v2.1.0.

## Graphics devices

There are now a comprehensive set of functions to deal with R's builtin
[graphics devices](http://withr.r-lib.org/reference/devices.html).

These functions open a new graphics device, run some code, then automatically
close the device.

```{r}
path <- file.path(Sys.glob("*withr*_files/figure-html"), "test.png")
with_png(path, width = 400, height = 300, {
  plot(mtcars$hp, mtcars$mpg)
})
```

![](`r path`)

Thanks to [Richard Cotton's](https://github.com/richierocks) great [pull
request](https://github.com/r-lib/withr/pull/37) for this feature!

## Connections

There are two new functions for cleaning up connections in R.
`with_connection()` allows you to automatically close R's file connections.
Here we create a writable file connection, write some lines to it with
`with_connection()`, then open a read-only connection and read the file using
`local_connection()`.

```{r}
with_connection(list(con = file("temp", "w")), {
  writeLines(c("foo", "bar"), con)
})
read_temp <- function() {
  con <- local_connection(file("temp", "r"))
  readLines(con)
}
read_temp()
```
```{r, echo = FALSE}
unlink("temp")
```

`with_db_connection()` provides [DBI](http://rstats-db.github.io/DBI)
connections to databases which automatically call
`DBI::dbDisconnect()`. Here we create a new [SQLite](https://www.sqlite.org/)
database, connect to it with `with_db_connection()`, and write a new table to
it. We then create another connection with `local_db_connection()` and read
from the table.

```{r}
db <- tempfile()
with_db_connection(
  list(con = DBI::dbConnect(RSQLite::SQLite(), db)), {
    DBI::dbWriteTable(con, "mtcars", mtcars)
})

head_db_table <- function(...) {
  con <- local_db_connection(DBI::dbConnect(RSQLite::SQLite(), db))
  head(DBI::dbReadTable(con, "mtcars"), ...)
}
head_db_table()
unlink(db)
```

## Packages

`with_package()` allows you to temporarily attach a package.

```{r}
with_package("lattice", {
  xyplot(y ~ x, data.frame(x = -2:2, y = dnorm(-2:2)))
})
```

## Tempfiles

`with_tempfile()` handy for creating a new temporary files that are removed,
often useful when writing tests.

```{r}
with_tempfile("file1", {
  print(file1)
  writeLines("foo", file1)
  readLines(file1)
})
```

## Other changes

There are many other bug fixes and other minor improvements in this release.
You can see a complete list in the [release
notes](https://github.com/r-lib/withr/releases/tag/v2.1.0).

A big thanks goes to all the community members who contributed code and opened
issues since the last release!

```{r, eval = FALSE, include = FALSE}
x <- gh::gh("/repos/:owner/:repo/issues", owner = "r-lib", repo = "withr", since = "2017-06-16", state = "all")
users <- unique(purrr::map_chr(x, c("user", "login")))
clipr::write_clip(glue::collapse(glue::glue("[\\@{users}](https://github.com/{users})"), ", ", last = ", and "))
```

[\@QuLogic](https://github.com/QuLogic),
[\@krlmlr](https://github.com/krlmlr),
[\@hadley](https://github.com/hadley),
[\@wlandau-lilly](https://github.com/wlandau-lilly),
[\@jimhester](https://github.com/jimhester),
[\@kevinushey](https://github.com/kevinushey),
and [\@richierocks](https://github.com/richierocks)
