---
output: hugodown::hugo_document
slug: workflows-0-2-0
title: workflows 0.2.0
date: 2020-09-16
author: Davis Vaughan
description: >
    workflows 0.2.0 is now on CRAN!
photo:
  url: https://unsplash.com/photos/nN5L5GXKFz8
  author: Mark Fletcher-Brown
categories: [package] 
tags: [tidymodels]
editor_options: 
  chunk_output_type: console
---

We're excited to announce the release of [workflows](https://workflows.tidymodels.org/) 0.2.0. workflows is a [tidymodels](https://www.tidymodels.org/) package for bundling a model specification from [parsnip](https://parsnip.tidymodels.org/) with a preprocessor, such as a formula or [recipe](https://recipes.tidymodels.org/). Doing this can streamline the model fitting workflow and combines nicely with [tune](https://tune.tidymodels.org/) for performing hyperparameter tuning.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("workflows")
```

```{r setup, warning=FALSE}
library(workflows)
library(parsnip)
```

## Variables

The main change in this release of workflows is the introduction of a new preprocessor method: `add_variables()`. This adds a third method to specify model terms, in addition to `add_formula()` and `add_recipe()`. 

`add_variables()` has a tidyselect interface, where `outcomes` are specified using bare column names, followed by `predictors`.

```{r}
linear_spec <- linear_reg() %>%
  set_engine("lm")

wf <- workflow() %>%
  add_model(linear_spec) %>%
  add_variables(mpg, c(cyl, disp))

wf
```

```{r}
model <- fit(wf, mtcars)
mold <- pull_workflow_mold(model)

mold$predictors
mold$outcomes
```

`outcomes` are removed before `predictors` is evaluated, which means that formula specifications like `y ~ .` can be easily reproduced as:

```{r}
workflow() %>%
  add_variables(mpg, everything())
```

Importantly, `add_variables()` doesn't do any preprocessing to your columns whatsoever. This is in contrast to `add_formula()`, which uses the standard `model.matrix()` machinery from R, and `add_recipe()`, which will `recipes::prep()` the recipe for you. It is especially useful when you aren't using a recipe, but you do have S3 columns that you don't want run through `model.matrix()` for fear of losing the S3 class, like with Date columns.

```{r, warning=FALSE}
library(modeltime)

arima_spec <- arima_reg() %>%
    set_engine("arima")

df <- data.frame(
  y = sample(5),
  date = as.Date("2019-01-01") + 0:4
)

wf <- workflow() %>%
  add_variables(y, date) %>%
  add_model(arima_spec)

arima_model <- fit(wf, df)

arima_model
```

```{r}
mold <- pull_workflow_mold(arima_model)
mold$predictors
```

## Tune

workflows created with `add_variables()` do not work with the current CRAN version of tune (0.1.1). However, the development version of tune does have support for this, which you can install in the meantime until a new version of tune hits CRAN.

```{r, eval = FALSE}
devtools::install_github("tidymodels/tune")
```

## Acknowledgements

Thanks to the three contributors that helped with this version of workflows [&#x0040;EmilHvitfeldt](https://github.com/EmilHvitfeldt), [&#x0040;mdancho84](https://github.com/mdancho84), and [&#x0040;RaviHela](https://github.com/RaviHela)!
