---
output: hugodown::hugo_document

slug: forcats-1-0-0
title: forcats 1.0.0
date: 2023-01-12
author: Hadley Wickham
description: >
    There are no major new features in this version of forcats, but the 1.0.0
    label now clearly advertises that this a stable member of the tidyverse.

photo:
  url: https://unsplash.com/photos/NWwv0ETyzxc
  author: Diego Morales

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [forcats]
---

```{=html}
<!--
TODO:
* [ ] Look over / edit the post's title in the yaml
* [ ] Edit (or delete) the description; note this appears in the Twitter card
* [ ] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [ ] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->
```
We're so happy to announce the release of [forcats](https://forcats.tidyverse.org) 1.0.0.
The goal of the forcats package is to provide a suite of tools that solve common problems with factors, including changing the order of levels or the values.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("forcats")
```

While this is the 1.0.0 release of forcats, this is mainly a signal that we think forcats is stable, and don't anticipated any major changes in the future.
This blog post will outline the only major new features in this version: `fct_na_level_to_value()` and `fct_na_value_to_level()`.
As usual, you can see a full list of changes in the [release notes](https://github.com/tidyverse/forcats/releases/tag/v1.0.0).

```{r setup}
library(forcats)
```

## `NA` in levels vs `NA` in values

There are two ways to represent a missing value in a factor:

-   You can include it in the values of the factor; it does not appear in the levels and `is.na()` reports it as missing.
    This is how missing values are encoded in a factor by default:

    ```{r}
    f1 <- factor(c("x", "y", NA))
    levels(f1)
    is.na(f1)
    ```

-   You can include it in the levels of the factor and `is.na()` does not report it as missing.
    This requires a little more work to create, because, by default, `factor()` uses `exclude = NA`, meaning that missing values are not included in the levels.
    You can force `NA` to be included by setting `exclude = NULL`:

    ```{r}
    f2 <- factor(c("x", "y", NA), exclude = NULL)
    levels(f2)
    is.na(f2)
    ```

You can see the difference a little more clearly by looking at the underlying integer value of the factor:

```{r}
as.integer(f1)
as.integer(f2)
```

When the `NA` is stored in the levels, there's no missing value in the underlying integer values, because the value of level 3 is `NA`.

`NA`s in the values tend to be best for data analysis, since `is.na()` works as you'd expect.
`NA`s in the levels can be useful for display if you need to control where they appear in a table or a plot.
forcats now comes with tools for converting between these two forms: `fct_na_value_to_level()` and `fct_na_level_to_value()`.

Here's a practical example of why it matters from `vignette("forcats")`.
In the plot below, I've attempted to use `fct_infreq()` to reorder the levels of the factor so that the highest frequency levels are at the top of the bar chart:

```{r fct-infreq-hair}
#| fig-asp: 1
#| fig-width: 4
#| fig.alt: >
#|   The bar chart of hair color, now ordered so that the least
#|   frequent colours come first and the most frequent colors come last.
#|   This makes it easy to see that the most common hair color is none 
#|   (~35), followed by brown (~18), then black (~12). Surprisingly,
#|   NAs are at the top of the graph, even though there are ~5 NAs and 
#|   other colors have smaller values.
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

ggplot(starwars, aes(y = fct_rev(fct_infreq(hair_color)))) + 
  geom_bar() + 
  labs(y = "Hair color")
```

But unfortunately, because the `NA`s are stored in the values, `fct_infreq()` has no ability to affect them, and hence they appear in their default position.

We can make `fct_infreq()` do what we want by making an explicit `NA` level:

```{r}
#| fig-asp: 1
#| fig-width: 4
#| fig-alt: >
#|   The bar chart of hair color, now ordered so that NAs are 
#|   ordered where you'd expect: in between white (4) and black (12).
ggplot(starwars, aes(y = fct_rev(fct_infreq(fct_na_value_to_level(hair_color))))) + 
  geom_bar() + 
  labs(y = "Hair color")
```

That code is getting a little verbose so we might want to pull it out into a separate dplyr step:

```{r}
#| fig-asp: 1
#| fig-width: 4
#| results: false
starwars |> 
  mutate(
    hair_color = hair_color |> 
      fct_na_value_to_level() |> 
      fct_infreq() |> 
      fct_rev()
  ) |> 
  ggplot(aes(y = hair_color)) + 
  geom_bar() + 
  labs(y = "Hair color")
```

Pulling the factor manipulation out into it's own mini-pipeline makes it easier to make other adjustments.
For example, the code below uses a more informative label for the missing level, and lumps together all the hair colours with less than 2 observations.
I've left the (Other) category as a bar at the end, but by rearranging the other of `fct_infreq()` and `fct_lump_min()` I could other with the others, if I wanted.

```{r}
#| fig-width: 4
#| fig-alt: >
#|   The bar chart of hair color, with NA hair colour now labelled as (Unknown)
#|   and the low frequency bars lumped into (Other).
starwars |> 
  mutate(
    hair_color = hair_color |> 
      fct_na_value_to_level("(Unknown)") |> 
      fct_infreq() |> 
      fct_lump_min(2, other_level = "(Other)") |> 
      fct_rev() 
  ) |> 
  ggplot(aes(y = hair_color)) + 
  geom_bar() + 
  labs(y = "Hair color")
```

Looking closely at what got lumped together made me realise that there's also an existing "Unknown" level that should probably be represented as a missing value.
One way to fix that is with `fct_na_level_to_value()`:

```{r}
#| fig-width: 4
#| fig-alt: >
#|   The bar chart of hair color, with "unknown" hair colour now lumped in
#|   with (Unknown) instead of other
starwars |> 
  mutate(
    hair_color = hair_color |> 
      fct_na_level_to_value("Unknown") |> 
      fct_na_value_to_level("(Unknown)") |> 
      fct_infreq() |> 
      fct_lump_min(2, other_level = "(Other)") |> 
      fct_rev() 
  ) |> 
  ggplot(aes(y = hair_color)) + 
  geom_bar() + 
  labs(y = "Hair color")
```

## Acknowledgements
