---
output: hugodown::hugo_document

slug: magrittr-2-0
title: magrittr 2.0 is coming soon
date: 2020-07-30
author: Lionel Henry
description: >
    A new version of the magrittr package brings laziness, better performance, and leaner backtraces for debugging errors.

photo:
  url: https://unsplash.com/photos/X-NAMq6uP3Q
  author: Mike Benna

categories: [package]
tags: []
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

It is with unclouded exhilaration that we announce the upcoming release of [magrittr](https://magrittr.tidyverse.org/) 2.0. magrittr is the package home to the `%>%` pipe operator written by Stefan Milton Bache and used throughout the tidyverse.

This a major rewrite which adds laziness, gives better backtraces, and has much improved performance. Our testing suggests that it should be a drop in replacement, but we'd really like you to try it out and give us some feedback before we submit to CRAN. You can try the development version from GitHub with:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("magrittr")
```

If you discover any issues, please let us know by posting issues on the [Github repository](https://github.com/tidyverse/magrittr) of magrittr.

This blog post covers the three main changes in this new version of the magrittr pipe.

```{r setup}
library(magrittr)
```


## Laziness

The main user-visible change in this release is that the pipe expressions are now evaluated lazily, only when needed.

```{r}
ignore_arguments <- function(...) "value"

stop("foo") %>% ignore_arguments()
```

This has subtle implications but should be backward compatible with existing pipelines that run without error. The main source of behaviour change is that some code that previously failed may stop failing if the latter part of the pipeline specifically handled the error.

Similarly, warnings that were previously issued might now be suppressed by a function you're piping into. That's because the following expressions are now almost completely equivalent:

```{r}
# Piped
warning("foo") %>% suppressWarnings()

# Nested
suppressWarnings(warning("foo"))
```

Thanks to this change, you will now be able to pipe into testthat error expectations, for instance:

```{r}
library(testthat) %>%
  suppressMessages()

{ 1 + "a" } %>%
  expect_error("non-numeric argument")
```


## Backtraces

The R implementation of the magrittr pipe was rather costly in terms of backtrace clutter. This made it difficult to debug errors with functions using the pipe:

```{r, eval = FALSE}
foo <- function() bar()
bar <- function() 1 %>% identity() %>% baz()
baz <- function(x) rlang::abort("oh no")

foo()
#> Error: oh no

rlang::last_trace()
#> <error/rlang_error>
#> oh no
#> Backtrace:
#>      █
#>   1. └─global::foo()
#>   2.   └─global::bar()
#>   3.     └─1 %>% identity() %>% baz()
#>   4.       ├─base::withVisible(eval(quote(`_fseq`(`_lhs`)), env, env))
#>   5.       └─base::eval(quote(`_fseq`(`_lhs`)), env, env)
#>   6.         └─base::eval(quote(`_fseq`(`_lhs`)), env, env)
#>   7.           └─`_fseq`(`_lhs`)
#>   8.             └─magrittr::freduce(value, `_function_list`)
#>   9.               ├─base::withVisible(function_list[[k]](value))
#>  10.               └─function_list[[k]](value)
#>  11.                 └─global::baz(.)
```

This clutter is now completely resolved:

```{r, eval = FALSE}
foo()
#> Error: oh no

rlang::last_trace()
#> <error/rlang_error>
#> oh no
#> Backtrace:
#>     █
#>  1. ├─global::foo()
#>  2. │ └─global::bar()
#>  3. │   └─1 %>% identity() %>% baz()
#>  4. └─global::baz(.)
```

Note that one consequence of having a lazy pipe is that the whole pipeline will be shown on the call stack before any errors are thrown:

```{r, eval = FALSE}
f1 <- function(x) x
f2 <- function(x) x
f3 <- function(x) x
f4 <- function(x) x

stop("oh no") %>% f1() %>% f2() %>% f3() %>% f4()
#> Error in f1(.) : oh no

rlang::last_trace()
#> <error/rlang_error>
#> oh no
#> Backtrace:
#>     █
#>  1. ├─stop("oh no") %>% f1() %>% f2() %>% f3() %>% f4()
#>  2. ├─global::f4(.)
#>  3. ├─global::f3(.)
#>  4. ├─global::f2(.)
#>  5. └─global::f1(.)
```

The last function of the pipeline is `f4()`, so that's the first one to be run. It evaluates its argument which is provided by `f3()`, so that's the second function pushed on the stack. And so on until `f1()` needs the result of `stop("oh no")` which causes an error.


## Speed

The pipe is now written in C. This greatly improves the performance. Here is a benchmark for the old R implementation:

```{r, eval = FALSE}
bench::mark(
  `1` = NULL %>% f1(),
  `2` = NULL %>% f1() %>% f2(),
  `3` = NULL %>% f1() %>% f2() %>% f3(),
  `4` = NULL %>% f1() %>% f2() %>% f3() %>% f4(),
)
#>   expression     min  median `itr/sec` mem_alloc `gc/sec` n_itr  n_gc
#>   <bch:expr> <bch:t> <bch:t>     <dbl> <bch:byt>    <dbl> <int> <dbl>
#> 1 1           59.4µs  68.9µs    13648.      280B     59.1  6004    26
#> 2 2           82.6µs 101.6µs     9252.      280B     42.8  3894    18
#> 3 3          106.4µs 124.7µs     7693.      280B     18.8  3690     9
#> 4 4          130.9µs 156.1µs     6173.      280B     18.8  2956     9
```

The new implementation is about 10 times less costly per pipe expression:

```{r, eval = FALSE}
bench::mark(
  `1` = NULL %>% f1(),
  `2` = NULL %>% f1() %>% f2(),
  `3` = NULL %>% f1() %>% f2() %>% f3(),
  `4` = NULL %>% f1() %>% f2() %>% f3() %>% f4(),
)
#>   expression      min   median `itr/sec` mem_alloc `gc/sec` n_itr  n_gc
#>   <bch:expr> <bch:tm> <bch:tm>     <dbl> <bch:byt>    <dbl> <int> <dbl>
#> 1 1            4.47µs   5.95µs   159655.        0B     79.9  9995     5
#> 2 2            5.97µs    8.8µs   109534.        0B     32.9  9997     3
#> 3 3            8.83µs  10.63µs    89902.        0B     27.0  9997     3
#> 4 4           10.99µs  13.18µs    72330.        0B     36.2  9995     5
```

We don't generally except this to have much impact on data analysis code, but it might yield meaningful speed ups if you had a pipe inside a tight loop.


## Towards a release

Though we have changed the behaviour of the pipe, none of the 2600 reverse dependencies of magrittr, purrr, tidyr, and dplyr were broken by the change. To be extra sure, we'd be grateful for any additional testing on real-life scripts with this development version.

If you're interested in the design tradeoffs involved in the creation of a pipe operator in R, see the [tradeoffs](https://magrittr.tidyverse.org/articles/tradeoffs.html) vignette. Any comments about the choices we have made are welcome.
