---
output:
  hugodown::md_document

slug: pak-0-6-0
title: pak 0.6.0
date: 2023-08-28
author: Gábor Csárdi
description: >
    pak installs R packages from various sources.
    This post shows the improvements in system requirements installation
    on Linux, in the just released pak 0.6.0 version.

photo:
  url: https://www.pexels.com/photo/blue-white-orange-and-brown-container-van-163726/
  author: Pixabay

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package]
tags: []
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->
```

```{sh, include = FALSE}
docker build --platform linux/amd64 -t pak/ubuntu -f Dockerfile-ubuntu .
docker kill pak-ubuntu || true
docker rm pak-ubuntu || true
docker run -td --platform linux/amd64 --name pak-ubuntu pak/ubuntu bash
docker start pak-ubuntu
```

```{r, include = FALSE}
knitr::dep_prev()
source("docker.R")
knitr::knit_hooks$set(evaluate = hook_docker)
knitr::opts_chunk$set(cache = TRUE)
```

We're delighted to announce the release of [pak](https://pak.r-lib.org)
0.6.0.
pak helps with the installation of R packages, and many related tasks.

You can install pak from CRAN with:

```{r, eval = FALSE}
install.packages("pak")
```

This blog post focuses on the exciting new improvements in the matching
and installation of system requirements on Linux systems.

You can see a full list of changes in the
[release notes](https://github.com/r-lib/pak/releases/tag/v0.6.0)

## System requirements

Many R packages require the installation of external software, otherwise
they do not work, or even load.
For example, the RPostgres R package requires the PostgreSQL client library,
and by default dynamically links to it on Linux systems.
This means that you (or the administrators of your system) need to install
this library, typically in the form of a system package: `libpq-dev` on
Ubuntu and Debian systems, or `postgresql-server-devel` or `postgresql-devel`
on Red Hat, Fedora, etc. systems.

The good news is that pak helps you with this:

- it looks up the required system packages when installing R packages,
- it checks if the required system packages are installed, and
- it installs them automatically, if you are a superuser, or if you can use
  password-less `sudo` to start a superuser shell.

In addition, pak now also has some functions to query system requirements
and system packages.

## Supported platforms

pak 0.6.0 supports the following Linux systems currently:

- Ubuntu Linux,
- Debian Linux,
- Red Hat Enterprise Linux,
- SUSE Linux Enterprise,
- OpenSUSE,
- CentOS,
- Rocky Linux,
- Fedora Linux.

Call `pak::sysreqs_platforms()` to query the current list of supported
platforms:

```{r sysreqs-platforms}
pak::sysreqs_platforms()[,1:3]
```

Call `pak::system_r_platform()` to check if pak has detected your platform
correctly, and `pak::sysreqs_is_supported()` to see if it is supported:

```{r system-r-platform}
pak::system_r_platform()
```

```{r sysreqs-is-supported}
pak::sysreqs_is_supported()
```

## R package installation

If you are using pak as the `root` user, on a supported platform, then
during package installation pak will look up the required system packages,
and will install the missing ones. Here is an example:

```{r pkg-inst}
pak::pkg_install("RPostgres")
```

### Running R as a regular user

If you don't want to use R as the superuser, but you can set up `sudo`
without a password, that works as well.
pak will detect the password-less `sudo` capability, and use it to install
system packages, as needed.

If you run R as a regular (not root) user, and password-less `sudo` is
not available, then pak will print the system requirements, but it will
not try to install or update them.

If you are compiling R packages from source, and they need to link to system
libraries, then their installation will probably fail, until you install
these system packages.

If you are installing binary R packages (e.g. from
[P3M](https://packagemanager.posit.co/client/#/)), then the installation
typically succeeds, but you won't be able to load these packages into R,
until you install the required system packages.

To demonstrate this, let's remove the system package for the PostgreSQL
client library:

```{r pkg-inst-user-setup, results = "hide"}
system("apt-get remove -y libpq5")
```

If now we (re)install the binary RPostgres R package, the installation
will succeed, but then `library()` fails because of the missing system
package. (We will fix the broken R package below.)

```{r pkg-install-user, error = TRUE, echo = 2}
options(pkg.sysreqs = FALSE)
pak::pkg_install("RPostgres?reinstall")
```

```{r pkg-install-user-load}
library(RPostgres)
```

## System requirements queries

pak 0.6.0 also has a number of functions to query system requirements and
system packages.
The `pak::pkg_sysreqs()` function is similar to `pak::pkg_deps()` but in addition to
looking up package dependencies, it also looks up system dependencies, and
only reports the latter:

```{r pkg-sysreqs}
pak::pkg_sysreqs(c("curl", "r-lib/xml2", "devtools", "CHRONOS"))
```

See the manual of `pak::pkg_sysreqs()` to learn how to programmatically extract
information from its return value.

`pak::sysreqs_check_installed()` is a handy function that checks if all
system requirements are installed for some or all R packages in your
library.
This should report our broken RPostgres package:

```{r check-installed}
pak::sysreqs_check_installed()
```

`pak::sysreqs_fix_installed()` goes one step further and also tries to install
the missing system requirements:

```{r fix-installed}
pak::sysreqs_fix_installed()
```

Now we can load RPostgres again:

```{r fix-installed-2}
library(RPostgres)
```

## Configuration

There are several pak configuration options you can use to adjust how
system requirements are handled.
See the complete list in the `config?pak` manual page.

## Other related pak functions

- `pak::sysreqs_db_list()`, `pak::sysreqs_dbmatch()` and `pak::sysreqs_db_update()`
  list, query and update the built-in system requirements database.
- `pak::sysreqs_list_system_packages()` lists system packages, including virtual
  packages and the features they provide.

## More information

- [pak documentation](https://pak.r-lib.org/)
- [System requirements manual page](https://pak.r-lib.org/dev/reference/sysreqs.html)
- [System requirements database](https://github.com/rstudio/r-system-requirements)

## Acknowledgements

A big thank you to all those who have contributed to pak, or one of its
workhorse packages since the v0.5.1 release:

[&#x0040;alexpate30](https://github.com/alexpate30),
[&#x0040;averissimo](https://github.com/averissimo),
[&#x0040;ArnaudKunzi](https://github.com/ArnaudKunzi),
[&#x0040;billdenney](https://github.com/billdenney),
[&#x0040;Darxor](https://github.com/Darxor),
[&#x0040;drmowinckels](https://github.com/drmowinckels),
[&#x0040;Fan-iX](https://github.com/Fan-iX),
[&#x0040;gongyh](https://github.com/gongyh),
[&#x0040;hadley](https://github.com/hadley),
[&#x0040;idavydov](https://github.com/idavydov),
[&#x0040;jefferis](https://github.com/jefferis),
[&#x0040;joan-yanqiong](https://github.com/joan-yanqiong),
[&#x0040;kevinushey](https://github.com/kevinushey),
[&#x0040;kkmann](https://github.com/kkmann),
[&#x0040;klmr](https://github.com/klmr),
[&#x0040;krlmlr](https://github.com/krlmlr),
[&#x0040;lgaborini](https://github.com/lgaborini),
[&#x0040;maelle](https://github.com/maelle),
[&#x0040;maxheld83](https://github.com/maxheld83),
[&#x0040;maximsmol](https://github.com/maximsmol),
[&#x0040;michaelmayer2](https://github.com/michaelmayer2),
[&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel),
[&#x0040;olivroy](https://github.com/olivroy),
[&#x0040;pascalgulikers](https://github.com/pascalgulikers),
[&#x0040;pawelru](https://github.com/pawelru),
[&#x0040;royfrancis](https://github.com/royfrancis),
[&#x0040;tanho63](https://github.com/tanho63),
[&#x0040;thomasyu888](https://github.com/thomasyu888),
[&#x0040;vincent-hanlon](https://github.com/vincent-hanlon), and
[&#x0040;VincentGuyader](https://github.com/VincentGuyader).
