---
output: hugodown::hugo_document

slug: bigrquery-1-4-0
title: bigrquery 1.4.0
date: 2021-08-04
author: Jenny Bryan
description: >
    bigrquery 1.4.0 fixes a bug in `bq_table_download()`.

photo:
  url: https://unsplash.com/photos/uzw4MvfG5ps
  author: Henry & Co.

categories: [package] 
tags: [bigrquery, gargle, dbplyr, databases]
---

We're gratified to announce the release of [bigrquery](https://bigrquery.r-dbi.org) 1.4.0.
bigrquery makes it easy to work with data stored in [Google BigQuery](https://developers.google.com/bigquery/), a hosted database for big data.

You can install bigrquery from CRAN with:

```{r, eval = FALSE}
install.packages("bigrquery")
```

This blog post will ...

You can see a full list of changes in the [release notes]({ github_release })

```{r setup}
library(bigrquery)
```

## Topic 1

## Topic 2

## Acknowledgements

We'd like to thank everyone who has furthered the development of bigrquery, since the last major release (v1.0.0), through their contributions in issues and pull requests:

[&#x0040;414theodore](https://github.com/414theodore), [&#x0040;abalter](https://github.com/abalter), [&#x0040;acvelozo](https://github.com/acvelozo), [&#x0040;adhi-r](https://github.com/adhi-r), [&#x0040;afalcioni](https://github.com/afalcioni), [&#x0040;ahmohamed](https://github.com/ahmohamed), [&#x0040;ajhindle](https://github.com/ajhindle), [&#x0040;AlekseyBuzmakov](https://github.com/AlekseyBuzmakov), [&#x0040;analyse9823](https://github.com/analyse9823), [&#x0040;andirey](https://github.com/andirey), [&#x0040;ArbenKqiku](https://github.com/ArbenKqiku), [&#x0040;arvhug](https://github.com/arvhug), [&#x0040;batpigandme](https://github.com/batpigandme), [&#x0040;bbrewington](https://github.com/bbrewington), [&#x0040;bhargavimoorthyrao](https://github.com/bhargavimoorthyrao), [&#x0040;btrx-sreddy](https://github.com/btrx-sreddy), [&#x0040;byapparov](https://github.com/byapparov), [&#x0040;carbocation](https://github.com/carbocation), [&#x0040;CartWill](https://github.com/CartWill), [&#x0040;chrisherold](https://github.com/chrisherold), [&#x0040;ChrisJohnsonUMG](https://github.com/ChrisJohnsonUMG), [&#x0040;cpcgoogle](https://github.com/cpcgoogle), [&#x0040;danny-molamola](https://github.com/danny-molamola), [&#x0040;deflaux](https://github.com/deflaux), [&#x0040;dmoimpact](https://github.com/dmoimpact), [&#x0040;downloaderfan](https://github.com/downloaderfan), [&#x0040;dsolito](https://github.com/dsolito), [&#x0040;dujm](https://github.com/dujm), [&#x0040;eamcvey](https://github.com/eamcvey), [&#x0040;eddelbuettel](https://github.com/eddelbuettel), [&#x0040;edgararuiz-zz](https://github.com/edgararuiz-zz), [&#x0040;eduardodrc](https://github.com/eduardodrc), [&#x0040;evandropp10](https://github.com/evandropp10), [&#x0040;everron](https://github.com/everron), [&#x0040;geotheory](https://github.com/geotheory), [&#x0040;gikis1](https://github.com/gikis1), [&#x0040;gjuggler](https://github.com/gjuggler), [&#x0040;gkmuralimech](https://github.com/gkmuralimech), [&#x0040;grantmcdermott](https://github.com/grantmcdermott), [&#x0040;guillaumed90](https://github.com/guillaumed90), [&#x0040;hadley](https://github.com/hadley), [&#x0040;HarlanH](https://github.com/HarlanH), [&#x0040;hlynurhallgrims](https://github.com/hlynurhallgrims), [&#x0040;htappen](https://github.com/htappen), [&#x0040;Iuiu1234](https://github.com/Iuiu1234), [&#x0040;izzetagoren](https://github.com/izzetagoren), [&#x0040;j450h1](https://github.com/j450h1), [&#x0040;janejuenyang](https://github.com/janejuenyang), [&#x0040;jayBana](https://github.com/jayBana), [&#x0040;jberninger](https://github.com/jberninger), [&#x0040;jcheng5](https://github.com/jcheng5), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;jimmyg3g](https://github.com/jimmyg3g), [&#x0040;joetortorelli](https://github.com/joetortorelli), [&#x0040;jordanwebb10](https://github.com/jordanwebb10), [&#x0040;jpryda](https://github.com/jpryda), [&#x0040;jrecasens](https://github.com/jrecasens), [&#x0040;Ka2wei](https://github.com/Ka2wei), [&#x0040;KarimZaoui](https://github.com/KarimZaoui), [&#x0040;kesnalawrence](https://github.com/kesnalawrence), [&#x0040;kevinwang09](https://github.com/kevinwang09), [&#x0040;kkmann](https://github.com/kkmann), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;Kvit](https://github.com/Kvit), [&#x0040;ldanai](https://github.com/ldanai), [&#x0040;leemc-data-ed](https://github.com/leemc-data-ed), [&#x0040;LukasWallrich](https://github.com/LukasWallrich), [&#x0040;maelle](https://github.com/maelle), [&#x0040;mapinas](https://github.com/mapinas), [&#x0040;mauricioita](https://github.com/mauricioita), [&#x0040;meystingray](https://github.com/meystingray), [&#x0040;meztez](https://github.com/meztez), [&#x0040;mr2dark](https://github.com/mr2dark), [&#x0040;mwilson19](https://github.com/mwilson19), [&#x0040;paleolimbot](https://github.com/paleolimbot), [&#x0040;paulsendavidjay](https://github.com/paulsendavidjay), [&#x0040;philbrierley](https://github.com/philbrierley), [&#x0040;ras44](https://github.com/ras44), [&#x0040;rasmusab](https://github.com/rasmusab), [&#x0040;reliscu](https://github.com/reliscu), [&#x0040;riccardopinosio](https://github.com/riccardopinosio), [&#x0040;Saikri5hna](https://github.com/Saikri5hna), [&#x0040;samudzi](https://github.com/samudzi), [&#x0040;santic113](https://github.com/santic113), [&#x0040;saptarshiguha](https://github.com/saptarshiguha), [&#x0040;Schumzy](https://github.com/Schumzy), [&#x0040;SeagleLiu](https://github.com/SeagleLiu), [&#x0040;selcukakbas](https://github.com/selcukakbas), [&#x0040;selesnow](https://github.com/selesnow), [&#x0040;siroros](https://github.com/siroros), [&#x0040;skydavis435](https://github.com/skydavis435), [&#x0040;spgarbet](https://github.com/spgarbet), [&#x0040;spiddy69](https://github.com/spiddy69), [&#x0040;srkpratap](https://github.com/srkpratap), [&#x0040;stelsemeyer](https://github.com/stelsemeyer), [&#x0040;stevecondylios](https://github.com/stevecondylios), [&#x0040;svmakarovv](https://github.com/svmakarovv), [&#x0040;tchaithonov](https://github.com/tchaithonov), [&#x0040;tdsmith](https://github.com/tdsmith), [&#x0040;theclue](https://github.com/theclue), [&#x0040;tinoater](https://github.com/tinoater), [&#x0040;valentas-kurauskas](https://github.com/valentas-kurauskas), [&#x0040;valentinumbach](https://github.com/valentinumbach), [&#x0040;victorz-ca](https://github.com/victorz-ca), [&#x0040;warnes](https://github.com/warnes), [&#x0040;YuanyuanZhang1986](https://github.com/YuanyuanZhang1986), [&#x0040;zacdav](https://github.com/zacdav), [&#x0040;ZainRizvi](https://github.com/ZainRizvi), [&#x0040;zerobytes](https://github.com/zerobytes), and [&#x0040;zoews](https://github.com/zoews).
