---
title: lubridate 1.7.0
slug: lubridate-1.7.0
description: >
  A new version of lubridate is on CRAN!
author: Vitalie Spinu
date: 2017-11-01
photo:
  url: https://unsplash.com/photos/yBzrPGLjMQw
  author: Heather Zabriskie
categories: [package]
---

```{r setup, include = FALSE}
library("lubridate")
knitr::opts_chunk$set(collapse = TRUE, comment = "##=>")
```

We are pleased to announce that [lubridate
1.7.0](https://cran.r-project.org/web/packages/lubridate/index.html) is now on
CRAN! For a complete set of changes please see the release
[NEWS](https://github.com/tidyverse/lubridate/releases/tag/v1.7.0).

Lubridate is a package that makes working with date-time and time-span objects
easier. It provides fast and user friendly parsing of date-time strings,
extraction and updating of components of a date-time objects (years, months,
days etc.) and algebraic manipulation on date-time and time-span objects.

Here is a brief walk-through of the prominent new features in 1.7.0.

## Built-in CCTZ and much faster `update` and `force_tz`

From this version lubridate relies on google's
[CCTZ](https://github.com/google/cctz) library for the date-time updates and
time-zone manipulation. This means that `force_tz`, `update`, `round`ing and a
range of arithmetic operations on timespans are now considerably faster.

## Vectorized time-zone manipulation

Often date-time data comes with heterogeneous time-zones. For example, you might
have a character vector of date-times and a separate vector of time-zones. In
such cases you can parse your date-time vector in one time zone (`UTC`) and then
"enforce" the heterogeneous time-zones with `force_tzs`.


```{r}
x <- ymd_hms(c("2009-08-01 00:00:00", "2009-08-01 00:00:00"))
tzs <- c("America/New_York", "Europe/Amsterdam")

force_tzs(x, tzones = tzs)

force_tzs(x, tzones = tzs, tzone_out = "America/New_York")
```

Note that first `force_tzs` call produced a vector of instants in UTC time-zone,
for the second call we specified the desired time-zone of the output vector with
the `tzone_out` argument. This is needed as R's date-time vectors cannot
represent heterogeneous time-zones. For the same reason, the counterpart of
`with_tz`, `with_tzs`, does not exist. Instead, `local_time` should cover the
`with_tzs` use case in most situations.

New `local_time` function retrieves day clock time of the input vector in the
specified time zones. It returns a `difftime` object and it is vectorized over
both date-time and time-zone arguments:

```{r}
x <- ymd_hms(c("2009-08-01 01:02:03", "2009-08-01 10:20:30"))
local_time(x, units = "hours")

x <- ymd_hms(c("2009-08-01 00:00:00", "2009-08-01 00:00:00"))
tzs <- c("America/New_York", "Europe/Amsterdam")
local_time(x, tzs, units = "hours")
```

## New `cutoff_2000` parameter in parsing functions

Lower level lubridate parser functions (`parse_datetime2`,`fast_strptime`) now
accept `cutoff_2000` parameter to determine when the parsing of `yy` format
should output 19th or 20th century.

```{r}
x <- c("50-01-01", "70-01-01")
parse_date_time2(x, "ymd")
parse_date_time2(x, "ymd", cutoff_2000 = 30)
```

By default `cutoff_2000` is 68 to comply with R's `strptime` function.

This feature was not propagated to higher level parsing function because base
`strptime`, on which those occasionally rely, doesn't support this option.


## Functions `wday` and `month` are now localized

When `label=TRUE` `wday` and `month` extractors return localized strings:

```{r}
Sys.setlocale(locale = "zh_CN.utf8")
wday(now(), label = TRUE)
month(now(), label = TRUE)
```

Please note that for this reason using labels for comparison is not
recommended. You should use numeric output instead. Otherwise your code might
not work on other computers.

## Several functions gained the `week_start` parameter

Functions for which depend on the week-start conventions (`wday`, `wday<-`,
`floor_date`, `ceiling_date` and `round_date`) now accept `week_start` argument
which defaults to `getOption("lubridate.week.start", 7)`.

```{r}
x <- today()
wday(x, label = TRUE, abbr = FALSE)
wday(x)
wday(x,  week_start = 2)
```

## Rounding to fraction of a second

R's date-time format (POSIXct) is accurate up to microseconds, thus it makes
sense to round to fractions of a second. Unfortunately R currently prints
fractional seconds incorrectly which can lead to confusion:

```{r}
## print fractional seconds
options(digits.secs=6, digits = 13)

x <- ymd_hms("2009-08-03 12:01:59.031")
rx <- round_date(x, ".01sec")
cx <- ceiling_date(x, ".01sec")

c(rx, cx)
as.double(c(rx, cx))
```

So, fractional rounding and ceiling work, but don't be taken aback by incorrect printing;)

## New `epiweek` and `epiyear` functions

New functions `epiyear` and `epiweek` are the US CDC version of epidemiological
weeks and years. They follow same rules as `isoweek` and `isoyear` but with the
week starting on Sunday. In other parts of the world the convention is to start
epidemiological weeks on Monday, which is then the same as for `isoweek` and
`isoyear`.
