---
output: hugodown::hugo_document

slug: roxygen2-7-2-0
title: roxygen2 7.2.0
date: 2022-04-25
author: Hadley Wickham
description: >
    roxygen2 7.2.0 brings improvements to `NAMESPACE` generation, 
    better multiparameter argument inheritance, and improved warnings.
photo:
  url: https://unsplash.com/photos/hpTH5b6mo2s
  author: ian dooley

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [roxygen2, devtools]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->
```
We're tickled pink to announce the release of [roxygen2](https://roxygen2.r-lib.org) 7.2.0.
roxygen2 allows you to write specially formatted R comments that generate R documentation files (`man/*.Rd`) and the `NAMESPACE` file.
roxygen2 is used by over 9,000 CRAN packages.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("roxygen2")
```

There are five big improvements in this release:

-   The `NAMESPACE` roclet now preserves all existing non-import directives during its first pass.
    This will generally eliminate the pair of `"NAMESPACE has changed"` messages and should reduce the chances that you end up with a sufficiently broken `NAMESPACE` that you can't re-load and re-document your package.

-   `@inheritParams` now only inherits exact multi-parameter matches.
    For example take `my_plot()` below:

    ```{r}
    #' @param width,height The dimensions in inches
    my_plot <- function(x, width, height) {
      
    }
    ```

    Previously, `width` and `height` were inherited individually, so this roxygen2 block:

    ```{r}
    #' @inheritParams my_plot
    your_plot <- function(x, y, width, height) {
      
    } 
    ```

    Would be equivalent to:

    ```{r}
    #' @param width The dimensions in inches
    #' @param height The dimensions in inches
    your_plot <- function(x, y, width, height) {
      
    } 
    ```

    Now, multiparmeter arguments will only be inherited as a whole.
    This could break your documentation if you were relying on this behaviour, but in all the cases we've seen in tidyverse packages it's been an improvement.

-   We've done a thorough review of all warning messages to make them more informative and actionable.
    We've also fixed a number of bugs that lead to invalid Rd files or pointed you to something that wasn't actually the problem.

    If you have a daily build of RStudio, warnings will now include a clickable hyperlink that takes you directly to the problem.
    This technology is under active development across the IDE and the cli package but is something you should expect to see more of in the future.

You can see a full list of changes in the [release notes](%7B%20github_release%20%7D)

## Acknowledgements

A big thanks to everyone who contributed to this release through their issues, pull requests, and discussions! [&#x0040;AlexisDerumigny](https://github.com/AlexisDerumigny), [&#x0040;BenWiseman](https://github.com/BenWiseman), [&#x0040;billdenney](https://github.com/billdenney), [&#x0040;bobjansen](https://github.com/bobjansen), [&#x0040;brry](https://github.com/brry), [&#x0040;cderv](https://github.com/cderv), [&#x0040;cjyetman](https://github.com/cjyetman), [&#x0040;courtiol](https://github.com/courtiol), [&#x0040;DanChaltiel](https://github.com/DanChaltiel), [&#x0040;danielvartan](https://github.com/danielvartan), [&#x0040;DarioS](https://github.com/DarioS), [&#x0040;DavisVaughan](https://github.com/DavisVaughan), [&#x0040;dieghernan](https://github.com/dieghernan), [&#x0040;dmurdoch](https://github.com/dmurdoch), [&#x0040;dwachsmuth](https://github.com/dwachsmuth), [&#x0040;flrd](https://github.com/flrd), [&#x0040;gaborcsardi](https://github.com/gaborcsardi), [&#x0040;hadley](https://github.com/hadley), [&#x0040;IndrajeetPatil](https://github.com/IndrajeetPatil), [&#x0040;JantekM](https://github.com/JantekM), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;karoliskoncevicius](https://github.com/karoliskoncevicius), [&#x0040;kongdd](https://github.com/kongdd), [&#x0040;kpagacz](https://github.com/kpagacz), [&#x0040;lionel-](https://github.com/lionel-), [&#x0040;lorenzwalthert](https://github.com/lorenzwalthert), [&#x0040;maelle](https://github.com/maelle), [&#x0040;malcolmbarrett](https://github.com/malcolmbarrett), [&#x0040;mbojan](https://github.com/mbojan), [&#x0040;MichaelChirico](https://github.com/MichaelChirico), [&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [&#x0040;MislavSag](https://github.com/MislavSag), [&#x0040;mschilli87](https://github.com/mschilli87), [&#x0040;Nelson-Gon](https://github.com/Nelson-Gon), [&#x0040;netique](https://github.com/netique), [&#x0040;pnacht](https://github.com/pnacht), [&#x0040;ramiromagno](https://github.com/ramiromagno), [&#x0040;romainfrancois](https://github.com/romainfrancois), [&#x0040;saicharanp18](https://github.com/saicharanp18), [&#x0040;simonsays1980](https://github.com/simonsays1980), [&#x0040;ThierryO](https://github.com/ThierryO), [&#x0040;wch](https://github.com/wch), [&#x0040;wurli](https://github.com/wurli), and [&#x0040;yogat3ch](https://github.com/yogat3ch).
