---
output: hugodown::hugo_document

slug: roxygen2-7-2-0
title: roxygen2 7.2.0
date: 2022-05-13
author: Hadley Wickham
description: >
    roxygen2 7.2.0 brings improvements to `NAMESPACE` generation, 
    better multiparameter argument inheritance, and improved warnings.
photo:
  url: https://unsplash.com/photos/hpTH5b6mo2s
  author: ian dooley

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [roxygen2, devtools]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->
```
We're tickled pink to announce the release of [roxygen2](https://roxygen2.r-lib.org) 7.2.0.
roxygen2 allows you to write specially formatted R comments that generate R documentation files (`man/*.Rd`) and the `NAMESPACE` file.
roxygen2 is used by over 9,000 CRAN packages.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("roxygen2")
```

There are three big improvements in this release:

-   The `NAMESPACE` roclet now preserves all existing non-import directives during its first pass.
    This will generally eliminate the pair of `"NAMESPACE has changed"` messages and should reduce the chances that you end up with a sufficiently broken `NAMESPACE` that you can't re-load and re-document your package.

-   `@inheritParams` now only inherits exact multi-parameter matches.
    For example take `my_plot()` below:

    ```{r}
    #' @param width,height The dimensions in inches
    my_plot <- function(x, width, height) {
      
    }
    ```

    Previously, `width` and `height` were inherited individually, so this roxygen2 block:

    ```{r}
    #' @inheritParams my_plot
    your_plot <- function(x, y, width, height) {
      
    } 
    ```

    Would be equivalent to:

    ```{r}
    #' @param width The dimensions in inches
    #' @param height The dimensions in inches
    your_plot <- function(x, y, width, height) {
      
    } 
    ```

    Now, multi-parameter arguments will be inherited as a whole.
    This could potentially break your documentation if you (e.g.) only had one of `width` and `height` in your function.
    But we've only seen this problem a few places in the tidyverse, it was easily fixed, and inherited arguments are generally much improved.

-   We've done a thorough review of all warning messages to make them more informative and actionable.
    We've also fixed a number of bugs that led to invalid Rd files or pointed you to the wrong place.

    If you have a daily build of RStudio, warnings now include a clickable link that takes you directly to the problem.
    This technology is under active development across the IDE and the [cli](https://cli.r-lib.org) package and you can expect to see more of it in the future.

You can see a full list of changes in the [release notes](https://github.com/r-lib/roxygen2/blob/main/NEWS.md).

## Acknowledgements

A big thanks to everyone who contributed to this release through their issues, pull requests, and discussions!
[\@AlexisDerumigny](https://github.com/AlexisDerumigny), [\@BenWiseman](https://github.com/BenWiseman), [\@billdenney](https://github.com/billdenney), [\@bobjansen](https://github.com/bobjansen), [\@brry](https://github.com/brry), [\@cderv](https://github.com/cderv), [\@cjyetman](https://github.com/cjyetman), [\@courtiol](https://github.com/courtiol), [\@DanChaltiel](https://github.com/DanChaltiel), [\@danielvartan](https://github.com/danielvartan), [\@DarioS](https://github.com/DarioS), [\@DavisVaughan](https://github.com/DavisVaughan), [\@dieghernan](https://github.com/dieghernan), [\@dmurdoch](https://github.com/dmurdoch), [\@dwachsmuth](https://github.com/dwachsmuth), [\@flrd](https://github.com/flrd), [\@gaborcsardi](https://github.com/gaborcsardi), [\@hadley](https://github.com/hadley), [\@IndrajeetPatil](https://github.com/IndrajeetPatil), [\@JantekM](https://github.com/JantekM), [\@jennybc](https://github.com/jennybc), [\@karoliskoncevicius](https://github.com/karoliskoncevicius), [\@kongdd](https://github.com/kongdd), [\@kpagacz](https://github.com/kpagacz), [\@lionel-](https://github.com/lionel-), [\@lorenzwalthert](https://github.com/lorenzwalthert), [\@maelle](https://github.com/maelle), [\@malcolmbarrett](https://github.com/malcolmbarrett), [\@mbojan](https://github.com/mbojan), [\@MichaelChirico](https://github.com/MichaelChirico), [\@mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [\@MislavSag](https://github.com/MislavSag), [\@mschilli87](https://github.com/mschilli87), [\@Nelson-Gon](https://github.com/Nelson-Gon), [\@netique](https://github.com/netique), [\@pnacht](https://github.com/pnacht), [\@ramiromagno](https://github.com/ramiromagno), [\@romainfrancois](https://github.com/romainfrancois), [\@saicharanp18](https://github.com/saicharanp18), [\@simonsays1980](https://github.com/simonsays1980), [\@ThierryO](https://github.com/ThierryO), [\@wch](https://github.com/wch), [\@wurli](https://github.com/wurli), and [\@yogat3ch](https://github.com/yogat3ch).
