---
output: hugodown::hugo_document

slug: dtplyr-1-2-0
title: dtplyr 1.2.0
date: 2021-11-28
author: Hadley Wickham
description: >
    dtplyr 1.2.0 adds two new authors, a bunch of tidyr translations,
    new join translations, and many minor translation improvements.
    
photo:
  url: https://unsplash.com/photos/uRQlCmfOCRg
  author: Zdeněk Macháček

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [dplyr]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're thrilled to announce that [dtplyr](https://dtplyr.tidyverse.org) 1.0.0 is now on CRAN. dtplyr gives you the speed of [data.table](http://r-datatable.com/) with the syntax of dplyr; you write dplyr (and tidyr) code and dtplyr translates it to the data.table equivalent.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("dtplyr")
```

I'll discuss three major changes in this blog post:

* New authors
* New tidyr translations
* Improvements to join translations

There are also over 20 minor improvements to the quality of translations. You can see a full list of changes in the [release notes](https://github.com/tidyverse/dtplyr/blob/main/NEWS.md)

```{r setup}
library(dtplyr)

library(dplyr, warn.conflicts = FALSE)
library(tidyr)
```

## New authors

The biggest news in this release is the addition of two new [authors](https://github.com/tidyverse/tidyups/blob/main/004-governance.md#authors): [Mark Fairbanks](https://github.com/markfairbanks) and [Maximilian Girlich](https://github.com/mgirlich) are now dtplyr authors in recognition of their significant and sustained contributions. Along with @eutwt, they implemented the bulk of the improvements in this release!

## tidyr translations

Thanks to Mark and Maximilian, dtplyr gains translations for many more tidyr verbs including `complete()`, `drop_na()`, `expand()`, `fill()`, `nest()`, `pivot_longer()`, `replace_na()` and `separate()`. A few examples are shown below:

```{r}
dt <- lazy_dt(data.frame(x = c(NA, "x.y", "x.z", "y.z")))
dt %>% 
  separate(x, c("A", "B"), sep = "\\.", remove = FALSE) %>% 
  show_query()

dt <- lazy_dt(data.frame(x = c(1, NA, NA, 2, NA)))
dt %>% 
  fill(x) %>% 
  show_query()

dt %>% 
  replace_na(list(x = 99)) %>% 
  show_query()

dt <- lazy_dt(relig_income)
dt %>%
  pivot_longer(!religion, names_to = "income", values_to = "count") %>% 
  show_query()
```

## Improvements to joins

The join functions have been overhauled: `inner_join()`, `left_join()`, and `right_join()` now all produce a call to `[`, rather than to `merge()`:

```{r}
dt1 <- lazy_dt(data.frame(x = 1:3))
dt2 <- lazy_dt(data.frame(x = 2:3, y = c("a", "b")))

dt1 %>% inner_join(dt2, by = "x") %>% show_query()
dt1 %>% left_join(dt2, by = "x") %>% show_query()
dt2 %>% right_join(dt1, by = "x") %>% show_query()
```

This can make the translation a little longer for simple joins, but simplifies the underlying code base. This simplification has made it easier to more closely match dplyr behaviour for column order, handling of named `by`, Cartesian joins (`by = character()`), and dealing with duplicated variables names.

## Acknowledgements

As always, tidyverse packages wouldn't be possible with the community, so a big thanks goes out to all 35 folks who helped to make this release a reality: [&#x0040;akr-source](https://github.com/akr-source), [&#x0040;batpigandme](https://github.com/batpigandme), [&#x0040;bguillod](https://github.com/bguillod), [&#x0040;cgoo4](https://github.com/cgoo4), [&#x0040;chenx2018](https://github.com/chenx2018), [&#x0040;D-Se](https://github.com/D-Se), [&#x0040;eutwt](https://github.com/eutwt), [&#x0040;hadley](https://github.com/hadley), [&#x0040;jatherrien](https://github.com/jatherrien), [&#x0040;jdmoralva](https://github.com/jdmoralva), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;jtlandis](https://github.com/jtlandis), [&#x0040;kmishra9](https://github.com/kmishra9), [&#x0040;lutzgruber](https://github.com/lutzgruber), [&#x0040;lutzgruber-quantco](https://github.com/lutzgruber-quantco), [&#x0040;markfairbanks](https://github.com/markfairbanks), [&#x0040;mgirlich](https://github.com/mgirlich), [&#x0040;mrcaseb](https://github.com/mrcaseb), [&#x0040;nassuphis](https://github.com/nassuphis), [&#x0040;nigeljmckernan](https://github.com/nigeljmckernan), [&#x0040;NZambranoc](https://github.com/NZambranoc), [&#x0040;PMassicotte](https://github.com/PMassicotte), [&#x0040;psads-git](https://github.com/psads-git), [&#x0040;quid-agis](https://github.com/quid-agis), [&#x0040;romainfrancois](https://github.com/romainfrancois), [&#x0040;roni-fultheim](https://github.com/roni-fultheim), [&#x0040;samlipworth](https://github.com/samlipworth), [&#x0040;sanjmeh](https://github.com/sanjmeh), [&#x0040;sbashevkin](https://github.com/sbashevkin), [&#x0040;StatsGary](https://github.com/StatsGary), [&#x0040;torema-ed](https://github.com/torema-ed), [&#x0040;verajosemanuel](https://github.com/verajosemanuel), [&#x0040;Waldi73](https://github.com/Waldi73), [&#x0040;wurli](https://github.com/wurli), and [&#x0040;yiugn](https://github.com/yiugn).
