---
output: hugodown::hugo_document

slug: bigrquery-1-5-0
title: bigrquery 1.5.0
date: 2024-01-22
author: Hadley Wickham
description: >
    This release fixes a bunch of annoyances and catches up with 
    innovations in DBI and dbplyr.

photo:
  url: https://unsplash.com/photos/gr06IVY2YpM
  author: Aaron Santelices

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [bigrquery]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->
```
We're stoked to announce the release of [bigrquery](http://bigrquery.r-dbi.org/) 1.4.2.
bigrquery makes it easy to work with data stored in [Google BigQuery](https://developers.google.com/bigquery/), a hosted database for big data.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("bigquery")
```

This has been the first major update to bigrquery for a while, and is mostly about catching up with innovations elsewhere as well as squashing a bunch of smaller annoyances.

```{r}
library(bigrquery)
```

Here's a summary of the biggest changes:

-   bigrquery is now [MIT licensed](https://www.tidyverse.org/blog/2021/12/relicensing-packages/).

-   Deprecated functions (i.e. those not starting with `bq_`) have been removed.
    These have been superseded for a long time and were formally deprecated in bigrquery 1.3.0 (2020).

-   `bq_table_download()` now returns unknown fields as character vectors.
    In particular, this means that `BIGNUMERIC` and `JSON` columns are downloaded into R for you to process as you wish.
    `bq_table_download()` now uses the [clock package](https://clock.r-lib.org) to parse dates, leading to a considerable performance improvement and correct parsing for dates prior to 1970-01-01.

-   bigquery datasets and tables will now appear in the [RStudio connections pane](https://docs.posit.co/ide/user/ide/guide/data/data-connections.html) when connecting with `DBI::dbConnect()`.

-   `DBI::dbAppendTable(),` `DBI::dbCreateTable()`, and `DBI::dbExecute()` are now supported, and `DBI::dbGetQuery()`/`DBI::dbSendQuery()` support parameterised queries via the `params` argument.
    `DBI::dbReadTable()`, `DBI::dbWriteTable()`, `DBI::dbExistsTable()`, `DBI::dbRemoveTable()`, and `DBI::dbListFields()` now all work with `DBI::Id()`.

-   bigrquery now uses 2nd edition of dbplyr interface and is compatible with dbplyr 2.4.0.

See the [release notes](https://github.com/r-dbi/bigrquery/releases/tag/v1.5.0) for a full list of changes.

## Acknowledgements

A big thanks to all 14 folks who helped make this release happen with questions, comments, and code: [\@abalter](https://github.com/abalter), [\@ablack3](https://github.com/ablack3), [\@evanrollinsdrumline](https://github.com/evanrollinsdrumline), [\@hadley](https://github.com/hadley), [\@husseyd](https://github.com/husseyd), [\@jacobmpeters](https://github.com/jacobmpeters), [\@jennybc](https://github.com/jennybc), [\@Kvit](https://github.com/Kvit), [\@meztez](https://github.com/meztez), [\@mgirlich](https://github.com/mgirlich), [\@MichaelChirico](https://github.com/MichaelChirico), [\@mjbroerman](https://github.com/mjbroerman), [\@ncuriale](https://github.com/ncuriale), and [\@rdavis120](https://github.com/rdavis120).
