---
output: hugodown::hugo_document

slug: dplyr-1-0-4-if-any
title: dplyr 1.0.4 if_any() and if_all()
date: 2021-02-01
author: Romain Francois

photo:
  url: https://unsplash.com/photos/nQz49efZEFs
  author: Mattias Olsson

categories: [package] 
tags: [dplyr]
---

We're happy to announce the release of [dplyr](https://dplyr.tidyverse.org) 1.0.4. dplyr is the data
manipulation package of the [tidyverse](https://www.tidyverse.org). 

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("dplyr")
```

This blog post will discuss the two most important changes of the 1.0.4 release: 
the new `if_any()` and `if_all()` functions, and performance improvements for `across()`.

You can see a full list of changes in the [release notes]({ github_release })

```{r setup}
library(dplyr)
```

## if_any() and if_all()

## Faster across()

## Acknowledgements

[&#x0040;abalter](https://github.com/abalter), [&#x0040;cuixueqin](https://github.com/cuixueqin), [&#x0040;eggrandio](https://github.com/eggrandio), [&#x0040;everetr](https://github.com/everetr), [&#x0040;hadley](https://github.com/hadley), [&#x0040;hjohns12](https://github.com/hjohns12), [&#x0040;iago-pssjd](https://github.com/iago-pssjd), [&#x0040;jahonamir](https://github.com/jahonamir), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;lionel-](https://github.com/lionel-), [&#x0040;lotard](https://github.com/lotard), [&#x0040;luispfonseca](https://github.com/luispfonseca), [&#x0040;mbcann01](https://github.com/mbcann01), [&#x0040;mutahiwachira](https://github.com/mutahiwachira), [&#x0040;Robinlovelace](https://github.com/Robinlovelace), [&#x0040;romainfrancois](https://github.com/romainfrancois), [&#x0040;rpruim](https://github.com/rpruim), [&#x0040;shahronak47](https://github.com/shahronak47), [&#x0040;shangguandong1996](https://github.com/shangguandong1996), [&#x0040;sylvainDaubree](https://github.com/sylvainDaubree), [&#x0040;tomazweiss](https://github.com/tomazweiss), [&#x0040;vhorvath](https://github.com/vhorvath), [&#x0040;wasdoff](https://github.com/wasdoff), and [&#x0040;Yunuuuu](https://github.com/Yunuuuu).
