---
output: hugodown::hugo_document

slug: purrr-1-0-0
title: purrr 1.0.0
date: 2022-11-10
author: Hadley Wickham
description: >
    purrr 1.0.0 brings a basket of updates. We deprecated a number of
    seldom used functions to hone in on the core purpose of purrr and 
    implemented a swath of new features including progress bars, improved 
    error reporting, and much much more!

photo:
  url: https://unsplash.com/photos/YCPkW_r_6uA
  author: Jari Hytönen

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [purrr]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->
```
```{r}
#| include: false
asciicast::init_knitr_engine()
```

We're happy to announce the release of [purrr](http://purrr.tidyverse.org/) 1.0.0!
purrr enhances R's functional programming toolkit by providing a complete and consistent set of tools for working with functions and vectors.
You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("purrr")
```

purrr is 7 years old, but it's finally made it to 1.0.0!
This is a big release, adding some long-needed functionality (like progress bars!) as well as really refining the core purpose of purrr.
In this post, we'll start with an overview of the breaking changes, then briefly review some documentation changes.
Then we'll get to the good stuff: improvements to the `map` family, new `keep_at()` and `discard_at()` functions, and improvements to flattening and simplification.
You can see a full list of changes in the [release notes](%7B%20github_release%20%7D).

```{r setup}
library(purrr)
```

## Breaking changes

purrrr 1.0.0 was an opportunity to really refine the core purrrpose of purrr, making it more purrrsimonious by moving a number of functions to purrrgatory.
We've done our best to ensure that changes are not cat-istrophic, which I can now explain, having got these puns out of my system.

This release is more aggressive than usual because a 1.0.0 release is our opportunity to make bigger changes that form a strong foundation for the next 10 years.
These changes will break some existing code, but we've done our best to make it affect as little code as possible.
For example, I [made pull requests](https://github.com/tidyverse/purrr/issues/969) to all \~40 CRAN packages (out of \~1,400) that purrr 1.0.0 broke.
Making these fixes helped give me confidence that, though we're deprecating quite a few functions and changing a few special cases, it shouldn't affect too much code in the wild.

There are four important changes that you should be aware of:

-   `pluck()` behaves differently when extracting 0-length vectors.
-   The map function family uses the tidyverse rules for coercion and recycling.
-   All functions that modify lists handle `NULL` consistently.
-   We've deprecated functions that aren't related to the core purpose of purr.

### pluck and zero-length vectors

Previously, `pluck()` replaced 0-length vectors with the value of `default`.
Now `default` is only used for `NULL`s and absent elements:

```{r}
x <- list(y = list(a = character(), b = NULL))
x |> pluck("y", "a", .default = NA)
x |> pluck("y", "b", .default = NA)
x |> pluck("y", "c", .default = NA)
```

This also influences the map family because using an integer vector, character vector, or list instead of a function automatically calls `pluck()`:

```{r}
x <- list(list(1), list(), list(NULL), list(character()))
x |> map(1, .default = 0) |> str()
```

We made this change because it makes purrr more consistent with the rest of the tidyverse and it looks like it was a bug in the original implementation of the function.

### Tidyverse consistency

We've tweaked the map family of functions to be more consistent with general tidyverse coercion and recycling rules, as implemented by the [vctrs](https://vctrs.r-lib.org) package.
`map_lgl()`, `map_int()`, `map_int()`, and `map_dbl()` now follow the same coercion rules as vctrs.

-   `map_chr(TRUE, identity)`, `map_chr(0L, identity)`, and `map_chr(1L, identity)` have been deprecated because we believe that converting a logical/integer/double to a character vector is potentially dangerous and should require an explicit coercion.

    ```{r}
    # previously you could write
    map_chr(1:4, \(x) x + 1)

    # now you need something like this:
    map_chr(1:4, \(x) as.character(x + 1))
    ```

-   `map_int()` requires that the numeric results be close to integers, rather than silently truncating to integers.
    Compare these two examples:

    ```{r}
    #| error: true
    map_int(1:3, \(x) x / 2)

    map_int(1:3, \(x) x * 2)
    ```

`map2()`, `modify2()`, and `pmap()` use tidyverse recycling rules, which mean that vectors of length 1 are recycled to any size but all other vectors must have the same length.
This has two major changes:

-   Previously, the presence of a zero-length input generated a zero-length output.
    Now its recycled using the same rules:

    ```{r}
    #| error: true
    map2(1:2, character(), paste)

    # Works because length-1 vector gets recycled to length-0
    map2(1, character(), paste)
    ```

-   And now must explicitly recycle vectors that aren't length 1:

    ```{r}
    #| error: true
    map2_int(1:4, c(10, 20), `+`)

    map2_int(1:4, rep(c(10, 20), 2), `+`)
    ```

### Assigning `NULL`

purrr has a number of functions that modify a list: `pluck<-`, `assign_in()`, `modify()`, `modify2()`, `modify_if()`, `modify_at()`, and `list_modify()`.
Previously, these functions had inconsistent behaviour when you attempted to modify an element with `NULL`: some functions would delete that element, and some would set it to `NULL`.
That inconsistency arose because base R handles `NULL` in different ways depending on whether or not use you `$`/`[[` or `[`:

```{r}
x1 <- x2 <- x3 <- list(a = 1, b = 2)

x1$a <- NULL
str(x1)

x2["a"] <- list(NULL)
str(x2)
```

Now all list modifying functions will create an element containing `NULL`:

```{r}
x3 |> 
  list_modify(a = NULL) |> 
  str()

x3 |> 
  modify_at("b", \(x) NULL) |> 
  str()
```

If you want to delete the element, you can use the special `zap()` sentinel:

```{r}
x3 |> 
  list_modify(a = zap()) |> 
  str()
```

`zap()` does not work in `modify*()` because those functions are designed to always return the same top-level structure as the input.

### Core purpose refinements

We have **deprecated** a number of functions to keep purrr focused on its core purpose: facilitating functional programming in R.
Deprecation means that the functions will continue to work, but you'll be warned once every 8 hours if you use them.
In several years time, we'll release an update which causes the warnings to occur on every time you use them, and a few years after that they'll transformed to throwing errors.

-   `cross()` and all its variants have been deprecated because they're slow and buggy, and a better approach already exists in `tidyr::expand_grid()`.

-   `update_list()`, `rerun()`, and the use of tidyselect with `map_at()` and friends have been deprecated because we no longer believe that non-standard evaluation is a good fit for purrr.

-   The `lift_*` family of functions has been deprecated because they promote a style of function manipulation that is not commonly used in R.

-   `prepend()`, `rdunif()`, `rbernoulli()`, `when()`, and `list_along()` have been deprecated because they're not directly related to functional programming.

-   `splice()` has been deprecated because we no longer believe that automatic splicing makes for good UI and there are other ways to achieve the same result.

Consult the documentation for the alternatives that we now recommend.

Deprecating these functions makes purrr easier to maintain because it reduces the surface area for bugs and issues, and it makes purrr easier to learn because there's a clearer common thread that ties together all functions.

## Documentation

As you've seen in the code above, we are moving from magrittr's pipe (`%>%`) to the base pipe (`|>`) and from formula syntax (`~ .x + 1`) to R's new anonymous function short hand (`\(x) x + 1`).
We believe that it's better to use these new base tools because they work everywhere: the base pipe doesn't requite that you load magrittr and the new function shorthand works everywhere, not just in purrr functions.
Additionally, being able to specify the argument name for the anonymous function can often lead to clearer code.

```{r}
# Previously we wrote
1:10 %>%
  map(~ rnorm(10, .x)) %>%
  map_dbl(mean)

# Now we recommend
1:10 |>
  map(\(mu) rnorm(10, mu)) |>
  map_dbl(mean) 
```

Due to the [tidyverse R dependency policy](https://www.tidyverse.org/blog/2019/04/r-version-support/), purrr works in R 3.5, 3.6, 4.0, 4.1, and 4.2, but the base pipe and anonymous function syntax are only available in R 4.0 and later.
So the examples are automatically disabled on R 3.5 and 3.6 to allow purrr to continue to pass `R CMD check`.

## Mapping

With that out of the way, we can now talk about the exciting new features in purrr 1.0.0.
We'll start with the map family of functions which have three big new features:

-   Progress bars.
-   Better errors.
-   A new family member: `map_vec()`.

These are described in the following sections.

### Progress bars

```{asciicast}
#| include: false
library(purrr)
options(cli.progress_show_after = 0)
options(cli.progress_clear = FALSE)
```

The map family of function can now produce a progress bar.
This is very useful for long running jobs:

```{asciicast progress}
#| asciicast_at: "all"
#| cache: true
x <- map(1:100, \(x) Sys.sleep(0.1), .progress = TRUE)
```

(For interactive use, the progress bar uses some simple heuristics so that it doesn't show up for very simple jobs.)

In most cases, we expect that `.progress = TRUE` is enough, but if you're wrapping `map()` in another function, you might want to set `.progress` to a string that identifies the progress bar:

```{asciicast named-progress}
#| asciicast_at: "all"
#| cache: true
walk(1:100, \(i) Sys.sleep(0.1), .progress = "Saving plots")
```

### Better errors

If there's an error in the function you're mapping, `map()` and friends now tell you which element caused the problem:

```{r}
#| error: true
x <- sample(1:500)
x |> map(\(x) if (x == 1) stop("Error!") else 10)
```

We hope that this makes your debugging life just a little bit easier!
(Don't forget about `safely()` and `possibly()` if you expect failures and want to either ignore or capture them.)

We have also generally reviewed the error messages throughout purrr in order to make them more actionable.
If you hit a confusing error message, please let us know!

### New `map_vec()`

We've added `map_vec()` (along with `map2_vec()`, and `pmap_vec()`) to handle more types of vector.
`map_vec()` extends `map_lgl()`, `map_int()`, `map_dbl()`, and `map_chr()` to arbitrary types of vector, like dates, factors, and date-times:

```{r}
1:3 |> map_vec(\(i) factor(letters[i]))
1:3 |> map_vec(\(i) factor(letters[i], levels = letters[4:1]))

1:3 |> map_vec(\(i) as.Date(ISOdate(i + 2022, 10, 5)))
1:3 |> map_vec(\(i) ISOdate(i + 2022, 10, 5))
```

`map_vec()` exists somewhat in the middle of base R's `sapply()` and `vapply()`.
Unlike `sapply()` it will always return a simpler vector, erroring if there's no common type:

```{r}
#| error: true
list("a", 1) |> map_vec(identity)
```

If you want to require a certain type of output, supply `.ptype`, making `map_vec()` behave more like `vapply()`.
`ptype` is short for prototype, and should be a vector that exemplifies the type of output you expect.

```{r}
#| error: true

x <- list("a", "b") 
x |> map_vec(identity, .ptype = character())

# will error if the result can't be automatically coerced
# to the specified ptype
x |> map_vec(identity, .ptype = integer())
```

We don't expect you to know or memorise the [rules that vctrs uses for coercion](https://vctrs.r-lib.org/reference/faq-compatibility-types.html); our hope is that they'll become second nature as we steadily ensure that every tidyverse function follows the same rules.

## `keep_at()` and `discard_at()`

purrr has gained a new pair of functions, `keep_at()` and `discard_at()`, that work like `keep()` and `discard()` but operate on names rather than values:

```{r}
x <- list(a = 1, b = 2, c = 3, D = 4, E = 5)

x |> 
  keep_at(c("a", "b", "c")) |> 
  str()

x |> 
  discard_at(c("a", "b", "c")) |> 
  str()
```

Alternatively, you can supply a function that is called with the names of the elements and should return a logical vector describing which elements to keep/discard:

```{r}
is_lower_case <- function(x) x == tolower(x)

x |> keep_at(is_lower_case)
```

You can now also pass such a function to all other `_at()` functions:

```{r}
x |> 
  modify_at(is_lower_case, \(x) x * 100) |> 
  str()
```

## Flattening and simplification

Last, but not least, we've reworked the family functions that flatten and simplify lists.
These caused us a lot of confusion internally because folks (and different packages) used the same words to mean different things.
Now there are three main functions that share a common prefix that makes it clear that they all operate on lists:

-   `list_flatten()` removes a single level of hierarchy from a list; the output is always a list.
-   `list_simplify()` reduces a list to a homogeneous vector; the output is always the same length as the input.
-   `list_c()`, `list_cbind()`, and `list_rbind()` concatenate the elements of a list to produce a vector or data frame. There are no constraints on the output.

These functions have lead us to **supersede** a number of functions.
This means that they are not going away but we no longer recommend them, and they will receive only critical bug fixes.

-   `flatten()` has been superseded by `list_flatten()`.
-   `flatten_lgl()`, `flatten_int()`, `flatten_dbl()`, and `flatten_chr()` have been superseded by `list_c()`.
-   `flatten_dfr()` and `flatten_dfc()` have been superseded by `list_rbind()` and `list_cbind()` respectively. `flatten_dfr()` had some particularly puzzling edge cases when the inputs would be flattened into columns.
-   `map_dfc()` and `map_dfr()` (and their `map2` and `pmap` variants) have been superseded in favour of using the appropriate map function along with `list_rbind()` or `list_cbind()`.
-   `simplify()`, `simplify_all()`, and `as_vector()` have been superseded in favour of `list_simplify()`.

### Flattening

`list_flatten()` which removes one layer of hierarchy from a list.
In other words, if any of the children of the list are themselves lists, the contents of those lists are inlined into the parent:

```{r}
x <- list(1, list(2, list(3, 4), 5))
x |> str()
x |> list_flatten() |> str()
x |> list_flatten() |> list_flatten() |> str()
```

`list_flatten()` always returns a list; once a list is as flat as it can get (i.e. none of its children contain lists), it leaves the input unchanged.

```{r}
x |> list_flatten() |> list_flatten() |> list_flatten() |> str()
```

### Simplification

`list_simplify()` maintains the length of the input, but produces a simpler type:

```{r}
list(1, 2, 3) |> list_simplify()
list("a", "b", "c") |> list_simplify()
```

Because the length must stay the same, it will only succeed if every element has length 1:

```{r}
#| error: true
list_simplify(list(1, 2, 3:4))
list_simplify(list(1, 2, integer()))
```

Because the result must be a simpler vector, all the components must be compatible:

```{r}
#| error: true
list_simplify(list(1, 2, "a"))
```

If you need to simplify if it's possible, but otherwise leave the input unchanged, use `strict = FALSE`:

```{r}
list_simplify(list(1, 2, "a"), strict = FALSE)
```

If you want to be specific the type you want, `list_simplify()` can take the same prototype argument as `map_vec()`:

```{r}
#| error: true
list(1, 2, 3) |> list_simplify(ptype = integer())

list(1, 2, 3) |> list_simplify(ptype = factor())
```

### Concatenation

`list_c()`, `list_cbind()`, and `list_rbind()` concatenate all elements together in a similar way to using `do.call(c)` or `do.call(rbind)`[^1] .
Unlike `list_simplfy()`, this allows the elements to be different lengths:

[^1]: But of they use the tidyverse coercion rules.

```{r}
list(1, 2, 3) |> list_c()
list(1, 2, 3:4, integer()) |> list_c()

```

The downside of this flexibility is that these functions break the connection between the input and the output.
This reveals that `map_dfr()` and `map_dfc()` don't really belong to the map family because they don't maintain a 1-to-1 mapping between input and output: there's reliable no way to associate a row in the output with an element in an input.

For this reason, `map_dfr()` and `map_dfc()` (and the `map2` and `pmap`) variants are superseded and we recommend switching to an explicit call to `list_rbind()` or `list_cbind()` instead:

```{r}
#| eval: false

paths |> map_dfr(read_csv, .id = "path")
# now
paths |> 
  map(read_csv) |> 
  list_rbind(names_to = "path")
```

This new behaviour has some follow on consequences to `accumulate()` and `accumulate2()`, which previously had an idiosyncratic approach to simplification.

### `list_update()` functions

There's one other new function that isn't directly related to flattening and friends, but shares the `list_` prefix: `list_update()`.
`list_update()` is similar to `list_modify()` but it doesn't work recursively.
This is a mildly confusing feature of `list_modify()` that it's easy to miss in the documentation.

```{r}
list(x = 1, y = list(a = 1)) |> 
  list_modify(y = list(b = 1)) |> 
  str()
```

`list_update()` doesn't recurse into sublists making it a bit easier to reason about:

```{r}
list(x = 1, y = list(a = 1)) |> 
  list_update(y = list(b = 2)) |> 
  str()
```

## Acknowledgements
