---
output: hugodown::hugo_document

slug: purrr-1-0-0
title: purrr 1.0.0
date: 2022-11-10
author: Hadley Wickham
description: >
    purrr 1.0.0 brings a basket of updates to purrr. We deprecated a number of
    seldom used functions to hone in on the core purpose of purrr while 
    implemented a swath of new features including progress bars, improved 
    error reporting, and much much more!

photo:
  url: https://unsplash.com/photos/YCPkW_r_6uA
  author: Jari Hytönen

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [purrr]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [] `hugodown::use_tidy_thumbnails()`
* [ ] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->
```
```{r}
#| include: false
asciicast::init_knitr_engine()
```

We're happy to announce the release of [purrr](http://purrr.tidyverse.org/) 1.0.0!
purrr enhances R's functional programming toolkit by providing a complete and consistent set of tools for working with functions and vectors.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("purrr")
```

purrr is 7 years old, but it's finally made it to 1.0.0!
This was an opportunity to really refine the core purrrpose of purrr, making it more purrrsimonious by moving a number of functions to purrrgatory.
Hopefully these changes are not cat-istrophic

Ok, now that I've got all the purrr related puns out of system ...

You can see a full list of changes in the [release notes](%7B%20github_release%20%7D)

```{r setup}
library(purrr)
```

## Core purpose refinements

A number of functions have been deprecated to keep purrr focused on it's core purpose of faciliating functional programming in R:

-   `cross()` and all its variants have been deprecated because they're slow and buggy, and a better approach already exists in `tidyr::expand_grid()`.

-   `update_list()`, `rerun()`, and the use of tidyselect with `map_at()` and friends have been deprecated, because we no longer believe that non-standard evaluation is a good fit for purrr.

-   The `lift_*` family of functions has been deprecated because they rely on a style of function manipulation that is uncommon in R.

-   `prepend()`, `rdunif()`, `rbernoulli()`, `when()`, and `list_along()` have all been deprecated because they don't align with the core purpose of purrr.

-   `splice()` was deprecated because we no longer believe that automatic splicing makes for good UI.

These functions will continue to work for a number of years, but will warn you once every 8 hours when you use them.
A future release in several years time will cause the warnings to happen on every use, and we'll remove them in a release several years after that.

Deprecating these functions makes purrr easier to maintain because it reduces the surface area for bugs and issues, and it makes purrr easier to learn because there's a clearer common thread that ties together all functons.

## Other breaking changes

As part of our new policy, I made pull requests to all [CRAN packages that broke](https://github.com/tidyverse/purrr/issues/969) (except for the 1 that wasn't on GitHub).
Out of \~1,400 dependencies only \~40 had problems.
I've found making these PRs very empathy building and I'm getting much faster at parachuting into a random package that I have no idea what it does and fixing the problems.
This act also gave me confidence that we'll we're deprecating quite a few functions and changing a few special cases, it shouldn't affect much code in the wild.

### pluck and zero-length vectors

Previously, `pluck()` would replace 0-length vectors with the value of `default`.
Now only `NULL` and absent elements will be replaced with `default`:

```{r}
x <- list(y = list(a = character(), b = NULL))
pluck(x, "y", "a", .default = NA)
pluck(x, "y", "b", .default = NA)
pluck(x, "y", "c", .default = NA)
```

This also influences the map family because using an integer vector, character vector, or list automatically calls `pluck()`:

```{r}
x <- list(list(1), list(), list(NULL), list(character()))
map(x, 1, .default = NA)
```

We made this change because it makes purrr more consistent with the rest of the tidyverse which distinguishes zero-length vectors from `NULL`s.

### Tidyverse consistency

Similarly, `map2()` and `pmap()` now obey the tidyverse recycling rules, more on that below.

And `map_chr()` is no longer so permissive.

### `map_depth()`

There was a bug in `map_depth()` - it was documented to error if you requested a depth that didn't exist.
This caused three packages to fail.

### Assigning `NULL`

In `pluck<-`, `assign_in()`, `modify()`, `modify2()`, `modify_if()`, `list_modify()` assigning a `NULL` value now creates a `NULL` value in the output, rather than deleting that element.
If you want to delete an entry, now use `zap()`.

```{r}
x <- list(a = 1, b = 2)
x$a <- NULL
str(x)
```

So how do you insert a `NULL` using base R?
You have to switch to `[` and wrap the `NULL` in a list:

```{r}
x <- list(a = 1, b = 2)
x["a"] <- list(NULL)
str(x)
```

With purrr:

```{r}
x |> list_modify(a = NULL) |> str()
```

If you want to delete it, you'll need to use the special `zap()` sentinel:

```{r}
x |> list_modify(a = zap()) |> str()
```

## Documentation and licensing

In purrr's documentation, we have switched to using the base pipe (`|>`) instead of magrittr's pie (`%>%`) and R's anonymous function short hand (`\(x) x + 1`) instead of formula syntax (`~ .x + 1`).
We believe that these are more readable because they can be used in every package.

Note, that due to the [tidyverse R dependency policy](https://www.tidyverse.org/blog/2019/04/r-version-support/), purrr works in R 3.5, 3.6, 4.0, 4.1, and 4.2, but the base pipe and anonymous function sytnax are only available in R 4.0 and later.
To allow purrr to continue to pass `R CMD check`, the examples are automatically disabled in older versions of R.

Similarly, inline with our new tidyverse policy [purrr has been re-licensed](https://www.tidyverse.org/blog/2021/12/relicensing-packages/) with the MIT license.

## Mapping

The map functions have received a major overhaul:

-   Progress bars
-   Errors give location
-   New `map_vec()` generalises `map_lgl()`, `map_int()` and friends.
-   Tidyverse consistency

### Progress bars

```{asciicast}
#| include: false
library(purrr)
options(cli.progress_show_after = 0)
options(cli.progress_clear = FALSE)
```

```{asciicast progress}
#| asciicast_at: "all"
#| cache: true
x <- map(1:500, \(x) Sys.sleep(0.01), .progress = TRUE)
```

In most cases, we expect that `.progress = TRUE` will give you a decent progress bar.
But if you're wrapping the `map()` in a function, you might want to set it to a string that 's used to identify the progress bar.
And if needed, you can supply a list, allowing you full control of the progress bar.

### Better errors

If there's an error in the function you're mapping, `map()` and friends now tell which element caused the problem:

```{r}
#| error: true
x <- sample(1:500)
map(x, \(x) if (x == 1) stop("Error!") else 10)
```

We hope that this makes your debugging life just a little bit easier!

(And don't forget about `safely()` and `possibly()` if you expect failures and want to either ignore or capture them.)

### New `map_vec()`

New `map_vec()`, `map2_vec()`, and `pmap_vec()` work on all types of vectors, extending `map_lgl()`, `map_int()`, and friends so that you can easily work with dates, factors, date-times and more

```{r}
#| error: true

1:3 %>% map_vec(factor)
1:3 %>% map_vec(\(year) ISOdate(year + 2022, 10, 5))
```

Obeys the vctrs rules, which are also used in `dplyr::bind_rows()`, dplyr joins, and many other places.

If you want to require a certain type of output, supply `.ptype`.
`ptype` is short for prototype, and should be vector that exemplifies the type of output you expect (this vector is usually empty, but it doesn't have to be).

```{r}
#| error: true

# in vctrs, converting a factor to a character is generally a free transformation:
1:3 %>% map_vec(factor, .ptype = character())

# but converting it to an integer is an error
1:3 %>% map_vec(factor, .ptype = integer())
```

We don't expect you to know or memorise these rules; our hope is that as we slowly ensure that every tidyverse functions follows the same rules that these will be become second nature.

### Tidyverse consistency

vctrs has also had an influence on `map_lgl()`, `map_int()`, `map_int()`, and `map_dbl()`, and they now follow the same coercion rules as vctrs.
This means that:

-   `map_chr(TRUE, identity)`, `map_chr(0L, identity)`, and `map_chr(1L, identity)` are deprecated because we now believe that converting a logical/integer/double to a character vector should require an explicit coercion.

    ```{r}
    # previously
    map_chr(1:4, \(x) x + 1)

    # now
    map_chr(1:4, \(x) as.character(x + 1))
    ```

-   `map_int(1.5, identity)` now fails because we believe that silently truncating doubles to integers is dangerous.
    But note that `map_int(1, identity)` still works since no numeric precision is lost.

    ```{r}
    #| error: true
    map_int(1:3, \(x) x / 2)

    map_int(1:3, \(x) x * 2)
    ```

Additionally, `map2()`, `modify2()`, and `pmap()` use tidyverse recycling rules where vectors of length 1 are recycled to any size but all others must have the same length.

## `keep_at()` and `discard_at()`

-   New `keep_at()` and `discard_at()` that work like `keep()` and `discard()` but operation on element names rather than element contents (#817).

-   `*_at()` can now take a function (or formula) that's passed the vector of element names and returns the elements to select.

## Flattening and simplification

-   New `list_c()`, `list_rbind()`, and `list_cbind()` make it easy to `c()`, `rbind()`, or `cbind()` all of the elements in a list.

-   New `list_simplify()` reduces a list of length-1 vectors to a simpler atomic or S3 vector (#900).

-   New `list_transpose()` which automatically simplifies if possible (#875).

-   `flatten()` and friends are superseded in favour of `list_flatten()`, `list_c()`, `list_cbind()`, and `list_rbind()`.

-   `*_dfc()` and `*_dfr()` have been superseded in favour of using the appropriate map function along with `list_rbind()` or `list_cbind()` (#912).

-   `simplify()`, `simplify_all()`, and `as_vector()` have been superseded in favour of `list_simplify()`.
    It provides a more consistent definition of simplification (#900).

-   `transpose()` has been superseded in favour of `list_transpose()` (#875).
    It has built-in simplification.

### `list_` functions

-   New `list_update()` which is similar to `list_modify()` but doesn't work recursively (#822).

-   `list_modify()` no longer recurses into data frames (and other objects built on top of lists that are fundamentally non-list like) (#810).

## Acknowledgements
