---
output: hugodown::hugo_document

slug: scales-1-4-0
title: scales 1.4.0
date: 2024-10-30
author: Teun van den Brand
description: >
    The new 1.4.0 release of the scales package adds some colourful updates.
    Read about colour manipulation, palettes and new label functions.

photo:
  url: https://unsplash.com/photos/a-close-up-of-a-person-holding-a-paintbrush-Xrelr7cTYm4
  author: Jennie Razumnaya

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [scales]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're stoked to announce the release of [scales]({https://scales.r-lib.org/}) 1.4.0. 
scales is a package that provides much of the scaling logic that is used in ggplot2 to a general framework, along with utility functions for e.g. formatting labels or creating colour palettes.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("scales")
```

This blog post will give an overview of the 1.4.0 release, which has some nifty upgrades for working with colours.

You can see a full list of changes in the [release notes](https://scales.r-lib.org/news/index.html)

```{r setup}
library(scales)
```

## Colour manipulation

The `alpha()` and `muted()` functions have been part of scales for a long time. 
Back in the 1.1.0 release we swapped to [farver](https://farver.data-imaginist.com/) to power these functions.
We felt it was appropriate to use this package for other common colour tasks, and so `col_shift()`, `col_lighter()`, `col_darker()`, `col_saturate()` and `col_mix()` were born.

```{r}
my_colours <- c("red", "green", "blue")

m <- rbind(
  original = my_colours,
  shift    = col_shift(my_colours, 90),
  lighter  = col_lighter(my_colours, 20),
  darker   = col_darker(my_colours, 20),
  duller   = col_saturate(my_colours, -50),
  mixed    = col_mix(my_colours, "orchid")
)

show_col(t(m), ncol = ncol(m))
text(x = ncol(m) + 0.25, y = -(1:nrow(m)) + 0.5, rownames(m), adj = 0)
```

## Palettes

Also palettes have been reworked this release to reflect more useful properties.
Palettes now come in one of two classes: 'pal_discrete' or 'pal_continuous'.

```{r}
my_palette <- manual_pal(c("palegreen", "deepskyblue", "magenta"))
class(my_palette)
```

Having palettes as a class rather than plain functions, allows us to store useful metadata about the palette.
In addition, most colour palette functions also allow the aforementioned colour manipulation functions to work on the palette output.

```{r}
palette_type(my_palette)
palette_nlevels(my_palette)
col_shift(my_palette, 180)(3)
```

This metadata can then be used to expand discrete palettes to continuous palettes with `as_continuous_pal()` or vise versa to chop up a continuous palette into discrete palettes with `as_discrete_pal()`.

```{r}
plot(as_continuous_pal(my_palette))
```

Another thing to make working with palettes easier, is that the 'scales' package now keeps track of named palettes.
By default, the collection of 'known' palettes is pre-populated with colour palettes from the grDevices, RColorBrewer and viridisLite packages.


```{r}
head(palette_names())
get_palette("Okabe-Ito")(8)
```

If you're a developer of a palette package, you can use `set_palette()` to register your palette.
This has the advantage that your palette is now available to users by name, which at times might be more convenient than having to call the palette generator function.

```{r, error=TRUE}
get_palette("aurora")

set_palette("aurora", my_palette)
plot(get_palette("aurora"))
```

## Labels

Lastly, please let us introduce you to our two new labelling functions and two new convenience functions for labels.
In contrast to most of scales' label functions, these label functions are great for discrete input.
First up is `label_glue()`, which uses the string interpolation from the glue package to format your labels.

```{r}
label_glue("The {x} penguin")(c("Gentoo", "Chinstrap", "Adelie"))
```

The next labeling function is convenient when some variable you use consists of shortcodes or abbreviations.
You can provide `label_dictionary()` with a named vector that translates the values to prettier labels.
If you value didn't exist in the dictionary, these stay as-is by default.

```{r}
dict <- c(
  diy = "Do it yourself", eta = "Estimated time of arrival",
  asap = "As soon as possible", tldr = "Too long; didn't read"
)
label_dictionary(dict)(c("diy", "tldr", "bff"))
```

The first label convenience function we'd like to tell you about is the `compose_label()` function.
Similar to `compose_trans()`, it allows you to chain together different labelling functions.

```{r}
screaming_flowers <- compose_label(label_glue("The {x} flower"), toupper)
screaming_flowers(c("daffodil", "orchid", "tulip"))
```
Lastly, we haven't completely forgotton about numeric labels either.
We have introduced the `number_options()` functions to globally populate defaults for functions such as `label_number()` and `label_currency()`.
This can be convenient if you produce statistical reports in non-English languages.

```{r}
number_options(
  decimal.mark = ",",
  big.mark = ".",
  style_negative = "minus",
  currency.prefix = "",
  currency.suffix = "€",
  currency.decimal.mark = ",",
  currency.big.mark = " ",
  ordinal.rules = ordinal_french()
)

label_currency(accuracy = 0.01)(c(0.1, 10, 1000000, -1000))
label_ordinal()(1:4)
```



## Acknowledgements

We'd like to thank all people who have contributed in some way, whether it was filing issues, participating in discussion or contributing to code and documentation:

[&#x0040;Aariq](https://github.com/Aariq), [&#x0040;Aehmlo](https://github.com/Aehmlo), [&#x0040;Ali-Hudson](https://github.com/Ali-Hudson), [&#x0040;cb12991](https://github.com/cb12991), [&#x0040;colindouglas](https://github.com/colindouglas), [&#x0040;d-morrison](https://github.com/d-morrison), [&#x0040;davidhodge931](https://github.com/davidhodge931), [&#x0040;EricMarcon](https://github.com/EricMarcon), [&#x0040;kellijohnson-NOAA](https://github.com/kellijohnson-NOAA), [&#x0040;kmcd39](https://github.com/kmcd39), [&#x0040;lz1nwm](https://github.com/lz1nwm), [&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [&#x0040;mjskay](https://github.com/mjskay), [&#x0040;Moohan](https://github.com/Moohan), [&#x0040;muschellij2](https://github.com/muschellij2), [&#x0040;ppreshant](https://github.com/ppreshant), [&#x0040;rawktheuniversemon](https://github.com/rawktheuniversemon), [&#x0040;rogiersbart](https://github.com/rogiersbart), [&#x0040;SchmidtPaul](https://github.com/SchmidtPaul), [&#x0040;teunbrand](https://github.com/teunbrand), and [&#x0040;thomasp85](https://github.com/thomasp85).


