---
output: hugodown::hugo_document
slug: playing-on-the-same-team-as-your-dependecy
title: Playing on the same team as your dependecy
date: 2022-09-29
author: Thomas Lin Pedersen
description: >
    Using another package as a dependency is a two-way street, but the
    expectations can be murky. This blog post guides you towards becoming a 
    stellar reverse dependency.
photo:
  url: https://unsplash.com/photos/G1hIBdjQoAA
  author: Florian Schmetz
# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [learn] 
tags: []
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<!--
TODO:
* [ ] Look over / edit the post's title in the yaml
* [ ] Edit (or delete) the description; note this appears in the Twitter card
* [ ] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [ ] Find photo & update yaml metadata
* [ ] Create `thumbnail-sq.jpg`; height and width should be equal
* [ ] Create `thumbnail-wd.jpg`; width should be >5x height
* [ ] `hugodown::use_tidy_thumbnails()`
* [ ] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->
```
Developing packages for R is a matter of standing on the shoulders of
others. Very seldom does packages exist in a vacuum --- on the contrary,
we often rely on dependencies to avoid duplication of code or lean into
the work done by experts within an adjacent field.

It can easily feel like a one way relationship to take on a dependency
on another package. You are responsible for keeping your package working
and the developer of the dependency can ignore whatever goes on in your
package. Code flows only from the dependency to your package. This is
not true, though. By taking on a dependency you enter into a mutual
relationship with it. The dependency implicitly promises to not change
it's interface without providing an upgrade path to you. You, on the
other hand promises to only rely on the public interface of the package.
This blog post goes into detail with what your promise entails.

### Why does this matter?

As a developer you may be surprised to learn that the dependency's
promise is enforced by CRAN. When submitting a new version for release
the package goes through a battery of tests, among these a reverse
dependency check where all packages on CRAN that depends on the
submitted package is checked against the new version. If any regressions
have occurred it is flagged. The CRAN repository policy states:

> If an update will change the package\'s API and hence affect packages
> depending on it, it is expected that you will contact the maintainers
> of affected packages and suggest changes, and give them time (at least
> 2 weeks, ideally more) to prepare updates before submitting your
> updated package.

This is good in general --- it *is* important that a package maintains a
stable interface across versions --- but can become a huge obstacle to
updates if the packages that depends on you are reaching behind the
curtain and making assumptions you never promised to adhere to.

## What's in an API?

For better and worse, R as a language is extremely liberal with what you
can access as a user. There is practically no data or function you can't
access and modify which makes the concept of APIs a question of
conventions. Those conventions are quite well defined when it comes to
functions in packages, but much less so for everything else. We will
discuss functions first and then proceed into the more gray area of
objects and data.

### Exported functions

When creating a package you are now required to provide a NAMESPACE file
which states the functions you import *into* your package for use, and
the functions you export *out of* your package for others to use. The
NAMESPACE file demarcates in very clear terms the functional interface
of a package but is still based on mutual trust. While you cannot import
functions from a package that have not been exported, there is nothing
in the R language that prevents you from using them by accessing them
directly. Below we will talk about several ways of doing this and why
all of them has issues:

#### Using `:::`

R provides two operators for accessing objects in a package namespace:
`::` allows you to fetch exported objects and functions while `:::`
allows you to access *any* object and function. Thus, you could for
instance gain access to the internal `camelize()` function in ggplot2 to
convert geom function names into ggproto object names like so:

```{r}
ggplot2:::camelize('geom_point', first = TRUE)
```

*But* since there is no need for `:::` except for reaching beyond the
package interface it's use is actively checked and packages using it is
rejected from CRAN.

#### Using `utils::getFromNamespace()`

To circumvent the detection of `:::` we sometimes see code like the
following in packages:

```{r}
camelize <- utils::getFromNamespace("camelize", "ggplot2")
```

There are two huge issues with this approach. The first being that you
now have sneakily accessed something that was never meant for public
consumption (this is a general theme). The second is that you are
grabbing a function from another package *at build time*. This means
that the `camelize()` function living in your package is the one from
the ggplot2 version available when your package got build on CRAN. Why
is that a problem? Consider again `camelize()`:

```{r}
ggplot2:::camelize
```

We can see that it contains a call to `firstUpper()` which is another
internal function. As ggplot2 developers we might decide one day that
this factorization of code is too granular and inline the code of
`firstUpper()` into `camelize()`, allowing us to remove `firstUpper()`
altogether.

```{r, eval=FALSE}
# New version of camelize
camelize <- function(x, first = FALSE) {
  x <- gsub("_(.)", "\\U\\1", x, perl = TRUE)
  if (first) x <- paste0(to_upper_ascii(substring(x, 1, 1)), substring(x, 2))
  x
}
```

All of that would be perfectly fine for us to do. After all we are not
changing the public interface of ggplot2, we aren't even changing how
`camelize()` works. But, in packages that have fetched `camelize()` at
build time the function would be unchanged, still calling `firstUpper()`
who now no longer exists. As you might imagine this can lead to some
very hard to debug errors for, you, your users, and the maintainer of
the dependency.

#### Use `asNamespace()` inside a function

This rule can be extended beyond its use to access unexported functions:
**Never assign a function from another package to a variable in your own
package**. You might import a function from a package developed by
someone who prefers long and descriptive function names, say
`add_these_two_objects_together()`, and find it easier to create a
shorthand version:

```{r, eval=FALSE}
add2 <- add_these_two_objects_together
```

While `add_these_two_objects_together` is exported and you are doing
nothing wrong in terms of interfaces, you are still setting up a
build-time dependency that might cause breakage any time your dependency
gets updated on a system.

Thus we arrive at the last approach: Fetching the function inside a
function call and then use it. In the example below we are using
`asNamespace()` but the same principle holds true for
`getFromNamespace()`

```{r}
camelize <- function(...) {
  asNamespace("ggplot2")$camelize(...)
}
```

Now, while this is many times better than what we did before, it is
still a big red flag. Consider the same situation as before. We inline
every use of `camelize()` in ggplot2 (it's only used once), and remove
the function. This will again lead to a breakage of your package when
ggplot2 got updated because you made assumptions that ggplot2 never
promised anything about.

#### What to do?

What if you really wanted that functionality? A good first approach is
to simply copy the code for the function into your own package. For
something like `camelize()` this is fairly simple as it doesn't call
into other internal functions (except `firstUpper()` but we saw that
that could be inlined). One thing to keep in mind here is to make sure
that the licence of the dependency doesn't prevent you from doing this
(e.g. a package released under MIT license can't copy code from a
package released under a GPL-2 licence).

If you can't copy the code into your own package, either due to
incompatible licenses or because the function is a rabbit hole of
internal function calls, you'll need to reach out to the maintainer and
ask whether the required function can be exported so you can use it.
Keep in mind that there are many good reasons why you could get a "no"
since every new export increases the maintenance burden of a package.
So, you can get a "yes" and all is well, or you might get a "no" and
have to accept that as well. Getting a "no" is not a blanket approval to
do any of the above things we have discussed, for the exact reasons we
have discussed, rather it means you have to reframe your solution so it
doesn't require this functionality or abandon it altogether.

### Exported structures

While the situation with functions is quite clear-cut --- there are
*do's* and *don't* --- we enter a much grayer area when it comes to any
sort of data/object structure you get from a dependency, either as an
object exported by the package or as a return value from an exported
function. The reason why it is a gray area is that there is no formal
way to specify an interface to on object in R and the users are used to
an "everything goes" mentality when it comes to reaching into data
structures. For example, while attributes are a bit more "hidden away"
than elements in a list, there is no notion of these being prohibited
from access. There might be a mutual understanding that if you alter
attributes in some way it might lead to breakage somewhere downstream,
but merely reading attributes as a pretty common thing to do. The same
goes for more complex object that contains more than just data. An
example is the object created by a call to `ggplot()`

```{r}
str(ggplot2::ggplot())
```

This is obviously more than just data, but what, if any, is actually
fair to access as a package developer? This is a tough question to
answer in general terms.

If you want to be a very polite (and who wouldn't), the best way to go
about it is to look for accessor functions for the part of the object
you are interested in and in the absence of one, ask the maintainer to
add one. The reason why accessor functions are so much better than
relying on e.g. `attr()` to extract some information stored in an
attribute, is that it frees the maintain to change the *structure* of
the data/object, while keeping the *interface* constant. Asking a
maintainer for a public accessor function will also alert the maintainer
to the fact that others are actually interested in said information
which could inform future development.

#### Testing, testing

You may be the most polite package developer, using only the finest
public accessor functions in your code and keeping out of any data
structure you don't control the provenance off and still be reliant of
implementation details in objects from other packages. How? You may
inadvertently test for their internal details in your unit tests when
you are comparing objects wholesale, or if you have saved complex
objects and load these up during testing.

Once again we are certainly in a gray area here, but one guide line to
help you is to ask yourself whether your unit test is only testing for
parts that your own package influence, or does it also include
assumptions about implementation details of another package. As an
example (once again from ggplot2), you might want to ensure that a plot
function in your package works as intended. On one extreme end you can
save a working ggplot object returned from your function and then test
for equivalence with that during unit testing. This is not a great idea
because anything we might change internally in ggplot2 would likely
result in changes to the created ggplot2 object, and while it still
works, it may look slightly different. On the other end you may instead
do visual testing using the vdiffr package where you only look at the
actual output. However, that also makes a lot of assumptions about how
ggplot2 chooses to render it's objects and internal changes may again
break your tests without there being anything broken in reality.

> Visual testing in general is something that is mainly intended for
> packages providing graphic rendering, e.g. ggplot2 and it's extension
> package ecosystem. If you are using other packages to create your
> plots you should in general lean on them to test for visual
> regressions.

The Goldilocks zone for your testing is to figure out which exact
elements your high-level plot function influences and then get to these,
preferably using public accessor functions. For ggplot2 it will often be
enough to extract the data for each layer (using `layer_data()`) and
test specific columns of that (never test against the full layer data
since ggplot2 may add to this etc.).

If you find that you are missing public accessor function in order to do
proper testing, once again reach out to the maintainer and ask. You may
learn that this information is not exposed because it is subject to
change, thus a poor fit for unit testing, or you may get your function
and end up with more robust tests in your own package.
