---
output: hugodown::hugo_document
slug: dplyr-1-1-0-per-operation-grouping
title: "dplyr 1.1.0: Per-operation grouping"
date: 2023-02-01
author: Davis Vaughan
description: >
    dplyr now supports an experimental per-operation grouping syntax. This serves as an
    alternative to `group_by()` and always returns an ungrouped data frame, meaning that you
    never need to remember to `ungroup()`.
photo:
  url: https://www.pexels.com/photo/fruit-stand-375897/
  author: Clem Onojeghuo
categories: [package] 
tags: [dplyr, dplyr-1-1-0]
editor_options: 
  chunk_output_type: console
---

Today we are going to look at one of the major new features in [dplyr 1.1.0](https://dplyr.tidyverse.org/news/index.html#dplyr-110), per-operation grouping with [`.by`/`by`](https://dplyr.tidyverse.org/reference/dplyr_by.html).
Per-operation grouping is an experimental alternative to `group_by()` which is only active within a single dplyr verb.
This is another of the new dplyr features that was inspired by [data.table](https://cran.r-project.org/web/packages/data.table/index.html), this time by their own grouping syntax with `by`.

To see the other blog posts in this series, head [here](https://www.tidyverse.org/tags/dplyr-1-1-0/).

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("dplyr")
```

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
```

## Persistent grouping with `group_by()`

In dplyr, grouping radically affects the computation of the verb that you use it with.
Since the very beginning of dplyr, you've been able to perform grouped operations by modifying your data frame with `group_by()`.
This grouping is *persistent*, meaning that it typically sticks around in some form for more than one operation.
As an example, take a look at this `transactions` dataset which tracks revenue brought in from various transactions across multiple companies.
If we wanted to add a column for the total yearly revenue per company, we might do:

```{r}
transactions <- tibble(
  company = c("A", "A", "A", "B", "B", "B"),
  year = c(2019, 2019, 2020, 2021, 2023, 2023),
  revenue = c(20, 50, 4, 10, 12, 18)
)

transactions
```

```{r}
transactions |>
  group_by(company, year) |>
  mutate(total = sum(revenue))
```

Notice that the result is still grouped by both `company` and `year`.
This is useful if you need to follow up with additional grouped operations (with the exact same grouping columns), but many people follow this `mutate()` with an `ungroup()`.

If we only need the totals, we could also use `summarise()`, which peels off 1 layer of grouping by default:

```{r}
transactions |>
  group_by(company, year) |>
  summarise(total = sum(revenue))
```

Here the grouping of the output isn't exactly the same as the input, but we still consider this persistent grouping because some of the groups outlive the verb they were used with.

## Per-operation grouping with `.by`/`by`

In dplyr 1.1.0, we've added an alternative to `group_by()` known as [`.by`](https://dplyr.tidyverse.org/reference/dplyr_by.html) that introduces the idea of *per-operation* grouping:

```{r}
transactions |>
  mutate(total = sum(revenue), .by = c(company, year))

transactions |>
  summarise(total = sum(revenue), .by = c(company, year))
```

There are a few things about `.by` worth noting:

-   The result is always ungrouped, regardless of the number of grouping columns.
    With `.by`, you never need to remember to call `ungroup()`.

-   We used [tidyselect](https://tidyselect.r-lib.org/reference/language.html) to group by multiple columns.

-   `summarise()` didn't emit a message about regrouping.

One of the things we like about `.by` is that it allows you to place the grouping specification alongside the code that uses it, rather than in a separate `group_by()` line.
This idea was inspired by data.table's grouping syntax, which looks like:

```{r, eval=FALSE}
transactions[, .(total = sum(revenue)), by = .(company, year)]
```

To see a complete list of dplyr verbs that support `.by`, look [here](https://dplyr.tidyverse.org/reference/dplyr_by.html#supported-verbs).

### `.by` or `by`?

As you use per-operation grouping in dplyr, you'll likely notice that some verbs use `.by` and others use `by`, for example:

```{r}
transactions |>
  slice_max(revenue, n = 2, by = company)
```

This is a technical difference resulting from the fact that some verbs consistently use a `.` prefix for their arguments, and others don't (see our design notes on the [dot prefix](https://design.tidyverse.org/dots-prefix.html) for more details).
Most dplyr verbs use `.by`, and we've tried to ensure that the cases that are most likely to result in typos instead generate an informative error:

```{r, error=TRUE}
# Uses `by` to be consistent with `n` and `prop`
transactions |>
  slice_max(revenue, n = 2, .by = company)

# Uses `.by` to be consistent with `.preserve`
transactions |>
  slice(revenue, by = company)
```

### Translating from `group_by()`

You shouldn't feel pressured to translate existing code using `group_by()` to use `.by` instead.
`group_by()` won't ever disappear, and is not currently being superseded.

That said, if you do want to start using `.by`, there are a few differences from `group_by()` to be aware of.

-   `.by` always returns an ungrouped data frame.
    This is one of the main reasons to use `.by`, but is worth keeping in mind if you have existing code that takes advantage of persistent grouping from `group_by()`.

-   `.by` uses tidy-selection.
    `group_by()`, on the other hand, works more like `mutate()` in that it allows you to create grouping columns on the fly, i.e. `df |> group_by(month = floor_date(date, "month"))`.
    With `.by`, you must create your grouping columns ahead of time.
    An added benefit of `.by`'s usage of tidy-selection is that you can supply an external character vector of grouping variables using `.by = all_of(groups_vec)`.

-   `.by` doesn't sort grouping keys.
    `group_by()` always sorts keys in ascending order, which affects the results of verbs like `summarise()`.

The last point might seem strange, but consider what happens if we preferred our transactions data in order by descending year so that the most recent transactions are at the top.

```{r}
transactions2 <- transactions |>
  arrange(company, desc(year))

transactions2
```

```{r}
# Note that `group_by()` re-ordered
transactions2 |>
  group_by(company, year) |>
  summarise(total = sum(revenue), .groups = "drop")

# But `.by` used whatever order was already there
transactions2 |>
  summarise(total = sum(revenue), .by = c(company, year))
```

Notice that `.by` doesn't re-sort the grouping keys.
Instead, the previous call to `arrange()` is "respected" in the summary (this is also useful in combination with the new `.locale` argument to `arrange()`).

We expect that most code won't depend on the ordering of these group keys, but it is worth keeping in mind if you are switching to `.by`.
If you did rely on sorted group keys, you currently need to explicitly call `arrange()` either before or after the call to `summarise(.by =)`.
In a future release, we may add [an argument](https://github.com/tidyverse/dplyr/issues/6663) to control this.

## `nest(.by = )`

```{r}
library(tidyr)
```

The idea behind `.by` turns out to be useful in contexts outside of dplyr.
In [tidyr 1.3.0](https://www.tidyverse.org/blog/2023/01/tidyr-1-3-0/#nestby), `nest()` gained a `.by` argument, allowing you to specify the columns you want to nest *by* rather than the columns that appear in the nested results, which often makes for more natural calls to `nest()`.

```{r}
# Specify what to nest by
transactions |>
  nest(.by = company)

# Specify what to nest
transactions |>
  nest(data = -company)

# Specify both, allowing you to drop `year` along the way
transactions |>
  nest(data = revenue, .by = company)
```

We currently have 3 different nesting variants in the tidyverse: `tidyr::nest()`, `dplyr::group_nest()`, and `dplyr::nest_by()`.
Because the tidyr variant is now the most flexible of all of these, and because `unnest()` also lives in tidyr, we are likely to deprecate the two experimental dplyr options in the future.
