---
output: hugodown::hugo_document

slug: censored-0-1-0
title: censored 0.1.0
date: 2022-07-19
author: Hannah Frick
description: >
    censored 0.1.0 is a new tidymodels package for survival models.

photo:
  url: https://unsplash.com/photos/n6vS3xlnsCc
  author: Kelley Bozarth

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [tidymodels, parsnip, censored]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [ ] Find photo & update yaml metadata
* [ ] Create `thumbnail-sq.jpg`; height and width should be equal
* [ ] Create `thumbnail-wd.jpg`; width should be >5x height
* [ ] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're extremely pleased to announce the release of [censored](https://censored.tidymodels.org) 0.1.0. censored is a extension package for survival models to parsnip, the unified modeling interface for tidymodels.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("censored")
```

This blog post will introduce a new model type, a new mode, and new prediction types for survival analysis in the tidymodels framework. We have [previously](https://www.tidyverse.org/blog/2021/11/survival-analysis-parsnip-adjacent/) blogged about these changes while they were in development, now they have been released!

```{r setup}
library(censored)
```

## Model types, modes, and engines

A parsnip model specification consists of three elements:

-   a **model type** such as linear model, random forest, support vector machine, etc
-   a computational **engine** such as a specific R package or tools outside of R like Keras or Stan
-   a **mode** such as regression or classification

parsnip 1.0.0 introduces a new mode `"censored regression"` and the censored package provides engines to fit various models in this new mode. With the addition of the new `proportional_hazards()` model type, the available models cover parametric, semi-parametric, and tree-based models:

| model                    | engine   |
|:-------------------------|:---------|
| `bag_tree()`             | rpart    |
| `boost_tree()`           | mboost   |
| `decision_tree()`        | rpart    |
| `decision_tree()`        | partykit |
| `proportional_hazards()` | survival |
| `proportional_hazards()` | glmnet   |
| `rand_forest()`          | partykit |
| `survival_reg()`         | survival |
| `survival_reg()`         | flexsurv |

All models can be fitted through a formula interface. Stratification can be specified through adding a `strata()` term to the formula, as in the survival package.

To illustrate the new modelling function `proportional_hazards()` and the formula interface for glmnet, let's fit a penalized Cox model. The `cetaceans` data set contains information about dolphins and whales living in captivity in the USA.

```{r glmnet-fit}
#devtools::install_github("hfrick/cetaceans")
library(censored)
library(cetaceans)

cox_penalized <- proportional_hazards(penalty = 0.1) %>%
  set_engine("glmnet") %>%
  set_mode("censored regression") %>%
  fit(
    Surv(age, event) ~ sex + transfers + strata(born_in_captivity),
    data = cetaceans
  )
```

## Prediction types

For censored regression, parsnip now also includes new prediction types:

- `"time"` for the survival time
- `"survival"` for the survival probability
- `"hazard"` for the hazard
- `"quantile"` for quantiles of the event time distribution
- `"linear_pred"` for the linear predictor

Predictions made with censored respect the tidymodels principles of:

- The predictions are always inside a tibble.
- The column names and types are unsurprising and predictable.
- The number of rows in `new_data` and the output are the same.

Let's demonstrate that with a small data set to predict on: just three observations, and the first one includes a missing value for one of the predictors.

```{r new_data}
cetaceans_3 <- cetaceans[1:3,]
cetaceans_3$sex[1] <- NA
```

Predictions of types `"time"` and `"survival"` are available for all model/engine combinations in censored.

```{r glmnet-time}
predict(cox_penalized, new_data = cetaceans_3, type = "time")
```

Survival probability can be predicted at multiple time points, specified through the `time` argument to `predict()`. Here we are predicting survival probability at age 10, 20, 30, and 40 years.

```{r glmnet-survival}
pred <- predict(cox_penalized, new_data = cetaceans_3, type = "survival", time = c(10, 20, 30, 40))
pred
```

The `.pred` column is a list-column, containing nested tibbles:

```{r glmnet-survival-unnested}
# for the observation with NA
pred$.pred[[1]]

# without NA
pred$.pred[[1]]
```

This can be used to visualize an approximation of the underlying survival curve.

```{r survival-curve}
library(ggplot2)

predict(cox_penalized, new_data = cetaceans[2:3,], type = "survival", time = seq(0, 80, 1)) %>% 
  dplyr::mutate(id = factor(2:3)) %>% 
  tidyr::unnest(cols = .pred) %>% 
  ggplot(aes(x = .time, y = .pred_survival, col = id)) +
  geom_step() +
  theme_bw()
```

More examples of available models, engines, and prediction types can be found in the article [Fitting and Predicting with censored](https://censored.tidymodels.org/articles/examples.html).


## Acknowledgements

[&#x0040;bcjaeger](https://github.com/bcjaeger), [&#x0040;brunocarlin](https://github.com/brunocarlin), [&#x0040;caimiao0714](https://github.com/caimiao0714), [&#x0040;DavisVaughan](https://github.com/DavisVaughan), [&#x0040;dvdsb](https://github.com/dvdsb), [&#x0040;EmilHvitfeldt](https://github.com/EmilHvitfeldt), [&#x0040;erikvona](https://github.com/erikvona), [&#x0040;gvelasq](https://github.com/gvelasq), [&#x0040;hfrick](https://github.com/hfrick), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;mattwarkentin](https://github.com/mattwarkentin), [&#x0040;mikemahoney218](https://github.com/mikemahoney218), [&#x0040;schelhorn](https://github.com/schelhorn), and [&#x0040;topepo](https://github.com/topepo).
