---
output: hugodown::hugo_document

slug: broom-0-7-0
title: broom 0 7 0
date: 2020-06-19
author: Simon Couch
description: >
    The newest release of broom features many new tidier methods, bug fixes, and
    improvements to internal consistency.

photo:
  url: https://unsplash.com/photos/3gS-lDkOuJ4
  author: Timothy Dykes

categories: [package] 
tags:
  - tidymodels
---

I'm excited to announce the release of broom 0.7.0 on CRAN!

broom is a package for summarizing statistical model objects in tidy tibbles. 
While several compatibility updates have been released in recent months, this 
is the first major update to broom in almost two years. This update includes 
many new tidier methods, bug fixes, improvements to existing tidier methods 
and their documentation, and improvements to maintainability and internal 
consistency. The full list of changes is available in the 
package [release notes](https://broom.tidymodels.org/news/index.html).

This release was made possible by the RStudio internship program, which has 
allowed me ([Simon Couch](https://github.com/simonpcouch)) to work on broom 
full-time for the last month.

You can install the most recent broom update with the following code:

```{r, eval = FALSE}
install.packages("broom")
```

Then attach it for use with:

```{r, eval = FALSE}
library(broom)
```

I'll outline some of the more notable changes below!

### New Tidier Methods

For one, this release includes support for several new model objects---many of
these additions came from first-time contributors to broom!

* `anova` objects from the `car` package
* `pam` objects from the `cluster` package
* `drm` objects from the `drc` package
* `summary_emm` objects from the `emmeans` package
* `epi.2by2` objects from the `epiR` package
* `fixest` objects from the `fixest` package
* `regsubsets` objects from the `leaps` package
* `lm.beta` objects from the `lm.beta` package
* `rma` objects from the `metafor` package
* `mfx`, `logitmfx`, `negbinmfx`, `poissonmfx`, `probitmfx`, and `betamfx` 
objects from the`mfx` package
* `lmrob` and `glmrob` objects from the `robustbase` package
* `sarlm` objects from the `spatialreg` package
* `speedglm` objects from the `speedglm` package
* `svyglm` objects from the `survey` package
* We have restored a simplified version of `glance.aov()`

### Improvements and Bug Fixes for Existing Tidiers

This update also features many bug fixes improvements to existing tidiers. 
Some of the more notable ones:

* Many improvements to the consistency of `augment.*()` methods:
    - If you pass a dataset to `augment()` via the `data` or `newdata` arguments,
  you are now guaranteed that the augmented dataset will have exactly the same
  number of rows as the original dataset. This differs from previous behavior
  primarily when there are missing values. Previously `augment()` would drop
  rows containing `NA`. This should no longer be the case. As a result, 
  `augment.*()` methods no longer accept an `na.action` argument.
    - In previous versions, several `augment.*()` methods inherited the 
  `augment.lm()` method, but required additions to the `augment.lm()` method
  itself. We have shifted away from this approach in favor of re-implementing
  many `augment.*()` methods as standalone methods making use of internal 
  helper functions. As a result, `augment.lm()` and some related methods have
  deprecated (previously unused) arguments.
    - The `.resid` column in the output of `augment().*` methods is now 
  consistently defined as `y - y_hat`.
    - `augment()` tries to give an informative error when `data` isn't the original
  training data.
* Several `glance.*()` methods have been refactored in order to return a 
one-row tibble even when the model matrix is rank-deficient.
* Many `glance()` methods now return a `nobs` column, which contains the number 
of data points used to fit the model!
* Various warnings resulting from changes to the tidyr API in v1.0.0 have been 
fixed.
* Added options to provide additional columns in the outputs of 
`glance.biglm()`, `tidy.felm()`, `tidy.lmsobj()`, `tidy.lmodel2()`, 
`tidy.polr()`, `tidy.prcomp()`, `tidy.zoo()`, `tidy_optim()`

### Breaking Changes and Deprecations

This release also contains a number of breaking changes and deprecations 
meant to improve maintainability and internal consistency.

* We have changed how we report degrees of freedom for `lm` objects. 
This is especially important for instructors in statistics 
courses. Previously the `df` column in `glance.lm()` reported the rank of the 
design matrix. Now it reports degrees of freedom of the numerator for the 
overall F-statistic. This is equal to the rank of the model matrix minus one 
(unless you omit an intercept column), so the new `df` should be the old 
`df` minus one.
* We are moving away from supporting `summary.*()` objects. In particular, we 
have removed `tidy.summary.lm()` as part of a major overhaul of internals. 
Instead of calling `tidy()` on `summary`-like objects, please call `tidy()` 
directly on model objects moving forward.
* We have removed all support for the `quick` argument in `tidy()` methods. 
This is to simplify internals and is for maintainability purposes. We anticipate
this will not influence many users as few people seemed to use it. If this 
majorly cramps your style, let us know, as we are considering a new verb to 
return only model parameters. In the meantime, `stats::coef()` together with 
`tibble::enframe()` provides most of the functionality 
of `tidy(..., quick = TRUE)`.
* All `conf.int` arguments now default to `FALSE`, and all `conf.level` 
arguments now default to `0.95`. This should primarily affect `tidy.survreg()`, 
which previously always returned confidence intervals, although there are 
some others.
* Tidiers for `emmeans`-objects use the arguments `conf.int` and `conf.level` 
instead of relying on the argument names native to 
the `emmeans::summary()`-methods (i.e., `infer` and `level`). 
Similarly, `multcomp`-tidiers now include a call to `summary()` as previous 
behavior was akin to setting the now removed argument `quick = TRUE`. Both 
families of tidiers now use the `adj.p.value` column name when appropriate. 
Finally, `emmeans`-, `multcomp`-, and `TukeyHSD`-tidiers now consistently 
use the column names `contrast` and `null.value` instead 
of `comparison`, `level1` and `level2`, or `lhs` and `rhs`.

This release of broom also deprecates several helper functions as well as
tidier methods for a number of non-model objects, each in favor of more
principled approaches from other packages (outlined in the NEWS file). Notably,
though, tidiers have been deprecated for data frames, rowwise data frames,
vectors, and matrices. Further, we have moved forward with the planned
transfer of tidiers for mixed models to `broom.mixed`.

### Other Changes

Most all unit testing for the package is now supported by the
[modeltests](https://github.com/alexpghayes/modeltests) package!

Also, we have revised several vignettes and moved them to the tidymodels 
website. For backward compatibility, the existing vignettes will now simply 
link to the revised versions.

Finally, the package's website has moved from its previous
tidyverse domain to [broom.tidymodels.org](https://broom.tidymodels.org/).

### Looking Forward

Most notably, **the broom dev team generally will no longer be adding new tidier 
methods to the package.** Instead, we ask that issues/PRs requesting support for 
new model objects be directed to the model-owning package (i.e. the package 
that the model is exported from) rather than to broom. 

We are actively working to provide resources to both ease the process of 
writing new tidiers methods and reduce the dependency burden of taking 
on broom generics and helpers. As for the first point, we recently posted
an [article](https://www.tidymodels.org/learn/develop/broom/) on the
tidymodels website providing notes on best practices for writing tidiers.
This article will be kept up to date as we develop new resources for
easing the process of writing new tidier methods. As for the latter, 
the [`r-lib/generics`](https://github.com/r-lib/generics)
package provides lightweight dependencies for the main broom generics. We hope
to soon provide a coherent suite of helper functions for use in external
broom methods.

We anticipate that the most active development on the broom package, looking
forward, will center on `augment()` methods. 

### Contributors

This release features work and input from over 140 contributors (over 50 of them
for their first time) since the last major release. See the package
[release notes](https://broom.tidymodels.org/news/index.html) to see more 
specific notes on contributions. Thank you all for your thoughtful comments, 
patience, and hard work!

[&#x0040;abbylsmith](https://github.com/abbylsmith), 
[&#x0040;acoppock](https://github.com/acoppock), 
[&#x0040;ajb5d](https://github.com/ajb5d),  
[&#x0040;aloy](https://github.com/aloy), 
[&#x0040;AndrewKostandy](https://github.com/AndrewKostandy), 
[&#x0040;angusmoore](https://github.com/angusmoore), 
[&#x0040;anniew](https://github.com/anniew), 
[&#x0040;aperaltasantos](https://github.com/aperaltasantos), 
[&#x0040;asbates](https://github.com/asbates), 
[&#x0040;asondhi](https://github.com/asondhi), 
[&#x0040;asreece](https://github.com/asreece), 
[&#x0040;atyre2](https://github.com/atyre2), 
[&#x0040;bachmeil](https://github.com/bachmeil), 
[&#x0040;batpigandme](https://github.com/batpigandme), 
[&#x0040;bbolker](https://github.com/bbolker), 
[&#x0040;benjbuch](https://github.com/benjbuch), 
[&#x0040;bfgray3](https://github.com/bfgray3), 
[&#x0040;BibeFiu](https://github.com/BibeFiu), 
[&#x0040;billdenney](https://github.com/billdenney), 
[&#x0040;BrianOB](https://github.com/BrianOB), 
[&#x0040;briatte](https://github.com/briatte), 
[&#x0040;bruc](https://github.com/bruc), 
[&#x0040;brunaw](https://github.com/brunaw), 
[&#x0040;brunolucian](https://github.com/brunolucian), 
[&#x0040;bschneidr](https://github.com/bschneidr), 
[&#x0040;carlislerainey](https://github.com/carlislerainey), 
[&#x0040;CGMossa](https://github.com/CGMossa), 
[&#x0040;CharlesNaylor](https://github.com/CharlesNaylor), 
[&#x0040;ChuliangXiao](https://github.com/ChuliangXiao), 
[&#x0040;cimentadaj](https://github.com/cimentadaj), 
[&#x0040;crsh](https://github.com/crsh), 
[&#x0040;cwang23](https://github.com/cwang23), 
[&#x0040;DavisVaughan](https://github.com/DavisVaughan), 
[&#x0040;dchiu911](https://github.com/dchiu911), 
[&#x0040;ddsjoberg](https://github.com/ddsjoberg), 
[&#x0040;dgrtwo](https://github.com/dgrtwo), 
[&#x0040;dmenne](https://github.com/dmenne), 
[&#x0040;dylanjm](https://github.com/dylanjm), 
[&#x0040;ecohen13](https://github.com/ecohen13), 
[&#x0040;economer](https://github.com/economer), 
[&#x0040;EDiLD](https://github.com/EDiLD), 
[&#x0040;ekatko1](https://github.com/ekatko1), 
[&#x0040;ellessenne](https://github.com/ellessenne), 
[&#x0040;ethchr](https://github.com/ethchr), 
[&#x0040;florencevdubois](https://github.com/florencevdubois), 
[&#x0040;GegznaV](https://github.com/GegznaV), 
[&#x0040;gershomtripp](https://github.com/gershomtripp), 
[&#x0040;grantmcdermott](https://github.com/grantmcdermott), 
[&#x0040;gregmacfarlane](https://github.com/gregmacfarlane), 
[&#x0040;hadley](https://github.com/hadley), 
[&#x0040;haozhu233](https://github.com/haozhu233), 
[&#x0040;hasenbratan](https://github.com/hasenbratan), 
[&#x0040;HenrikBengtsson](https://github.com/HenrikBengtsson), 
[&#x0040;hermandr](https://github.com/hermandr), 
[&#x0040;hideaki](https://github.com/hideaki), 
[&#x0040;hughjonesd](https://github.com/hughjonesd), 
[&#x0040;iago-pssjd](https://github.com/iago-pssjd), 
[&#x0040;ifellows](https://github.com/ifellows), 
[&#x0040;IndrajeetPatil](https://github.com/IndrajeetPatil), 
[&#x0040;Inferrator](https://github.com/Inferrator), 
[&#x0040;istvan60](https://github.com/istvan60), 
[&#x0040;jamesmartherus](https://github.com/jamesmartherus), 
[&#x0040;JanLauGe](https://github.com/JanLauGe), 
[&#x0040;jasonyang5](https://github.com/jasonyang5), 
[&#x0040;jaspercooper](https://github.com/jaspercooper), 
[&#x0040;jcfisher](https://github.com/jcfisher), 
[&#x0040;jennybc](https://github.com/jennybc), 
[&#x0040;jessecambon](https://github.com/jessecambon), 
[&#x0040;jkylearmstrongibx](https://github.com/jkylearmstrongibx), 
[&#x0040;jmuhlenkamp](https://github.com/jmuhlenkamp), 
[&#x0040;JulianMutz](https://github.com/JulianMutz), 
[&#x0040;Jungpin](https://github.com/Jungpin), 
[&#x0040;jwilber](https://github.com/jwilber), 
[&#x0040;jyuu](https://github.com/jyuu), 
[&#x0040;karissawhiting](https://github.com/karissawhiting), 
[&#x0040;karldw](https://github.com/karldw), 
[&#x0040;khailper](https://github.com/khailper), 
[&#x0040;krauskae](https://github.com/krauskae), 
[&#x0040;kuriwaki](https://github.com/kuriwaki), 
[&#x0040;kyusque](https://github.com/kyusque), 
[&#x0040;KZARCA](https://github.com/KZARCA), 
[&#x0040;Laura-O](https://github.com/Laura-O), 
[&#x0040;ldlpdx](https://github.com/ldlpdx), 
[&#x0040;ldmahoney](https://github.com/ldmahoney), 
[&#x0040;lilymedina](https://github.com/lilymedina), 
[&#x0040;llendway](https://github.com/llendway), 
[&#x0040;lrose1](https://github.com/lrose1), 
[&#x0040;ltobalina](https://github.com/ltobalina), 
[&#x0040;LukasWallrich](https://github.com/LukasWallrich), 
[&#x0040;lukesonnet](https://github.com/lukesonnet), 
[&#x0040;lwjohnst86](https://github.com/lwjohnst86), 
[&#x0040;malcolmbarrett](https://github.com/malcolmbarrett), 
[&#x0040;margarethannum](https://github.com/margarethannum), 
[&#x0040;mariusbarth](https://github.com/mariusbarth),
[&#x0040;MatthieuStigler](https://github.com/MatthieuStigler),
[&#x0040;mattle24](https://github.com/mattle24), 
[&#x0040;mattpollock](https://github.com/mattpollock), 
[&#x0040;mattwarkentin](https://github.com/mattwarkentin), [&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), 
[&#x0040;mkirzon](https://github.com/mkirzon), 
[&#x0040;mlaviolet](https://github.com/mlaviolet), 
[&#x0040;Move87](https://github.com/Move87),
[&#x0040;namarkus](https://github.com/namarkus),
[&#x0040;nlubock](https://github.com/nlubock),
[&#x0040;nmjakobsen](https://github.com/nmjakobsen),
[&#x0040;ns-1m](https://github.com/ns-1m),
[&#x0040;nt-williams](https://github.com/nt-williams),
[&#x0040;oij11](https://github.com/oij11), 
[&#x0040;petrhrobar](https://github.com/petrhrobar),
[&#x0040;PirateGrunt](https://github.com/PirateGrunt),
[&#x0040;pjpaulpj](https://github.com/pjpaulpj),
[&#x0040;pkq](https://github.com/pkq),
[&#x0040;poppymiller](https://github.com/poppymiller), 
[&#x0040;QuLogic](https://github.com/QuLogic), 
[&#x0040;randomgambit](https://github.com/randomgambit),
[&#x0040;riinuots](https://github.com/riinuots), 
[&#x0040;RobertoMuriel](https://github.com/RobertoMuriel),
[&#x0040;Roisin-White](https://github.com/Roisin-White), 
[&#x0040;romainfrancois](https://github.com/romainfrancois),
[&#x0040;rsbivand](https://github.com/rsbivand), 
[&#x0040;serina-robinson](https://github.com/serina-robinson), 
[&#x0040;shabbybanks](https://github.com/shabbybanks),
[&#x0040;Silver-Fang](https://github.com/Silver-Fang), 
[&#x0040;Sim19](https://github.com/Sim19),
[&#x0040;simonpcouch](https://github.com/simonpcouch), 
[&#x0040;sjackson1236](https://github.com/sjackson1236), 
[&#x0040;softloud](https://github.com/softloud), 
[&#x0040;stefvanbuuren](https://github.com/stefvanbuuren), 
[&#x0040;strengejacke](https://github.com/strengejacke), [&#x0040;sushmitavgopalan16](https://github.com/sushmitavgopalan16), 
[&#x0040;tcuongd](https://github.com/tcuongd), 
[&#x0040;thisisnic](https://github.com/thisisnic),
[&#x0040;topepo](https://github.com/topepo), 
[&#x0040;tyluRp](https://github.com/tyluRp), 
[&#x0040;vincentarelbundock](https://github.com/vincentarelbundock), 
[&#x0040;vjcitn](https://github.com/vjcitn), 
[&#x0040;vnijs](https://github.com/vnijs), 
[&#x0040;weiyangtham](https://github.com/weiyangtham), 
[&#x0040;william3031](https://github.com/william3031), 
[&#x0040;x249wang](https://github.com/x249wang), 
[&#x0040;xieguagua](https://github.com/xieguagua), 
[&#x0040;yrosseel](https://github.com/yrosseel), 
and [&#x0040;zoews](https://github.com/zoews)
