---
output: hugodown::hugo_document
slug: clock-0-1-0
title: clock 0.1.0
date: 2021-02-26
author: Davis Vaughan
description: >
    Introducing, clock, a new package for working with date-times.
photo:
  
  url: https://unsplash.com/photos/FlHdnPO6dlw
  author: Jon Tyson
categories: [package] 
tags: ["r-lib"]
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE}
# Fix the TZ to America/New_York, as this is used by `as.POSIXct(<Date>)`.
# This is reset at the end of the Rmd with `withr::deferred_run()`.
withr::local_envvar(TZ = "America/New_York")
```

We're thrilled to announce the first release of [clock](https://r-lib.github.io/clock/index.html). clock is a new package providing a comprehensive set of tools for working with date-times. It is packed with features, including tools for: parsing, formatting, arithmetic, rounding, and extraction/updating of individual components. Additionally, it:

-   Provides a new family of orthogonal date-time classes (durations,
    time points, zoned-times, and calendars) that partition
    responsibilities so that you only have to think about time zones
    when you need them.

-   Implements a [high level API](https://r-lib.github.io/clock/reference/index.html#section-high-level-api) for Date and POSIXct classes that lets
    you get productive quickly without having to learn the details of
    the underlying classes.

-   Requires explicit handling of invalid dates (e.g. what date is one
    month after January 31st?) and nonexistent or ambiguous times
    (caused by daylight saving time issues).

-   Is built on the C++ [date](https://github.com/HowardHinnant/date)
    library, which provides a correct and high-performance backend.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("clock")
```

This blog post will show off a few of clock's unique features. To learn more, you'll want to take a look at clock's vignettes:

- [Getting Started](https://r-lib.github.io/clock/articles/clock.html)

- [Motivations for clock](https://r-lib.github.io/clock/articles/articles/motivations.html)

- [Examples and Recipes](https://r-lib.github.io/clock/articles/recipes.html)

- [Frequently Asked Questions](https://r-lib.github.io/clock/articles/faq.html)

```{r setup, message=FALSE, warning=FALSE}
library(clock)
library(nycflights13)
library(tidyverse)
library(lubridate)
```

## What about lubridate?

If you've ever worked with dates or date-times in R, you've probably used [lubridate](https://lubridate.tidyverse.org/). lubridate has powerful capabilities for working with this kind of data. So, why clock?

One of the primary motivations for creating clock was to improve on lubridate's handling of invalid dates and daylight saving time. As you'll see in the following sections, clock tries extremely hard to guard you from unexpected problems that can arise from these two complex concepts.

Additionally, clock provides a variety of new types for working with date-times. While lubridate is solely focused on working with R's native Date and POSIXct classes, clock goes many steps further with types such as: date-times without an implied time zone, nanosecond precision date-times, built-in granular types such as year-month and year-quarter, and a type for representing a weekday.

lubridate will never go away, and is not being deprecated or superseded. As of now, we consider clock to be an _alternative_ to lubridate. You can stick with one or the other, or use them together, as there are no name conflicts between the two. Keep in mind that clock is a young package, with plenty of room to grow. If you have any feedback about clock, or questions about its design, we'd love for you to [open an issue](https://github.com/r-lib/clock/issues).

## First steps

The best place to start learning about clock is by checking out the [High-Level API](https://r-lib.github.io/clock/reference/index.html#section-high-level-api). This lists all of the utilities in clock that work with R's native date (Date) and date-time (POSIXct) types. You'll notice that all of these helpers start with one of the following prefixes:

- `get_*()`: Get a component

- `set_*()`: Set a component

- `add_*()`: Add a unit of time

- `date_*()`: General date manipulation

We'll explore some of these with a trimmed down version of the `flights` dataset from the nycflights13 package.

```{r, echo=FALSE}
set.seed(1234)

flights <- nycflights13::flights

flights <- select(flights, year, month, day, dep_time, dep_delay)

magic_rows <- c(24287, 25177, 142797)

flights <- flights %>%
  slice_sample(n = 97) %>%
  bind_rows(slice(flights, magic_rows)) %>%
  arrange(year, month, day)
```

```{r}
flights
```

The flight departure date is separated into year, month, and day fields. We can combine these together into a Date with `date_build()`.

```{r}
flights <- flights %>%
  mutate(
    date = date_build(year, month, day), 
    .keep = "unused", 
    .before = 1
  )

flights
```

If you need to get those individual components back, extract them with the corresponding `get_*()` function.

```{r}
mutate(flights, year = get_year(date), month = get_month(date))
```

To summarize the average departure delay by month, one option is to use `date_group()` to group by the current month of the year. For Dates, this ends up setting every day of the month to `1`.

```{r}
flights %>%
  mutate(date = date_group(date, "month")) %>%
  group_by(date) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop")
```

If you instead wanted to summarize by week, starting on Wednesdays, you could use `date_shift()` to shift your date to the previous or next target weekday. This requires using one of clock's new types, weekday, to set the target to shift to.

```{r}
wed <- weekday(clock_weekdays$wednesday)
wed

flights %>%
  select(date) %>%
  mutate(
    shifted = date_shift(date, wed, which = "previous"),
    wday_date = as_weekday(date),
    wday_shifted = as_weekday(shifted)
  )
```

If you've used lubridate before, you would probably have done both of these summarizations with `lubridate::floor_date()`. In clock, date summarization is broken into three groups: grouping, shifting, and rounding. This separation leads to code that is both less surprising, and more powerful, giving you the ability to summarize in new ways, such as: flooring by multiple weeks, grouping by day of the quarter, and flooring by rolling sets of, say, 60 days.

Be sure to check out the many other high-level tools for working with dates, including powerful utilities for formatting ([`date_format()`](https://r-lib.github.io/clock/reference/date_format.html)) and parsing ([`date_parse()`](https://r-lib.github.io/clock/reference/date_parse.html) and [`date_time_parse()`](https://r-lib.github.io/clock/reference/date-time-parse.html)).

As a lubridate user, none of the above should seem particularly revolutionary, and that's the entire idea of the high-level API. We've tried to make transitioning over to clock as easy as possible. In the following sections, you'll see some of the benefits you'll get from doing so.

## Invalid dates

Using our `flights` data, imagine we want to add 1 month to `date`, perhaps to set up some kind of forward looking variable. With lubridate, we can use `+ months(1)`.

```{r}
mutate(flights, date2 = date + months(1))
```

Huh, what's up with those `NA` values? Let's try with clock:

```{r, error=TRUE}
mutate(flights, date2 = add_months(date, 1))
```

What's this about an invalid date? Location 5? Taking a closer look, we can see that adding 1 month to `2013-01-29` theoretically results in `2013-02-29`, which doesn't exist. In clock, this is known as an _invalid date_. With lubridate, invalid dates result in a silent `NA`. With clock, an error is raised.

So, how do we handle this? Well, there are a number of things that you could do:

- Return `NA`

- Return the previous valid moment in time

- Return the next valid moment in time

- Overflow our invalid date into March by the number of days past the true end of February that it landed at

With lubridate, you can use `add_with_rollback()` (i.e. `%m+%`) to help with the second and third bullet points.

```{r}
problems <- flights %>%
  select(date) %>%
  slice(c(5, 6))

problems %>%
  mutate(
    date2 = date + months(1),
    date3 = date %m+% months(1),
    date4 = add_with_rollback(date, months(1), roll_to_first = TRUE)
  )
```

With Dates, the hardest part about `%m+%` is remembering to use it. It is a common bug to forget to use this helper until *after* you have been bitten by an invalid date issue with an unexpected `NA`.

The overarching goal of clock is to protect you, the hard-working data scientist who shouldn't have to be an expert in date-times, from issues like this by erroring early and often, rather than letting this slip through unnoticed, only to cause hard to debug issues down the line.

Looking at the other half of clock's error message, you're advised to use the `invalid` argument to `add_months()`. This allows you to specify one of many invalid date resolution strategies.

```{r}
problems %>%
  mutate(
    date2 = add_months(date, 1, invalid = "previous"),
    date3 = add_months(date, 1, invalid = "next"),
    date4 = add_months(date, 1, invalid = "overflow"),
    date5 = add_months(date, 1, invalid = "NA")
  )
```

## Daylight saving time

The `dep_time` column of `flights` contains the hour and minute of the actual departure time, encoded together into a single integer. Let's extract that.

```{r}
flights_hm <- flights %>%
  select(date, dep_time) %>%
  mutate(
    hour = dep_time %/% 100L,
    minute = dep_time %% 100L,
    .keep = "unused"
  )

head(flights_hm, n = 2)
```

We'd like to be able to add this time of day information to our `date` column. This flight information was recorded in the America/New_York time zone, so our resulting date-time should have that time zone as well. However, converting Date -> POSIXct will _always_ assume that Date starts as UTC, rather than being naive to any time zones, and the result will use your system's local time zone. This can have unintended side effects:

```{r}
# My local time zone is actually America/New_York.
# The conversion to POSIXct retains the underlying UTC instant, but
# the printed time changes unexpectedly, showing the equivalent time
# in the local time zone.
flights_hm %>%
  select(date) %>%
  mutate(
    datetime = as.POSIXct(date),
    datetime_utc = date_set_zone(datetime, "UTC")
  ) %>%
  head(n = 3)
```

To get what we want, we need to convince the `date` column to "forget" that it is UTC, then add on the America/New_York time zone. We'll do this by going through a new intermediate type called naive-time, a date-time type with a yet-to-be-specified time zone. We'll talk more about naive-times in the Time Points section below. This retains the printed time as we expected.

```{r}
flights_dt <- flights_hm %>%
  mutate(
    datetime = as.POSIXct(as_naive_time(date), "America/New_York"),
    .keep = "unused",
    .before = 1
  )

flights_dt
```

We can now add on our hours and minutes.

```{r}
flights_dt <- flights_dt %>%
  mutate(
    datetime = datetime %>%
      add_hours(hour) %>%
      add_minutes(minute),
    .keep = "unused"
  )

flights_dt
```

Now assume that we want to add two days to this `datetime` column, again to construct some forward looking variable.

```{r}
flights_dt_lubridate <- flights_dt %>%
  mutate(datetime2 = datetime + days(2))

flights_dt_lubridate
```

Looks reasonable. Now with clock:

```{r, error=TRUE}
flights_dt %>%
  mutate(datetime2 = add_days(datetime, 2))
```

Another problem! This time a _nonexistent time_ at row 18. Let's investigate what lubridate gave us here:

```{r}
flights_dt_lubridate[18,]
```

An `NA`? But why?

As it turns out, in the America/New_York time zone, on 2013-03-10 the clocks jumped forward from 01:59:59 -> 03:00:00, creating a daylight saving time gap, and a _nonexistent_ 2 o'clock hour. By adding two days, we've landed right in that gap (at 02:23:00). With nonexistent times like this, lubridate silently returns `NA`, while clock errors.

Like with invalid dates, clock tries to guard you from these issues by erroring as soon as they occur. You can resolve these particular issues with the `nonexistent` argument to `add_days()`. In this case, we could:

- Roll forward to the next valid moment in time

- Roll backward to the previous valid moment in time

- Shift forward by the size of the gap

- Shift backward by the size of the gap

- Return `NA`

```{r}
problem <- flights_dt$datetime[18]
problem

# 02:23:00 -> 03:00:00
add_days(problem, 2, nonexistent = "roll-forward")

# 02:23:00 -> 01:59:59
add_days(problem, 2, nonexistent = "roll-backward")

# 02:23:00 -> 03:23:00
add_days(problem, 2, nonexistent = "shift-forward")

# 02:23:00 -> 01:23:00
add_days(problem, 2, nonexistent = "shift-backward")

# 02:23:00 -> NA
add_days(problem, 2, nonexistent = "NA")
```

I recommend `"roll-forward"` or `"roll-backward"`, as these retain the _relative ordering_ of `datetime`, an issue that you can read about [here](https://r-lib.github.io/clock/articles/articles/motivations.html#nonexistent-time-1).

Unlike with invalid dates, lubridate does not provide any tools for resolving nonexistent times.

There are another class of daylight saving time issues related to _ambiguous times_. You resolve them in a similar way to what was done with nonexistent times. If you're interested, you can read more about those ambiguous times [here](https://r-lib.github.io/clock/articles/articles/motivations.html#ambiguous-time-1).

## Production

This new invalid date and daylight saving time behavior might sound great to you, but you might be wondering about usage of clock in production. What happens if `add_months()` worked in interactive development, but then you put your analysis into production, gathered new data, and all of the sudden it started failing?

```{r, error=TRUE}
dates <- flights$date

# All good! Ship it!
add_months(dates[1:4], 1) 

# Failed in production with new data! Oh no!
add_months(dates[1:10], 1)
```

To balance the usefulness of clock in interactive development with the strict requirements of production, you can set the `clock.strict` global option to `TRUE` to turn `invalid`, `nonexistent`, and `ambiguous` from optional arguments into required ones.

```{r, error=TRUE}
rlang::with_options(clock.strict = TRUE, .expr = {
  add_months(dates[1:4], 1)
})
```

Forcing yourself to specify these arguments up front during interactive development is a great way to explicitly document your assumptions about these possible issues, while also guarding against future problems in production.

## Advanced features

This blog post has only scratched the surface of what clock can do. Up until now, we've only explored clock's high-level API. There exists an entire world of more powerful utilities in the low-level API that powers clock. We'll briefly explore a few of those in the next few sections, but I'd encourage checking out the rest of the [reference page](https://r-lib.github.io/clock/reference/index.html) to get a bird's-eye view of all that clock can do.

### Calendars

Calendars allow you to represent a date using an alternative format. Rather than using a typical year, month, and day of the month format, you might want to specify the fiscal year, quarter, and day of the quarter. In the end, these point to the same moment in time, just in different ways. For example:

```{r}
ymd <- year_month_day(2019, 2, 25)

# Fiscal year starting in January
as_year_quarter_day(ymd)

# Fiscal year starting in April
as_year_quarter_day(ymd, start = clock_months$april)
```

There are 5 calendars that come with clock. The neat part about these is that they have _varying precision_, from year to nanosecond. This provides built-in granular types like year-month and year-quarter.

```{r}
# Gregorian year, month, and day of the month
year_month_day(2019, 1, 14)
year_month_day(2019, 2)
year_month_day(2019, 2, 14, 2, 30, 25, 12345, subsecond_precision = "nanosecond")

# Gregorian year, month, and indexed weekday of the month
# (i.e. the 2nd Wednesday)
year_month_weekday(2019, 2, day = clock_weekdays$wednesday, index = 2)

# Gregorian year and day of the year
year_day(2019, 105)

# Fiscal year, quarter, and day of the quarter
year_quarter_day(2019, 1, 14)
year_quarter_day(2019, 1, 14, start = clock_months$april)
year_quarter_day(2019, 2:4)

# ISO year, week, and day of the week
iso_year_week_day(2019, 2, clock_iso_weekdays$friday)
```

As shown above, you can convert from one calendar to another with functions like `as_year_quarter_day()`, and to Date or POSIXct with the standard `as.Date()` and `as.POSIXct()` functions.

One of the most unique features of calendars is the ability to represent invalid dates directly. In a previous section, we added 1 month to a Date and used the `invalid` argument to resolve invalid date issues. Let's swap to a year-month-day and try again. We can also use the cleaner `+ duration_months()` syntax here, which we can't use with Dates.

```{r}
invalids <- flights %>%
  select(date) %>%
  mutate(
    ymd = as_year_month_day(date),
    ymd2 = ymd + duration_months(1)
  )

invalids
```

The `ymd2` column directly contains the invalid dates, `2013-02-29` and `2013-02-30`! You also get a note up in the header, `i=2`, letting you know that you have two invalid dates in that column. You can resolve these dates at any time using `invalid_resolve()`, providing an invalid resolution strategy like we did earlier. Or, you can ignore them if you expect them to be resolved naturally in some other way. For example, if our end goal was to add 1 month, then fix the day of the month to the 15th, then these invalid dates would naturally resolve themselves:

```{r}
mutate(invalids, ymd3 = set_day(ymd2, 15))
```

The important thing is that they _eventually_ get resolved. You must resolve invalid dates before converting to another calendar or to a Date / POSIXct.

```{r, error=TRUE}
as.Date(invalids$ymd2)
```

### Time points and Zoned-times

The daylight saving time section of this post was complicated by the need to work around time zones. If your analysis doesn't actually require time zones, you can represent a date or date-time using a _naive-time_. This date-time type makes no assumption about the current time zone, instead assuming that there is a yet-to-be-specified time zone that hasn't been declared yet. Naive-times vary in precision from day up to nanosecond.

```{r}
flights_nt <- flights_hm %>%
  mutate(
    naive_day = as_naive_time(date),
    naive_time = naive_day + duration_hours(hour) + duration_minutes(minute)
  ) %>%
  select(date, starts_with("naive"))

flights_nt
```

Going from Date -> naive-time has dropped the UTC time zone assumption altogether, while keeping the printed time. This allowed us to convert back to POSIXct in an earlier example. Essentially, all that we were doing was declaring that yet-to-be-specified time zone as America/New_York, keeping the printed time where possible. We could have easily chosen a different time zone, like Europe/London.

```{r}
flights_nt %>%
  select(naive_time) %>%
  mutate(
    datetime_ny = as.POSIXct(naive_time, "America/New_York"),
    datetime_lo = as.POSIXct(naive_time, "Europe/London")
  )
```

If you're used to lubridate, converting to naive-time and back with a different time zone is similar to using `lubridate::force_tz()`, but with more control over possible daylight saving time issues (again using `nonexistent` and `ambiguous`, but supplied directly to `as.POSIXct()`).

In clock, a naive-time is a particular kind of _time point_, a type that counts units of time with respect to some origin. Time points are extremely efficient at daily and sub-daily arithmetic, but calendars are better suited for monthly and yearly arithmetic. Time points are also efficient at _rounding_ and _shifting_, through `time_point_floor()` and `time_point_shift()`, but calendars are better at _grouping_, through `calendar_group()`. In the high-level API for Date and POSIXct, we gloss over these details and internally switch between these two types for you.

There is a second type of time point in clock, the sys-time, which works exactly like a naive-time _except_ that it is assumed to be in UTC. If you never use a time zone aware class like POSIXct, then sys-time and naive-time are equivalent. However, once you start adding in time zones, the way you interpret each of them becomes extremely important.

```{r}
ymd <- year_month_day(2019, 1, 1)

# Yet-to-be-specified time zone
naive <- as_naive_time(ymd)
naive

# UTC time zone
sys <- as_sys_time(ymd)
sys

# - Keeps printed time
# - Changes underlying duration
as.POSIXct(naive, "America/New_York")

# - Changes printed time
# - Keeps underlying duration
as.POSIXct(sys, "America/New_York")
```

clock also provides its own time zone aware date-time type, the zoned-time. Converting to a zoned-time from a sys-time or naive-time works the same as converting to a POSIXct, but zoned-times can have up to nanosecond precision.

```{r}
naive %>%
  add_nanoseconds(100) %>%
  add_hours(2) %>%
  as_zoned_time("America/New_York")
```

There isn't actually a lot you can do with zoned-times directly. Generally, zoned-times are the start or end points of an analysis meant for humans to interpret. In the middle, you'll convert to naive-time, sys-time, or to a calendar type to perform any date-time specific manipulations.

```{r, echo=FALSE}
# Reset `TZ` envvar
withr::deferred_run()
```
