---
output: hugodown::hugo_document

slug: httr2-1-2-0
title: httr2 1.2.0
date: 2025-07-14
author: Hadley Wickham
description: >
    httr2 1.2.0 improves security for redacted headers, improves URL parsing
    and building, enhances debugging, and includes a bunch of other quality
    of life improvements.

photo:
  url: https://chatgpt.com/share/6870349d-20cc-8009-84b0-dd026c75cbb2
  author: chatGPT

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [httr2]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

# httr2 1.2.0

We're delighted to announce the release of httr2 1.2.0! httr2 (pronounced "hitter2") is a comprehensive HTTP client that provides a modern, pipeable API for working with web APIs. It builds on top of {curl} to provide features like explicit request objects, built-in rate limiting & retry tooling, comprehensive OAuth support, and secure handling of secrets and credentials.

You can install it from CRAN with:

```r
install.packages("httr2")
```

This blog post will walk you through the most important changes in 1.2.0: lifecycle updates, improved security for redacted headers, URL handling improvements, improved debugging tools, and a handful of other quality of life improvements. You can see a full list of changes in the [release notes](https://github.com/r-lib/httr2/releases/tag/v1.2.0)

```{r}
library(httr2)
```

## Lifecycle changes

Part of httr2's continued evolution is phasing out features that we now believe were mistakes. In this release:

* `req_perform_stream()` has been soft deprecated (not just superseded) in favour of `req_perform_connection()`.

* Deprecated functions `multi_req_perform()`, `req_stream()`, `with_mock()`, and `local_mock()`  have now been removed. Please use their modern replacements, `req_perform_parallel()`, `req_perform_stream()`, `with_mocked_responses()`, and `local_mocked_responses()`, instead.

* Deprecated arguments `req_perform_parallel(pool)`, 
  `req_oauth_auth_code(host_name, host_ip, port)`, and 
  `oauth_flow_auth_code(host_name, host_ip, port)` have been removed.
  Please use `req_perform_parallel(max_active)` and `req_oauth_auth_code(redirect_url)`/
  `oauth_flow_auth_code(redirect_url)` instead.

## Enhanced security for redacted headers

One of the most important improvements in this release improves the security of redacted headers. Redacted headers are used to conceal secrets, like API keys or passwords, that you don't want to accidentally reveal. For a long time, httr2 has automatically hidden these headers when you `print()` or `str()` them, ensuring that they don't accidentally end up in log files. You can see this in action with the `Authorization` header, which httr2 now automatically redacts:

```{r}
req <- request("http://example.com") |>
  req_auth_basic("username", "password")

req
str(req$headers)
req |> req_dry_run()
```

(If for you do really need to see the redacted values you can get with a bit of extra effort: call the new `req_get_headers()` function with `redacted = "reveal"`.)

In httr2 1.2.0, we've gone one step further, and prevented redacted headers from being saved to disk. Now if you save and reload a request, you'll notice that the redacted headers are no longer present:

```{r}
path <- tempfile()
saveRDS(req, path)

req2 <- readRDS(path)
req2 |> req_dry_run()
```

This protects you from accidentally revealing your credentials if you save a request to disk. This is easier to do than you might expect because httr2 includes the request object in every response (since this makes debugging much easier). That means if you're caching a slow response, it's very easy to accidentally store a secret, potentially leaking secure values. (Don't ask me how I discovered this!)

## URL handling improvements

URL construction is now powered by `curl::curl_modify_url()`, which correctly escapes the path component: 

```{r}
req <- request("https://api.example.com")
req |> req_url_path("/users/john doe/profile") |> req_get_url()
```

This means that `req_url_path()` can now only affect the path component of the URL, not the query parameters. If you previously relied on this behaviour, you'll need to switch to `req_url_query()`:

```{r}
# won't work any more:
req |> 
  req_url_path("/users?name=john-doe") |> 
  req_get_url()
# so now do this:
req |> 
  req_url_path("/users") |> 
  req_url_query(name = "john-doe") |> 
  req_get_url()
```

## Improved debugging tools

The vast majority of modern APIs use JSON, so httr2 now includes a few features to make debugging those APIs a little easier:

* `last_request()` and `last_response()` are now paired with `last_request_json()` and `last_response_json()` which pretty-print the JSON bodies of the last request and response. 
* `req_dry_run()` and `req_verbose()` automatically pretty print JSON bodies (turn this off by setting `options(httr2_pretty_json = FALSE)`).

We've also included a few general tools to make it easier to control httr2's default verbosity. You can now control the default via the `HTTR2_VERBOSITY` environment variable and there's a new `local_verbosity()` function to match the existing `with_verbosity()`.

## Quality of life improvements

This release also includes a bunch of few smaller quality of life improvements:

* `req_perform_parallel()` now lifts many of its restrictions. It now supports simplified versions of `req_throttle()` and `req_retry()`, it can refresh OAuth tokens, and it checks the cache before each request. 

* `req_get_url()`, `req_get_method()`, `req_get_headers()`, `req_body_get_type()`, and `req_get_body()` allow you to introspect request objects.

* `req_throttle()` now uses a "[leaky bucket](https://en.wikipedia.org/wiki/Leaky_bucket)". This maintains the same average rate limit as before, while allowing bursts of higher rates.

* `resp_timing()` exposes detailed timing information for a response.

## Acknowledgements

A big thanks to everyone who contributed to this release through issues, pull requests, and discussions on GitHub: [&#x0040;Aariq](https://github.com/Aariq), [&#x0040;annalena13](https://github.com/annalena13), [&#x0040;apsteinmetz](https://github.com/apsteinmetz), [&#x0040;arcresu](https://github.com/arcresu), [&#x0040;arnaudgallou](https://github.com/arnaudgallou), [&#x0040;atheriel](https://github.com/atheriel), [&#x0040;DavidRLovell](https://github.com/DavidRLovell), [&#x0040;dfalbel](https://github.com/dfalbel), [&#x0040;Eli-Berkow](https://github.com/Eli-Berkow), [&#x0040;fwimp](https://github.com/fwimp), [&#x0040;hadley](https://github.com/hadley), [&#x0040;jansim](https://github.com/jansim), [&#x0040;jcheng5](https://github.com/jcheng5), [&#x0040;jeffreyzuber](https://github.com/jeffreyzuber), [&#x0040;jjesusfilho](https://github.com/jjesusfilho), [&#x0040;jonthegeek](https://github.com/jonthegeek), [&#x0040;Kevanness](https://github.com/Kevanness), [&#x0040;m-muecke](https://github.com/m-muecke), [&#x0040;maelle](https://github.com/maelle), [&#x0040;mayeulk](https://github.com/mayeulk), [&#x0040;mdsumner](https://github.com/mdsumner), [&#x0040;noamross](https://github.com/noamross), [&#x0040;omuelle](https://github.com/omuelle), [&#x0040;pedrobtz](https://github.com/pedrobtz), [&#x0040;plietar](https://github.com/plietar), [&#x0040;ramiromagno](https://github.com/ramiromagno), [&#x0040;salim-b](https://github.com/salim-b), [&#x0040;sckott](https://github.com/sckott), [&#x0040;shikokuchuo](https://github.com/shikokuchuo), [&#x0040;vibalre](https://github.com/vibalre), [&#x0040;vladimirobucina](https://github.com/vladimirobucina), and [&#x0040;ZheFrench](https://github.com/ZheFrench).
