---
output: hugodown::hugo_document

slug: tidyr-1-3-0
title: tidyr 1.3.0
date: 2023-01-23
author: Hadley Wickham
description: >
    tidyr 1.3.0 brings a new family of string separating functions,
    along with improvements to `unnest_longer()`, `unnest_wider()`,
    `pivot_longer()`, and `nest()`.

photo:
  url: https://unsplash.com/photos/TEDo1eO8te4
  author: Jan Kopřiva

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package] 
tags: [tidyr]
---

```{=html}
<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->
```
We're pleased to announce the release of [tidyr](https://tidyr.tidyverse.org) 1.3.0.
tidyr provides a set of tools for transforming data frames to and from tidy data, where each variable is a column and each observation is a row.
Tidy data is a convention for matching the semantics and structure of your data that makes using the rest of the tidyverse (and many other R packages) much easier.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("tidyr")
```

This post highlights the biggest changes in this release:

-   A new family of `separate_*()` functions supersede `separate()` and `extract()` and come with useful debugging features.

-   `unnest_wider()` and `unnest_longer()` gain a bundle of useful improvements.

-   `pivot_longer()` gets a new `cols_vary` argument.

-   `nest(.by)` provides a new (and hopefully final) way to create nested datasets.

You should also notice generally improved errors with this release: we check function arguments more aggressively and take care to always report the name of the function that you called, not some internal helper.
As usual, you can find a full set of changes in the [release notes](http://github.com/tidyverse/tidyr/releases/tag/v1.3.0).

```{r setup}
library(tidyr)
library(dplyr, warn.conflicts = FALSE)
```

## `separate_*()` family of functions

The biggest feature of this release is a new, experimental, family of functions for separating string columns:

|                                  | Make columns                | Make rows                    |
|--------------------------|-----------------------|-----------------------|
| Separate with delimiter          | `separate_wider_delim()`    | `separate_longer_delim()`    |
| Separate by position             | `separate_wider_position()` | `separate_longer_position()` |
| Separate with regular expression | `separate_wider_regex()`    |                              |

These functions collectively supersede `extract()`, `separate()`, and `separate_rows()` because they have more consistent names and arguments, have better performance (thanks to stringr), and provide a new approach for handling problems.

|                                  | Make columns                     | Make rows         |
|---------------------------|---------------------------|------------------|
| Separate with delimiter          | `separate(sep = string)`         | `separate_rows()` |
| Separate by position             | `separate(sep = integer vector)` | N/A               |
| Separate with regular expression | `extract()`                      |                   |

Here I'll focus on the `wider` functions because they generally present the most interesting challenges.
Let's start by grabbing some census data with the [tidycensus](https://walker-data.com/tidycensus/) package:

```{r}
vt_census <- tidycensus::get_decennial(
  geography = "block",
  state = "VT",
  county = "Washington",
  variables = "P1_001N",
  year = 2020
)
vt_census
```

The `GEOID` column is made up of four components: a 2 digit state identifier, a 3 digit county identifier, a 6 digit tract identifier, and a 4 digit block identifier.
We could use `separate_wider_position()` to extract these into their own variables:

```{r}
vt_census |>
  select(GEOID) |> 
  separate_wider_position(
    GEOID,
    widths = c(state = 2, county = 3, tract = 6, block = 4)
  )
```

The `name` column contains this same information in a text form, with each component separated by a comma.
We can use `separate_wider_delim()` to break up this sort of data into individual variables:

```{r}
vt_census |>
  select(NAME) |> 
  separate_wider_delim(
    NAME,
    delim = ", ",
    names = c("block", "block_group", "tract", "county", "state")
  )
```

You'll notice that each row contains a lot of duplicated information ("Block", "Block Group", ...).
You could certainly use `mutate()` and string manipulation to clean this up, but there's a more direct approach that you can use if you're familiar with regular expressions.
The new `separate_wider_regex()` takes a vector of regular expressions that are matched in order, from left to right.
If you name the regular expression, it will appear in the output; otherwise, it will be dropped.
I think this leads to a particularly elegant solution to many problems.

```{r}
vt_census |>
  select(NAME) |> 
  separate_wider_regex(
    NAME,
    patterns = c(
      "Block ", block = "\\d+", ", ",
      "Block Group ", block_group = "\\d+", ", ",
      "Census Tract ", tract = "\\d+.\\d+", ", ",
      county = "[^,]+", ", ",
      state = ".*"
    )
  )
```

These functions also have a new way to report problems.
Let's start with a very simple example:

```{r}
#| error: true
df <- tibble(
  id = 1:3,
  x = c("a", "a-b", "a-b-c")
)

df |> separate_wider_delim(x, delim = "-", names = c("x", "y"))
```

We've requested two columns in the output (`x` and `y`), but the first row has only one element and the last row has three elements, so `separate_wider_delim()` can't do what we've asked.
The error lays out your options for resolving the problem using the `too_few` and `too_many` arguments.
I'd recommend always starting with `"debug"` to get more information about the problem:

```{r}
probs <- df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("a", "b"),
    too_few = "debug",
    too_many = "debug"
  )
probs
```

This adds three new variables: `x_ok` tells you if the `x` could be separated as you requested, `x_pieces` tells you the actual number of pieces, and `x_remainder` shows you anything that remains after the columns you asked for.
You can use this information to fix the problems in the input, or you can use the other options to `too_few` and `too_many` to tell `separate_wider_delim()` to fix them for you:

```{r}
df |> 
  separate_wider_delim(
    x,
    delim = "-",
    names = c("a", "b"),
    too_few = "align_start",
    too_many = "drop"
  )
```

`too_few` and `too_many` also work with `separate_wider_position()`, and `too_few` works with `separate_wider_regex()`.
The longer variants don't need these arguments because varying numbers of rows don't matter in the same way that varying numbers of columns do:

```{r}
df |> separate_longer_delim(x, delim = "-")
```

These functions are still experimental so we are actively seeking feedback.
Please try them out and let us know if you find them useful or if there are other features you'd like to see.

## `unnest_wider()` and `unnest_longer()` improvements

`unnest_longer()` and `unnest_wider()` have both received some quality of life and consistency improvements.
Most importantly:

-   `unnest_wider()` now gives a better error when unnesting an unnamed vector:

    ```{r}
    #| error: true
    df <- tibble(
      id = 1:2,
      x = list(c("a", "b"), c("d", "e", "f"))
    )
    df |> 
      unnest_wider(x)

    df |> 
      unnest_wider(x, names_sep = "_")
    ```

    And this same behaviour now also applies to partially named vectors.

-   `unnest_longer()` has gained a `keep_empty` argument like `unnest()`, and it now treats `NULL` and empty vectors the same way:

    ```{r}
    df <- tibble(
      id = 1:3,
      x = list(NULL, integer(), 1:3)
    )

    df |> unnest_longer(x)
    df |> unnest_longer(x, keep_empty = TRUE)
    ```

## `pivot_longer(cols_vary)`

By default, `pivot_longer()` creates its output row-by-row:

```{r}
df <- tibble(
  x = 1:2,
  y = 3:4,
  z = 5:6
)

df |> 
  pivot_longer(
    everything(),
    names_to = "name",
    values_to = "value"
  )
```

You can now request to create the output column-by-column with `cols_vary = "slowest":`

```{r}
df |> 
  pivot_longer(
    everything(),
    names_to = "name",
    values_to = "value",
    cols_vary = "slowest"
  )
```

## `nest(.by)`

A nested data frame is a data frame where one (or more) columns is a list of data frames.
Nested data frames are a powerful tool that allow you to turn groups into rows and can facilitate certain types of data manipulation that would be very tricky otherwise.
(One place to learn more about them is my 2016 talk "[Managing many models with R](https://www.youtube.com/watch?v=rz3_FDVt9eg)".)

Over the years we've made a number of attempts at getting the correct interface for nesting, including `tidyr::nest()`, `dplyr::nest_by()`, and `dplyr::group_nest()`.
In this version of tidyr we've taken one more stab at it by adding a new argument to `nest()`: `.by`, inspired by the upcoming [dplyr 1.1.0](https://www.tidyverse.org/blog/2022/11/dplyr-1-1-0-is-coming-soon/) release.
This means that `nest()` now allows you to specify the variables you want to nest by as an alternative to specifying the variables that appear in the nested data.

```{r}
# Specify what to nest by
mtcars |> 
  nest(.by = cyl)

# Specify what should be nested
mtcars |> 
  nest(data = -cyl)

# Specify both (to drop variables)
mtcars |> 
  nest(data = mpg:drat, .by = cyl)
```

If this function is all we hope it to be, we're likely to supersede `dplyr::nest_by()` and `dplyr::group_nest()` in the future.
This has the nice property of placing the functions for nesting and unnesting in the same package (tidyr).

## Acknowledgements

A big thanks to all 51 contributors who helped make this release possible, by writing code and documentating, asking questions, and reporting bugs! [&#x0040;AdrianS85](https://github.com/AdrianS85), [&#x0040;ahcyip](https://github.com/ahcyip), [&#x0040;allenbaron](https://github.com/allenbaron), [&#x0040;AnBarbosaBr](https://github.com/AnBarbosaBr), [&#x0040;ArthurAndrews](https://github.com/ArthurAndrews), [&#x0040;bart1](https://github.com/bart1), [&#x0040;billdenney](https://github.com/billdenney), [&#x0040;bknakker](https://github.com/bknakker), [&#x0040;bwiernik](https://github.com/bwiernik), [&#x0040;crissthiandi](https://github.com/crissthiandi), [&#x0040;daattali](https://github.com/daattali), [&#x0040;DavisVaughan](https://github.com/DavisVaughan), [&#x0040;dcaud](https://github.com/dcaud), [&#x0040;DSLituiev](https://github.com/DSLituiev), [&#x0040;elgabbas](https://github.com/elgabbas), [&#x0040;fabiangehring](https://github.com/fabiangehring), [&#x0040;hadley](https://github.com/hadley), [&#x0040;ilikegitlab](https://github.com/ilikegitlab), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;jic007](https://github.com/jic007), [&#x0040;Joao-O-Santos](https://github.com/Joao-O-Santos), [&#x0040;joeycouse](https://github.com/joeycouse), [&#x0040;jonspring](https://github.com/jonspring), [&#x0040;kevinushey](https://github.com/kevinushey), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;lionel-](https://github.com/lionel-), [&#x0040;lotard](https://github.com/lotard), [&#x0040;lschneiderbauer](https://github.com/lschneiderbauer), [&#x0040;lucylgao](https://github.com/lucylgao), [&#x0040;markfairbanks](https://github.com/markfairbanks), [&#x0040;martina-starc](https://github.com/martina-starc), [&#x0040;MatthieuStigler](https://github.com/MatthieuStigler), [&#x0040;mattnolan001](https://github.com/mattnolan001), [&#x0040;mattroumaya](https://github.com/mattroumaya), [&#x0040;mdkrause](https://github.com/mdkrause), [&#x0040;mgirlich](https://github.com/mgirlich), [&#x0040;millermc38](https://github.com/millermc38), [&#x0040;modche](https://github.com/modche), [&#x0040;moodymudskipper](https://github.com/moodymudskipper), [&#x0040;mspittler](https://github.com/mspittler), [&#x0040;olivroy](https://github.com/olivroy), [&#x0040;piokol23](https://github.com/piokol23), [&#x0040;ppreshant](https://github.com/ppreshant), [&#x0040;ramiromagno](https://github.com/ramiromagno), [&#x0040;Rengervn](https://github.com/Rengervn), [&#x0040;rjake](https://github.com/rjake), [&#x0040;roohitk](https://github.com/roohitk), [&#x0040;struckma](https://github.com/struckma), [&#x0040;tjmahr](https://github.com/tjmahr), [&#x0040;weirichs](https://github.com/weirichs), and [&#x0040;wurli](https://github.com/wurli).
