---
output: hugodown::hugo_document
slug: dplyr-1-1-1
title: dplyr 1.1.1
date: 2023-03-22
author: Davis Vaughan
description: >
    dplyr 1.1.1 is on CRAN! This patch release includes a number of performance regression fixes along with refinements to the multiple match join warnings that result in warnings being thrown much less often.
photo:
  url: https://unsplash.com/photos/jy8z4NBIYSQ
  author: Jon Tyson
categories: [package] 
tags: [dplyr, dplyr-1-1-0]
editor_options: 
  chunk_output_type: console
---

We're stoked to announce the release of [dplyr 1.1.1](https://dplyr.tidyverse.org/).
This blog post will outline the two most important features of this release:

-   Addressing various performance regressions
-   Refining the `multiple` match warning thrown by dplyr's joins

You can see a full list of changes in the [release notes](https://dplyr.tidyverse.org/news/index.html).
To see the other blog posts in the dplyr 1.1.0 series, head [here](https://www.tidyverse.org/tags/dplyr-1-1-0/).

You can install dplyr 1.1.1 from CRAN with:

```{r, eval = FALSE}
install.packages("dplyr")
```

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
```

## Performance regressions

In the [1.1.0 post on vctrs](/blog/2023/02/dplyr-1-1-0-vctrs), we discussed that we've rewritten all of dplyr's vector functions on top of [vctrs](https://vctrs.r-lib.org/) for improved versatility.
Unfortunately, we accidentally made two sets of functions much slower, especially when used on a data frame with many groups:

-   `case_when()` and `if_else()`

-   `nth()`, `first()`, and `last()`

These performance issues have been addressed, and should be back to 1.0.10 level of performance.
`case_when()` is still *slightly* slower than 1.0.10, but it isn't likely to be very noticeable, and we already have plans to improve this further in a future release.

## Revisiting multiple matches

In the [1.1.0 post on joins](/blog/2023/01/dplyr-1-1-0-joins), we discussed the new `multiple` argument that was added to `left_join()` and friends, which had a built in safety check that warned when you performed a join where a row from `x` matched more than one row from `y`.
The TLDR of the discussion below is that we've realized that this warning was being thrown in too many cases, so we've adjusted it in such a way that it now only catches the most dangerous type of join (a many-to-many join), meaning that you should see the warning *much* less often.

As a reminder, `multiple` determines what happens when a row from `x` matches more than one row from `y`.
You can choose to return `"all"` of the matches, the `"first"` or `"last"` match, or `"any"` of the matches if you are just interested in detecting if there is at least one.
`multiple` defaulted to a behavior similar to `"all"`, with the added side effect of throwing a warning if multiple matches were actually detected, like this:

```{r}
student <- tibble(
  student_id = c(1, 2, 3),
  transfer = c(FALSE, TRUE, TRUE),
  initial_term = c("fall 2018", "fall 2020", "fall 2020")
)

term <- tibble(
  student_id = c(1, 1, 2, 3, 3, 3),
  term = c("fall 2018", "spring 2019", "fall 2020", "fall 2020", "spring 2021", "fall 2021"),
  course_load = c(12, 15, 10, 14, 15, 12)
)
```

```{r}
# Information about students attending a university.
# One row per (student_id).
student

# Term specific information about each student.
# One row per (student_id, term) combination.
term
```

```{r, eval=FALSE}
student |>
  left_join(term, join_by(student_id))
#> Warning in left_join(student, term, join_by(student_id)): Each row in `x` is expected to match at most 1 row in `y`.
#> i Row 1 of `x` matches multiple rows.
#> i If multiple matches are expected, set `multiple = "all"` to silence this warning.
#> # A tibble: 6 × 5
#>   student_id transfer initial_term term        course_load
#>        <dbl> <lgl>    <chr>        <chr>             <dbl>
#> 1          1 FALSE    fall 2018    fall 2018            12
#> 2          1 FALSE    fall 2018    spring 2019          15
#> 3          2 TRUE     fall 2020    fall 2020            10
#> 4          3 TRUE     fall 2020    fall 2020            14
#> 5          3 TRUE     fall 2020    spring 2021          15
#> 6          3 TRUE     fall 2020    fall 2021            12
```

To silence this warning, we encouraged you to set `multiple = "all"` to be explicit about the fact that you expected a row from `x` to match multiple rows in `y`.

The original motivation for this behavior comes from a two-part hypothesis of ours:

-   Users are often surprised when a join returns more rows than the left-hand table started with (in the above example, `student` has 3 rows but the join result has 6).

-   It is dangerous to allow joins that can result in a Cartesian explosion of the number of rows (i.e. `nrow(x) * nrow(y)`).

This hypothesis led us to automatically warn on two types of join relationships, one-to-many joins and many-to-many joins.
If you aren't familiar with these terms, here is a quick rundown of the 4 types of join relationships (often discussed in a SQL context), which provide constraints on the number of allowed matches:

-   one-to-one:
    -   A row from `x` can match at most 1 row from `y`.
    -   A row from `y` can match at most 1 row from `x`.
-   one-to-many:
    -   A row from `x` can match any number of rows in `y`.
    -   A row from `y` can match at most 1 row from `x`.
-   many-to-one:
    -   A row from `x` can match at most 1 row from `y`.
    -   A row from `y` can match any number of rows in `x`.
-   many-to-many:
    -   A row from `x` can match any number of rows in `y`.
    -   A row from `y` can match any number of rows in `x`.

After gathering some valuable [user feedback](https://github.com/tidyverse/dplyr/issues/6717) and conducting an [in depth analysis](https://github.com/tidyverse/dplyr/issues/6731) of these join relationships, we've determined that the only relationship style actually worth warning on is many-to-many, because that is the one that can result in a Cartesian explosion of rows.
In retrospect, the one-to-many relationship is actually quite common, and is symmetrical with many-to-one, which we weren't warning on.
You could actually exploit this fact by switching the above join around, which would silence the warning:

```{r}
term |>
  left_join(student, join_by(student_id))
```

We still believe that new users are often surprised when a join returns more rows than they originally started with, but the many-to-one case of this is rarely a problem in practice.
So, as of dplyr 1.1.1, we no longer warn on one-to-many relationships, which should drastically reduce the amount of warnings that you see.

### Many-to-many relationships

A many-to-many relationship is much harder to construct (which is good).
In fact, a database system won't even let you create one of these "relationships" between two tables directly, instead requiring you to create a third bridge table that turns the many-to-many relationship into two one-to-many relationships.
We can "accidentally" create one of these in R though:

```{r}
course <- tibble(
  student_id = c(1, 1, 1, 2, 2, 3, 3, 3, 3),
  instructor_id = c(1, 2, 3, 1, 2, 1, 2, 3, 4),
  course = c(101, 110, 123, 110, 101, 110, 115, 110, 101),
  term = c(
    "fall 2018", "fall 2018", "spring 2019", "fall 2020", "fall 2020", 
    "fall 2020", "fall 2020", "spring 2021", "fall 2021"
  ),
  grade = c("A", "B", "A", "B", "C", "A", "C", "D", "B")
)

# Information about the courses each student took per semester.
# One row per (student_id, course, term) combination.
course
```

```{r}
# Forgetting to join by both `student_id` and `term`!
term |>
  left_join(course, by = join_by(student_id))
```

In the example above, we've forgotten to include the `term` column when joining these two tables together, which accidentally results in a small explosion of rows (we end up with 20 rows, more than in either original input, but not quite the maximum possible amount, which is a whopping 54 rows!).
Luckily, dplyr warns us that at least one row in each table matches more than one row in the opposite table - a sign that something isn't right.
At this point we can do one of two things:

-   Look into the new `relationship` argument that the warning mentions (we'll discuss this below)

-   Look at our join to see if we made a mistake

Of course, in this case we've messed up, and adding `term` into the by expression results in the correct (and silent) join:

```{r}
term |>
  left_join(course, by = join_by(student_id, term))
```

### Join `relationship`s

To adjust the joins to only warn on many-to-many relationships, we've done two things:

-   `multiple` now defaults to `"all"`, and is now focused solely on limiting the matches returned if multiple are detected, rather than also optionally warning/erroring.

-   We've added a new `relationship` argument.

The `relationship` argument allows you to explicitly specify the expected join relationship between the keys of `x` and `y` using the exact options we listed above: `"one-to-one"`, `"one-to-many"`, `"many-to-one"`, and `"many-to-many"`.
If the constraints of the relationship you choose are violated, an error is thrown.
For example, we could use this to require that the `student` + `term` join contains a one-to-many relationship between the two tables:

```{r}
student |>
  left_join(term, join_by(student_id), relationship = "one-to-many")
```

Let's violate this by adding a duplicate row in `student`:

```{r}
student_bad <- student |>
  tibble::add_row(
    student_id = 1, 
    transfer = FALSE, 
    initial_term = "fall 2019", 
    .after = 1
  )

student_bad
```

```{r, error=TRUE}
student_bad |>
  left_join(term, join_by(student_id), relationship = "one-to-many")
```

The default value of `relationship` doesn't add any constraints, but for equality joins it will check to see if a many-to-many relationship exists, and will warn if one occurs (like with the `term` + `course` join from above).
As mentioned before, this is quite hard to do, and often means you have a mistake in your join call or in the data itself.
If you really do want to perform a join with this kind of relationship, to silence the warning you can explicitly specify `relationship = "many-to-many"`.

One last thing to note is that `relationship` doesn't handle the case of an *unmatched* row.
For that, you should use the `unmatched` argument that was also added in 1.1.0.
The combination of `relationship` and `unmatched` provides a complete set of tools for adding production level quality control checks to your joins.

## Acknowledgements

The examples used in this blog post were adapted from [\@eipi10](https://github.com/eipi10) in [this issue](https://github.com/tidyverse/dplyr/issues/6717).

We'd like to thank all 66 contributors who help in someway, whether it was filing issues or contributing code and documentation: [\@alexhallam](https://github.com/alexhallam), [\@ammar-gla](https://github.com/ammar-gla), [\@arnaudgallou](https://github.com/arnaudgallou), [\@ArthurAndrews](https://github.com/ArthurAndrews), [\@AuburnEagle-578](https://github.com/AuburnEagle-578), [\@batpigandme](https://github.com/batpigandme), [\@billdenney](https://github.com/billdenney), [\@Bisaloo](https://github.com/Bisaloo), [\@bitplane](https://github.com/bitplane), [\@chrarnold](https://github.com/chrarnold), [\@D5n9sMatrix](https://github.com/D5n9sMatrix), [\@daattali](https://github.com/daattali), [\@DanChaltiel](https://github.com/DanChaltiel), [\@DavisVaughan](https://github.com/DavisVaughan), [\@dieghernan](https://github.com/dieghernan), [\@dkutner](https://github.com/dkutner), [\@eipi10](https://github.com/eipi10), [\@eitsupi](https://github.com/eitsupi), [\@emilBeBri](https://github.com/emilBeBri), [\@fawda123](https://github.com/fawda123), [\@fedassembly](https://github.com/fedassembly), [\@fkohrt](https://github.com/fkohrt), [\@gavinsimpson](https://github.com/gavinsimpson), [\@geogale](https://github.com/geogale), [\@ggrothendieck](https://github.com/ggrothendieck), [\@hadley](https://github.com/hadley), [\@hope-data-science](https://github.com/hope-data-science), [\@jaganmn](https://github.com/jaganmn), [\@jakub-jedrusiak](https://github.com/jakub-jedrusiak), [\@JorisChau](https://github.com/JorisChau), [\@krlmlr](https://github.com/krlmlr), [\@krprasangdas](https://github.com/krprasangdas), [\@larry77](https://github.com/larry77), [\@lionel-](https://github.com/lionel-), [\@lschneiderbauer](https://github.com/lschneiderbauer), [\@LukasWallrich](https://github.com/LukasWallrich), [\@maellecoursonnais](https://github.com/maellecoursonnais), [\@manhnguyen48](https://github.com/manhnguyen48), [\@mattansb](https://github.com/mattansb), [\@mgirlich](https://github.com/mgirlich), [\@mhaynam](https://github.com/mhaynam), [\@MichaelChirico](https://github.com/MichaelChirico), [\@mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [\@mkoohafkan](https://github.com/mkoohafkan), [\@moodymudskipper](https://github.com/moodymudskipper), [\@Moohan](https://github.com/Moohan), [\@msgoussi](https://github.com/msgoussi), [\@multimeric](https://github.com/multimeric), [\@osheen1](https://github.com/osheen1), [\@Pozdniakov](https://github.com/Pozdniakov), [\@psychelzh](https://github.com/psychelzh), [\@pur80a](https://github.com/pur80a), [\@robayo](https://github.com/robayo), [\@rszulkin](https://github.com/rszulkin), [\@salim-b](https://github.com/salim-b), [\@sda030](https://github.com/sda030), [\@sfirke](https://github.com/sfirke), [\@shannonpileggi](https://github.com/shannonpileggi), [\@stephLH](https://github.com/stephLH), [\@szabgab](https://github.com/szabgab), [\@tjebo](https://github.com/tjebo), [\@Torvaney](https://github.com/Torvaney), [\@twest820](https://github.com/twest820), [\@vanillajonathan](https://github.com/vanillajonathan), [\@warnes](https://github.com/warnes), and [\@zknitter](https://github.com/zknitter).
