---
output: hugodown::hugo_document

slug: waldo-0-3-0
title: waldo 0.3.0
date: 2021-08-20
author: Hadley Wickham
description: >
    waldo 0.3.0 improves the display of data frames differences, and gives the 
    objects being compared the ability to control the detail of their 
    comparisons.

photo:
  url: https://unsplash.com/photos/n6vS3xlnsCc
  author: Kelley Bozarth

# one of: "deep-dive", "learn", "package", "programming", or "other"
categories: [package] 
tags: [testthat, waldo]
---

<!--
TODO:
* [ ] Look over / edit the post's title in the yaml
* [ ] Edit (or delete) the description; note this appears in the Twitter card
* [ ] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [ ] Find photo & update yaml metadata
* [ ] Create `thumbnail-sq.jpg`; height and width should be equal
* [ ] Create `thumbnail-wd.jpg`; width should be >5x height
* [ ] `hugodown::use_tidy_thumbnails()`
* [ ] Add intro sentence, e.g. the standard tagline for the package
* [ ] `usethis::use_tidy_thanks()`
-->

We're delighted to announce the release of [waldo](https://waldo.r-lib.org) 0.3.0. waldo is designed to find and concisely describe the difference between a pair of R objects. It was designed primarily to improve failure messages for `testthat::expect_equal()`, but it turns out to be useful in a number of other situations.

You can install it from CRAN with:

```{r, eval = FALSE}
install.packages("waldo")
```

This blog post highlights the two biggest changes in this release: a new display format for data frame differences, and a new attribute that waldo can use to control the details of comparison. You can see a full list of changes in the [release notes](https://github.com/r-lib/waldo/blob/master/NEWS.md)

```{r setup}
library(waldo)
```

## Data frame differences

waldo 0.2.0 treated data frames in the same way as lists, which was useful if a column changed, but wasn't exactly informative if a row changed. In 0.3.0, data frames get a new method that summarise the changes in a row-oriented form:

```{r}
df1 <- data.frame(x = c(1, 2, 3, 4, 5), y = c("a", "b", "c", "d", "e"))
df2 <- data.frame(x = c(1, 2, 3, 10, 4, 5), y = c("a", "b", "c", "X", "d", "e"))
compare(df1, df2)
```

You'll notice that you still get the per column comparison as well. This is important for cases where the data frame print identically, but the underlying data is different. Here's a simple example:

```{r}
df1 <- data.frame(x = c("a", "b", "c"), stringsAsFactors = TRUE)
df2 <- data.frame(x = c("a", "b", "c"), stringsAsFactors = FALSE)
compare(df1, df2)
```

## Control of comparison

When developing new data structures, you often need to be able to control the details of the comparisons that waldo performs. For example, take the xml2 package, which uses the [libxml](http://xmlsoft.org) C library to work with xml. When you parse xml with xml2, it looks like it's just represented as a string:

```{r}
library(xml2)
x1 <- xml2::read_xml("<a>1</a>")
x1
```

But behind the scenes, it's actually two pointers to C data structures created by libxml:

```{r}
str(x1)
```

This means that a naive comparison isn't very useful:

```{r}
x2 <- xml2::read_xml("<a>2</a>")
compare(unclass(x1), unclass(x2))
```

To resolve this problem waldo provides the `compare_proxy()` generic which is called on every S3 object. You can use it to transform your object into something that's more easily compared. And waldo includes a built-in `compare_proxy.xml_node()` method that converts the C data structures back to strings:

```{r}
compare(x1, x2)
```

(You could also imagine converting the xml structure to some richer tree structure in R, but I didn't take the time to do so.)

This method has existed for some time, but waldo 0.3.0 generalised it so as well as returning the modifying object, it also returns a modifed "path" that describes how the object has been transformed:

```{r}
waldo:::compare_proxy.xml_node
```

This means that if the comparison fails, you get a clear path to the root cause.

Creating a new S3 method is reasonably heavy (and requires a little gymnastics in your package to correctly register without taking a hard dependency on waldo), so thanks to [Duncan Murdoch](http://github.com/dmurdoch) waldo 0.3.0 gains a new way of controlling comparisons: the `waldo_opts` attribute.

The `waldo_opts` attribute is a list with the same names and valid values as the arguments to `compare()`, and it will override the default arguments. This is a powerful tool because you can inject these attributes at any level of the object hierarchy, no matter how deep. 

For example, take these two lists which contain the same data but in different order:

```{r}
x1 <- list(a = 1, b = 2)
x2 <- list(b = 2, a = 1)
```

Usually waldo will report these to be different:

```{r}
compare(x1, x2)
```

But with the new `list_as_map` arugment (also thanks to an idea from Duncan Murdoch), you can request that the list be compared purely as mappings between names and values:

```{r}
compare(x1, x2, list_as_map = TRUE)
```

This is great if you want this comparison to happen at the top level of the object, but what if the difference is buried deep within a list of lists, and you only want `list_as_map` to affect one small part of the object? Well, now you can add the `waldo_opts` attribute:

```{r}
attr(x1, "waldo_opts") <- list(list_as_map = TRUE)
compare(x1, x2)
```

## Acknowledgements

