---
output: hugodown::hugo_document

slug: reprex-1-0-0
title: reprex 1 0 0
date: 2021-01-27
author: Jennifer Bryan
description: >
    We've never blogged about reprex before, so the release of v1.0.0 seems
    like a good reason occasion for it.

photo:
  url: https://www.flickr.com/photos/iamagenious/6251271390/sizes/o/
  author: Eric Molina

# one of: "deep-dive", "learn", "package", "programming", or "other"
categories: [package] 
tags: [reprex,tidyverse]
---

We're exhilarated to announce the release of reprex 1.0.0 ([reprex.tidyverse.org](https://reprex.tidyverse.org)).
reprex is a package that helps you prepare **REPR**oducible **EX**amples to share in places where people talk about code, e.g., on GitHub, on Stack Overflow, and in Slack or email messages.

You can install the current version of reprex from CRAN with[^tidyverse-install]:

[^tidyverse-install]: Another way you might get reprex is by installing the tidyverse meta-package.
reprex is one of the packages installed by `install.packages("tidyverse")`, however it is **not** among the core packages attached by `library(tidyverse)`.

```{r, eval = FALSE}
install.packages("reprex")
```

It turns out we've never blogged about reprex here on [tidyverse.org](https://www.tidyverse.org), so we'll start with a general overview for newcomers, then close with a summary of recent changes of interest to existing users.

You can see a full list of changes in the [release notes](https://reprex.tidyverse.org/news/index.html#reprex-1-0-0-2021-01-27).

## Why reprex exists

reprex is a convenience package that combines the power of R Markdown with conventional wisdom about what makes a good reproducible example.
We believe that conversations about code are more productive with:

* Code that **actually runs**
* Code that the reader **doesn't necessarily have to run**
* Code that the reader **can easily run**

Someone new to R might be puzzled by this code and want to discuss it:

```{r}
x <- factor("Hello, ")
y <- factor("world!")
x
y
c(x, y)
```

You, as a reader, just benefited from a few things:

1. The code was actually executed by the R interpreter, because this post is
   generated from an `.Rmd` document. I am probably not wasting your time with
   fictional or incomplete code.
2. You got to see the actual result without firing up R yourself and executing
   this code locally. Many R veterans will instantly recognize what's going on
   and be able to provide some useful explanation at this point.
3. If you do want to execute the code yourself, you can easily copy the entire
   chunk, paste it into R, and press enter[^reprex-clean]. Compare that with the
   fussy edits you'd have to make with this copy/paste from the R Console
   (removing prompts and output):

    ``` r
    > x <- factor("Hello, ")
    > y <- factor("world!")
    > x
    [1] Hello, 
    Levels: Hello, 
    > y
    [1] world!
    Levels: world!
    > c(x, y)
    [1] 1 1
    ```

[^reprex-clean]: Since the output is commented out, its presence is harmless. But you can even use `reprex::reprex_clean()` and friends to un-reprex code, if you like.

All of this is entirely possible without reprex.
People just need to create a suitable `.R` or `.Rmd` file, render it to an appropriate output format, in a fresh R session, with working directory set to session temp directory.
And also send me a pony!

Most of us won't do this, because it's a hassle.
`reprex::reprex()` aims to make doing the right thing so easy that people stop taking cell phone photos of their broken code.

## `reprex::reprex()`

`reprex::reprex()` is the main function in the package.
It takes a little bit of R code, probably from the clipboard or current selection, and does all the fiddly things we mentioned:

* Writes it to a file. By default, to a file below the session temp directory.
* Sets various R and knitr options that are especially favorable for reprexes,
  such as the `error = TRUE` chunk option.
* Renders the code in a fresh R session to ruthlessly expose missing
  `library()` calls and the use of objects that haven't been defined.
* Ensures the output is optimized for the target venue, e.g. GitHub-flavored
  Markdown vs. commented R code vs. Rich Text Format.
  
By default, the result is waiting on your clipboard and you'll also see an HTML preview of it.

Here is a 50 second video that shows an entire roundtrip: copying local code, `reprex()`ing, opening a GitHub issue, copying reprex from the issue, and re-executing the code in a local R session:

<iframe width="560" height="315" src="https://www.youtube.com/embed/35suhGR53wQ" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

There are many other handy features:

* Optionally include session info
* Post a figure to imgur.com and embed its link in the reprex
* Handy RStudio addin and gadget for even more convenience
* More ways to provide input and get output

Head over to [reprex.tidyverse.org](https://reprex.tidyverse.org) to learn more.

## What's new in v1.0.0

cool stuff!

## Acknowledgements

We've never blogged about reprex here, so we'll take this chance to thank all 116 people who have helped get reprex to this point. There have been 47 contributors since the previous release (v0.3.0).

[&#x0040;1029YanMa](https://github.com/1029YanMa), [&#x0040;Abhijitsj](https://github.com/Abhijitsj), [&#x0040;aegerton](https://github.com/aegerton), [&#x0040;alexpghayes](https://github.com/alexpghayes), [&#x0040;alistaire47](https://github.com/alistaire47), [&#x0040;andresrcs](https://github.com/andresrcs), [&#x0040;assignUser](https://github.com/assignUser), [&#x0040;atusy](https://github.com/atusy), [&#x0040;baptiste](https://github.com/baptiste), [&#x0040;barryrowlingson](https://github.com/barryrowlingson), [&#x0040;batpigandme](https://github.com/batpigandme), [&#x0040;billdenney](https://github.com/billdenney), [&#x0040;blairj09](https://github.com/blairj09), [&#x0040;brshallo](https://github.com/brshallo), [&#x0040;brunocarlin](https://github.com/brunocarlin), [&#x0040;carlmorgenstern](https://github.com/carlmorgenstern), [&#x0040;cderv](https://github.com/cderv), [&#x0040;chester-gan](https://github.com/chester-gan), [&#x0040;chris-prener](https://github.com/chris-prener), [&#x0040;chsafouane](https://github.com/chsafouane), [&#x0040;coatless](https://github.com/coatless), [&#x0040;ColinFay](https://github.com/ColinFay), [&#x0040;cooknl](https://github.com/cooknl), [&#x0040;crew102](https://github.com/crew102), [&#x0040;cstepper](https://github.com/cstepper), [&#x0040;cwickham](https://github.com/cwickham), [&#x0040;daattali](https://github.com/daattali), [&#x0040;david-romano](https://github.com/david-romano), [&#x0040;davidbody](https://github.com/davidbody), [&#x0040;DavisVaughan](https://github.com/DavisVaughan), [&#x0040;dchiu911](https://github.com/dchiu911), [&#x0040;dgrtwo](https://github.com/dgrtwo), [&#x0040;dpprdan](https://github.com/dpprdan), [&#x0040;dskard](https://github.com/dskard), [&#x0040;ellessenne](https://github.com/ellessenne), [&#x0040;emiltb](https://github.com/emiltb), [&#x0040;filipwastberg](https://github.com/filipwastberg), [&#x0040;franknarf1](https://github.com/franknarf1), [&#x0040;friendly](https://github.com/friendly), [&#x0040;gaborcsardi](https://github.com/gaborcsardi), [&#x0040;GegznaV](https://github.com/GegznaV), [&#x0040;gergness](https://github.com/gergness), [&#x0040;gvdr](https://github.com/gvdr), [&#x0040;hadley](https://github.com/hadley), [&#x0040;harrismcgehee](https://github.com/harrismcgehee), [&#x0040;HeidiSeibold](https://github.com/HeidiSeibold), [&#x0040;helix123](https://github.com/helix123), [&#x0040;Henrik-P](https://github.com/Henrik-P), [&#x0040;HughParsonage](https://github.com/HughParsonage), [&#x0040;IndrajeetPatil](https://github.com/IndrajeetPatil), [&#x0040;isteves](https://github.com/isteves), [&#x0040;j450h1](https://github.com/j450h1), [&#x0040;jasonmtroos](https://github.com/jasonmtroos), [&#x0040;jayhesselberth](https://github.com/jayhesselberth), [&#x0040;jemus42](https://github.com/jemus42), [&#x0040;jennformatics](https://github.com/jennformatics), [&#x0040;jennybc](https://github.com/jennybc), [&#x0040;JiaxiangBU](https://github.com/JiaxiangBU), [&#x0040;jimhester](https://github.com/jimhester), [&#x0040;joelgombin](https://github.com/joelgombin), [&#x0040;JohnMount](https://github.com/JohnMount), [&#x0040;jooyoungseo](https://github.com/jooyoungseo), [&#x0040;juliasilge](https://github.com/juliasilge), [&#x0040;jzadra](https://github.com/jzadra), [&#x0040;karawoo](https://github.com/karawoo), [&#x0040;kevinushey](https://github.com/kevinushey), [&#x0040;krlmlr](https://github.com/krlmlr), [&#x0040;lbusett](https://github.com/lbusett), [&#x0040;lionel-](https://github.com/lionel-), [&#x0040;lizhiwei1994](https://github.com/lizhiwei1994), [&#x0040;llrs](https://github.com/llrs), [&#x0040;lorenzwalthert](https://github.com/lorenzwalthert), [&#x0040;maelle](https://github.com/maelle), [&#x0040;marionlouveaux](https://github.com/marionlouveaux), [&#x0040;markdly](https://github.com/markdly), [&#x0040;mattfidler](https://github.com/mattfidler), [&#x0040;maurolepore](https://github.com/maurolepore), [&#x0040;mcanouil](https://github.com/mcanouil), [&#x0040;mdlincoln](https://github.com/mdlincoln), [&#x0040;mine-cetinkaya-rundel](https://github.com/mine-cetinkaya-rundel), [&#x0040;missaugustina](https://github.com/missaugustina), [&#x0040;moodymudskipper](https://github.com/moodymudskipper), [&#x0040;mrchypark](https://github.com/mrchypark), [&#x0040;mrdwab](https://github.com/mrdwab), [&#x0040;njtierney](https://github.com/njtierney), [&#x0040;no-reply](https://github.com/no-reply), [&#x0040;noamross](https://github.com/noamross), [&#x0040;npjc](https://github.com/npjc), [&#x0040;paleolimbot](https://github.com/paleolimbot), [&#x0040;pat-s](https://github.com/pat-s), [&#x0040;paternogbc](https://github.com/paternogbc), [&#x0040;PeteHaitch](https://github.com/PeteHaitch), [&#x0040;pgensler](https://github.com/pgensler), [&#x0040;PiotrKoller](https://github.com/PiotrKoller), [&#x0040;PMassicotte](https://github.com/PMassicotte), [&#x0040;prosoitos](https://github.com/prosoitos), [&#x0040;PublicHealthDataGeek](https://github.com/PublicHealthDataGeek), [&#x0040;r2evans](https://github.com/r2evans), [&#x0040;rcorty](https://github.com/rcorty), [&#x0040;restonslacker](https://github.com/restonslacker), [&#x0040;RLesur](https://github.com/RLesur), [&#x0040;robjhyndman](https://github.com/robjhyndman), [&#x0040;romainfrancois](https://github.com/romainfrancois), [&#x0040;rpruim](https://github.com/rpruim), [&#x0040;sckott](https://github.com/sckott), [&#x0040;scottcame](https://github.com/scottcame), [&#x0040;sfirke](https://github.com/sfirke), [&#x0040;sjspielman](https://github.com/sjspielman), [&#x0040;Tugsdelger](https://github.com/Tugsdelger), [&#x0040;tungmilan](https://github.com/tungmilan), [&#x0040;uribo](https://github.com/uribo), [&#x0040;wlandau](https://github.com/wlandau), [&#x0040;yonicd](https://github.com/yonicd), [&#x0040;yutannihilation](https://github.com/yutannihilation), [&#x0040;zkamvar](https://github.com/zkamvar), and [&#x0040;zx8754](https://github.com/zx8754).
