---
output: hugodown::hugo_document

slug: nanoparquet-0-4-0
title: nanoparquet 0.4.0
date: 2025-01-28
author: Gábor Csárdi
description: >
    nanoparquet 0.4.0 comes with a new and much faster `read_parquet()`,
    configurable type mappings in `write_parquet()`, and a new
    `append_parquet()`.

photo:
  url: https://www.pexels.com/photo/person-running-in-the-hallway-796545/
  author: Michael Foster

# one of: "deep-dive", "learn", "package", "programming", "roundup", or "other"
categories: [package]
tags: [parquet]
---

<!--
TODO:
* [x] Look over / edit the post's title in the yaml
* [x] Edit (or delete) the description; note this appears in the Twitter card
* [x] Pick category and tags (see existing with `hugodown::tidy_show_meta()`)
* [x] Find photo & update yaml metadata
* [x] Create `thumbnail-sq.jpg`; height and width should be equal
* [x] Create `thumbnail-wd.jpg`; width should be >5x height
* [x] `hugodown::use_tidy_thumbnails()`
* [x] Add intro sentence, e.g. the standard tagline for the package
* [x] `usethis::use_tidy_thanks()`
-->

We're thrilled to announce the release of [nanoparquet](
  https://nanoparquet.r-lib.org/) 0.4.0. nanoparquet is an R package that
reads and writes Parquet files.

You can install it from CRAN with:

```r
install.packages("nanoparquet")
```

This blog post will show the most important new features of nanoparquet
0.4.0: You can see a full list of changes in the [release notes]({ TODO }).

## Brand new `read_parquet()`

nanoparquet 0.4.0 comes with a completely rewritten Parquet reader.
The new version has an architecture that is easier to embed into R, and
facilitates great features, like `append_parquet()` and the new
`col_select` argument. (More to come!) The new reader is also much faster,
see the "Benchmarks" chapter.

## Read a subset of columns

`read_parquet()` now has a new argument called `col_select`, that lets you
read a subset of the columns from the Parquet file. Unlike for row oriented
file formats like CSV, this means that the reader never needs to touch the
columns that are not needed for analysis. Reading a handful of columns only
always takes the same amount of time, no matter how many more columns the
Parquet file might have!

You can either use column indices or column names to specify the columns
to read. Here is an example.

```{r setup}
library(nanoparquet)
library(pillar)
```

```{r include = FALSE}
if (!file.exists("flights.parquet")) {
  write_parquet(nycflights13::flights, "flights.parquet")
}
```

This is the `nycflights13::flights` data set:

```{r col_select}
read_parquet(
  "flights.parquet",
  col_select = c("dep_time", "arr_time", "carrier")
)
```

Use `read_parquet_schema()` if you want to see the structure of the Parquet
file first:

```{r read_parquet_schema}
read_parquet_schema("flights.parquet")
```

The output of `read_parquet_schema()` also shows you the R type that
nanoparquet will use for each column.

## Appending to Parquet files

The new `append_parquet()` function makes it easy to append new data to
a Parquet file, without first reading the whole file into memory.
The schema of the file and the schema new data must match of course. Lets
merge `nycflights13::flights` and `nycflights23::flights`:

```{r append_parquet}
file.copy("flights.parquet", "allflights.parquet", overwrite = TRUE)
append_parquet(nycflights23::flights, "allflights.parquet")
```

`read_parquet_info()` returns the most basic information about a Parquet
file:

```{r read_parquet_info}
read_parquet_info("flights.parquet")
read_parquet_info("allflights.parquet")
```

Note that you should probably still create a backup copy of the original
file when using `append_parquet()`. If the appending process is interrupted
by a power failure, then you might end up with an incomplete and invalid
Parquet file.

## Schemas and type conversions

In nanoparquet 0.4.0 `write_parquet()` takes a `schema` argument that
can customize the R to Parquet type mappings. E.g. by default
`write_parquet()` writes an R character vector as a `STRING` Parquet type.
If you'd like to write a certain character column as an `ENUM` type
instead, you'll need to specify that in `schema`:

```{r schema}
write_parquet(
  nycflights13::flights,
  "newflights.parquet",
  schema = parquet_schema(carrier = "ENUM")
)
read_parquet_schema("newflights.parquet")
```

Here we wrote the `carrier` column as `ENUM`, and left the other other
columns to use the default type mappings.

See the [`nanoparquet-types`](
  https://nanoparquet.r-lib.org/dev/reference/nanoparquet-types.html#r-s-data-types
) manual page for the possible type mappings (lots of new ones!) and also
for the default ones.

## Encodings

It is now also possible to customize the encoding of each column in
`write_parquet()`, via the `encoding` argument. By default
`write_parquet()` tries to choose a good encoding based on the type and the
values of a column. E.g. it checks a small sample for repeated values to
decide if it is worth using a dictionary encoding (`RLE_DICTIONARY`).

If `write_parquet()` gets it wrong, use the `encoding` argument to force an
encoding. The following forces the `PLAIN` encoding for all columns. This
encoding is very fast to write, but creates a larger file. You can also
specify different encodings for different columns, see the
[`write_parquet()` manual page](
  https://nanoparquet.r-lib.org/dev/reference/write_parquet.html).

```{r encoding}
write_parquet(
  nycflights13::flights,
  "plainflights.parquet",
  encoding = "PLAIN"
)
file.size("flights.parquet")
file.size("plainflights.parquet")
```

See more about the implemented encodings and how the defaults are
selected in the [`parquet-encodings` manual page](
  https://nanoparquet.r-lib.org/dev/reference/parquet-encodings.html).

## API changes

Some nanoparquet functions have new, better names in nanoparquet 0.4.0.
In particular, all functions that read from a Parquet file have a
`read_parquet` prefix now. The old functions still work, with a warning.

Also, the `parquet_schema()` function is now for creating a new Parquet
schema from scratch, and not for inferring a schema from a data frame
(use `infer_parquet_schema()`) or for reading the schema from a Parquet
file (use `read_parquet_schema()`). `parquet_schema()` falls back to the
old behaviour when called with a file name, with a warning, so this is not
a breaking change (yet), and old code still works.

See the complete list of API changes in the [Changelog](
  https://nanoparquet.r-lib.org/dev/news/index.html).

## Benchmarks

We are very excited about the performance of the new Parquet reader, and
the Parquet writer was always quite speedy, so we ran a simple benchmark.

We compared nanoparquet to the Parquet implementations in Apache Arrow and
DuckDB, and also to CSV readers and writers, on a real data set, for
samples of 330k, 6.7 million and 67.4 million rows (40MB, 800MB and 8GB in
memory). For these data nanoparquet is indeed very competitive with both
Arrow and DuckDB.

You can see the full results [on the website](
  https://nanoparquet.r-lib.org/dev/articles/benchmarks.html).

## Other changes

Other important changes in nanoparquet 0.4.0 include:

* `write_parquet()` can now write multiple row groups. By default it puts
  at most 10 million rows in a single row group. (This is subject to
    https://nanoparquet.r-lib.org/dev/references/parquet_options.html
  ) on how to change the default.

* `write_parquet()` now writes minimum and maximum statistics (by default)
  for most Parquet types. See the [`parquet_options()` manual page](
    https://nanoparquet.r-lib.org/dev/reference/parquet_options.html
  ) on how to turn this off, which will probably make the writer faster.

* `write_parquet()` can now write version 2 data pages. The default is
  still version 1, but it might change in the future.

* New `compression_level` option to select the compression level manually.

* `read_parquet()` can now read from an R connection.

## Acknowledgements

[&#x0040;alvarocombo](https://github.com/alvarocombo), [&#x0040;D3SL](https://github.com/D3SL), [&#x0040;gaborcsardi](https://github.com/gaborcsardi), and [&#x0040;RealTYPICAL](https://github.com/RealTYPICAL).
